---
title: "E1_AllTaxa_Analysis"
author: "Maya Parker-Smith"
date: "2023-11-13"
output: 
 html_document:
    smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---

Creates tabs in the HTML document \# {.tabset .tabset-pills .tabset-fade}

```{r setup, include=FALSE}
#Clear environment
rm(list=ls())

#Read in necessary libraries
library(tidyverse)
library(boot)
library(broom)
library(corrplot)
library(cowplot)
library(ggpubr)
library(lme4)
library(MASS)
library(multcomp)
library(purrr)
library(readxl)
library(reshape2)
library(tseries)
library(vegan)

#Call our custom functions for classification and analysis
source("classify_functions.R")
```

#Read in Data Tables This includes the watersheds tables and the species-level data for the 4 taxa.

```{r}
watershed_info <- read_excel("Datasets/Watershed_data/Watershed_info.xlsx")

#Reading in the cleaned taxa data sets
E0_smammals <- read_csv("Datasets/E0_cleaned_data/E0_smammals.csv")
E0_grasshoppers <- read_csv("Datasets/E0_cleaned_data/E0_grasshoppers.csv")
E0_birds <- read_csv("Datasets/E0_cleaned_data/E0_birds.csv")
E0_plants <- read_csv("Datasets/E0_cleaned_data/E0_plants.csv")

#read in plant species list
pl_info <- read_csv("Datasets/Raw_data/Plants/plant_sp_list.csv")
```

# Clean Data to Classified Incidence for All Taxa

This Rmarkdown is for taking the clean datasets for all 4 taxa (small-mammals, grasshoppers, birds, and plants) and using the incidence classification method to categorize each species. This file also estimates composition change using presence-absence information (Jaccard dissimilarity) and abundance information (Bray-Curtis dissimilarity), richness change (linear model), and abundance change (linear model) over time.

## Brief overview of Classification

Incidence Classification categories include:

-   **"No_change-absent"** \<- Species was detected at other surrounding watersheds, but it was absent at specified watershed for the duration of the time series.
-   **"Rare"** \<- Species was present at specified watershed for between 1-10% of years in the duration of the time series.
-   **"Common"** \<- Species at watershed is present in \>=90% of years in which data was collected.
-   **"Increasing"** \<- The presence of the species increases throughout the duration of the study (it is absent during the beginning of the study, but appears towards the end of the study)
-   **"Decreasing"** \<- The presence of the species decreases throughout the duration of the study (it is present during the beginning of the study, but then disappears towards the end of the study).
-   **"Recurrent"** \<- Species is detected in consecutive years and is followed by the absence of the species for a period of time, in a distinct pattern. Species must at least two instances of presences followed by absences.
-   **"Clumped"** \<- Species is detected in a single group of consecutive years, a cluster of incidences.
-   **"Random"** \<- Species is present in an intermediate number of years (11%-89%) and detected sporadically (statistically randomly) for the duration of the study at watershed.


Abundance Classification categories include:

-   **Invariant** \<- Species abundance did not vary across the time series. Cannot be analyzed.
-   **"No_change-absent"** \<- Species was detected at other surrounding watersheds, but it was absent at specified watershed for the duration of the time series (\# years present = 0).
-   **"Rare"** \<- Species was present at specified watershed for between 1-10% of years in the duration of the time series. Abundance will not be classified.
-   **"No_change-increasing"** \<- Species at watershed is above average abundance in \>=90% of years in which data was collected. #CHECKME: is this even possible?
-   **"No_change-decreasing"** \<- Species at watershed is below average abundance in \>=90% of years in which data was collected. #CHECKME: is this even possible?
-   **"Increasing"** \<- The second half of the study contains most of the above average records of the species (it is mostly below average during the beginning of the study)
-   **"Decreasing"** \<- The second half of the study contains most of the below average records of the species (it is mostly above average during the beginning of the study)
-   **"Recurrent"** \<- Species abundance fluctuations appear to be clumped into consecutive years, with multiple distinct periods of at or above average abundance.
-   **"Clumped"** \<- Species abundance fluctuations appear to have one distinct cluster of at or above average abundances.
-   **"Random"** \<- Species abundance fluctuates in an unpredictable pattern of above vs below average for the duration of the study.

#----------------------------------------------------------------------------------

# Determining Species Incidence Classification

## Small-mammals, Grasshoppers, and Birds {.tabset}
small mammals: 1992-2013
grasshoppers: 2002-2020
birds: 1992-2009

### Shared classification pipeline (smammals, grasshoppers, birds)

These three taxa share the same pipeline: sum counts by year/watershed/species, pivot wide,
run getTrends3.0 (incidence) and getAbundTrends (abundance), write results.

**Per-taxon differences handled below:**
- smammals: na.rm = TRUE for decostand (no replace_na needed; all years present by design)
- grasshoppers: replace_na to 0 across year columns before classification (species absent in
  a year have NA from pivot_wider, not zero); na.rm = FALSE for decostand
- birds: same as grasshoppers (replace_na + na.rm = FALSE); now uses Species column (= AOU
  code) following E0 standardization — verified identical results to original AOU_code grouping

Each taxon's output objects are named identically to the original
(e.g. smammals_classification, birds_years_wide) so all downstream code is unchanged.

```{r classification_smammals_grasshoppers_birds}

# ------ SMALL MAMMALS (1992-2013) ------
# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
smammals <- E0_smammals %>%
  group_by(Recyear, Watershed_name, Species) %>%
  reframe(Count = sum(Count))

#Combine the species & watershed variables to create a unique identifier
smammals <- smammals %>% unite("Species.watershed", Species:Watershed_name, sep = ".", remove = FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
smammals_years_wide <- smammals %>%
  pivot_wider(names_from = "Recyear", values_from = "Count")

#Create a separate species.watershed dataframe to add back to the dataset after standardization
smammals_env <- smammals_years_wide[, 1:3]

#Standardize by presence/absence
decostand_smammals_pa <- decostand(smammals_years_wide[, -(1:3)], "pa", na.rm = TRUE)

# ------ GRASSHOPPERS (2002-2020) ------
# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
grasshoppers <- E0_grasshoppers %>%
  group_by(Recyear, Watershed_name, Species) %>%
  reframe(Count = sum(Count)) %>%
  unite("Species.watershed", Species:Watershed_name, sep = ".", remove = FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
grasshoppers_years_wide <- grasshoppers %>%
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:22, ~replace_na(., 0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
grasshoppers_env <- grasshoppers_years_wide[, 1:3]

#Standardize by presence/absence
decostand_grasshoppers_pa <- decostand(grasshoppers_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

# ------ BIRDS (1992-2009) ------
# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
# NOTE: Species column now holds AOU code (standardized in E0); results identical to
#   original grouping by AOU_code
birds <- E0_birds %>%
  group_by(Recyear, Watershed_name, Species) %>%
  reframe(Count = sum(Count)) %>%
  unite("Species.watershed", Species:Watershed_name, sep = ".", remove = FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
birds_years_wide <- birds %>%
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:21, ~replace_na(., 0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
birds_env <- birds_years_wide[, 1:3]

#Standardize by presence/absence
decostand_birds_pa <- decostand(birds_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

# ------ SHARED PIPELINE: incidence + abundance classification for all three taxa ------
# Taxa list: name -> list of (years_wide, env, decostand_pa, outfile_prefix, na_rm)
# The pipeline below runs identically for each; only the input objects differ.

taxa_list <- list(
  smammals    = list(years_wide   = smammals_years_wide,
                     env          = smammals_env,
                     pa           = decostand_smammals_pa,
                     out          = "Datasets/E1_output_data/E1_smammals/E1_smammals"),
  grasshoppers = list(years_wide  = grasshoppers_years_wide,
                     env          = grasshoppers_env,
                     pa           = decostand_grasshoppers_pa,
                     out          = "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers"),
  birds       = list(years_wide   = birds_years_wide,
                     env          = birds_env,
                     pa           = decostand_birds_pa,
                     out          = "Datasets/E1_output_data/E1_birds/E1_birds")
)

for (tx in names(taxa_list)) {
  tx_data <- taxa_list[[tx]]

  # --- INCIDENCE CLASSIFICATION ---
  #Transform the presence/absence data into a matrix so I can feed it to the function
  pa_matrix <- as.matrix(tx_data$pa)

  #Apply the getTrends3.0 function to each row of the presence/absence matrix
  trends <- apply(pa_matrix, 1, getTrends3.0)

  # Create a table with the species.watershed and the getTrends info.
  #   puts everything back into a dataframe format
  #   1-10 stands for the number of rows in the output tibble from getTrends3.0
  classification <- tibble(tx_data$env,
                           do.call(rbind, lapply(trends, (function(v){v[c(1:10)]}))))

  #Add back in the presence/absence per year data
  classification <- classification %>% add_column(tx_data$pa)

  # Write results to file
  write_csv(classification, paste0(tx_data$out, "_classified.csv"))

  # --- ABUNDANCE CLASSIFICATION ---
  #use the wide format count data as input; convert to matrix format and remove labels
  years_matrix <- as.matrix(tx_data$years_wide[, -(1:3)])

  # input matrix into label_abundance method
  # returns a new matrix where 0=below average, 1=above or equal to average
  #   final column = proportion of years the species was absent (proportion of zeros)
  abund_matrix <- years_matrix %>%
    as.data.frame() %>%
    mutate(row_id = row_number()) %>%
    split(.$row_id) %>%
    map(~ label_abundance(as.numeric(.x[, -ncol(.x)]))) %>%
    do.call(rbind, .)

  #input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
  abun_trends <- apply(abund_matrix, 1, getAbundTrends)

  #Create table with the getAbundTrends results annotated with species & watershed labels
  abun_classification <- tibble(tx_data$env,
                                do.call(rbind, lapply(abun_trends, (function(v){v[c(1:9)]}))))

  # Write to file
  write_csv(abun_classification, paste0(tx_data$out, "_classified_abund.csv"))

  # Assign results back to named objects so all downstream code works unchanged
  assign(paste0(tx, "_classification"),      classification)
  assign(paste0(tx, "_abun_classification"), abun_classification)
}
```


## Plants (1992 - 2022) {.tabset}

### Plants: Presence/ Absence & Incidence Classification

For this section, I am reorganizing the plant data and reformatting it into a presence/absence matrix. This matrix will then be ran through the getTrends3.0 function to get the classifications for each species at each watershed.

**Note**: Plant data does not collect abundance as counts, but uses cover instead. 
It is recorded  on a scale of 1-7 where: 
  - 1 = <1% cover
  - 2 = 1-5% cover
  - 3 = 5-25% cover
  - 4 = 25-50% cover
  - 5 = 50-75% cover
  - 6 = 75-95% cover
  - 7 = 95-100% cover

Conduct the **incidence classification** method
*Note: Here, the column named 'Count' represents median Cover class (range 1-7) for plants*
```{r plants incidence classification}

# Average the data based on watershed (not using SUM because it's cover, not count, data)
# Here, we won't differentiate the data based on the trap lines or season
# NOTE: Count=median(cover) here, but it can be converted into pres/abs later
plants <- E0_plants %>% 
  group_by(Recyear, Watershed_name, Species) %>% 
  reframe(Count = median(Cover)) %>%
  unite("Species.watershed", Species:Watershed_name, sep=".", remove=FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
plants_years_wide <- plants %>% 
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:34, ~replace_na(.,0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
plants_env <- plants_years_wide[, 1:3]

#Standardize by presence/absence
decostand_plants_pa <- decostand(plants_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

#Transform the presence/absence data into a matrix so I can feed it to the function
plants_pa_matrix <- as.matrix(decostand_plants_pa)

#Apply the getTrends3.0 function to each row of the presence/absence matrix
plants_trends <- apply(plants_pa_matrix, 1, getTrends3.0)

# Create a table with the species.watershed and the getTrends info. 
#   puts everything back into a dataframe format
#   1-10 stands for the number of rows in the output tibble from getTrends3.0
plants_classification <- tibble((plants_env), do.call(rbind,lapply(plants_trends,(function(v){v[c(1:10)]}))))

#Add back in the presence/absence per year data
plants_classification <- plants_classification %>% add_column(decostand_plants_pa)

# Write to file
write_csv(plants_classification, "Datasets/E1_output_data/E1_plants/E1_plants_classified.csv")
```

Conduct the **abundance classification** method
*Note: Here, the column named 'Count' represents median Cover class (range 1-7) for plants*
#FIXME: Does it make sense for ordinal cover class to use mean instead of median? Think about it...?
```{r plants abundance classification}
#use the wide format count data as input; convert to matrix format and remove labels
plants_years_matrix <- as.matrix(plants_years_wide[, -(1:3)])

# input matrix into label_abundance method
# returns a new matrix where 0=below average, 1=above or equal to average
#   final column = proportion of years the species was absence (proportion of zeros)
abund_matrix <- plants_years_matrix %>%
  as.data.frame() %>%
  mutate(row_id = row_number()) %>%
  split(.$row_id) %>%
  map(~ label_abundance(as.numeric(.x[ , -ncol(.x)]))) %>%
  do.call(rbind, .)

#input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
plants_abun_trends <- apply(abund_matrix, 1, getAbundTrends) 

#Create table with the getAbunTrends results annotated with species & watershed labels
#   labels from birds_env, created earlier
plants_abun_classification <- tibble((plants_env), do.call(rbind,lapply(plants_abun_trends,(function(v){v[c(1:9)]}))))

# Write to file
write_csv(plants_abun_classification,
          "Datasets/E1_output_data/E1_plants/E1_plants_classified_abund.csv")

```


#----------------------------------------------------------------------------------------------
## Summary statistics on classification (all taxa) {.tabset}
This section makes some basic summaries of the incidence classification results.


```{r summstat_all_taxa}

# Named list: taxon label -> classification and abun_classification objects
##  inc stands for incidence classes
##  abun stands for abundance classes
taxa_class <- list(
  smammals     = list(inc  = smammals_classification,
                      abun = smammals_abun_classification,
                      ymax_switch = 7),    # y position for geom_text on switch plot
  grasshoppers = list(inc  = grasshoppers_classification,
                      abun = grasshoppers_abun_classification,
                      ymax_switch = 7),
  birds        = list(inc  = birds_classification,
                      abun = birds_abun_classification,
                      ymax_switch = 20),
  plants       = list(inc  = plants_classification,
                      abun = plants_abun_classification,
                      ymax_switch = 90)
)

for (tx in names(taxa_class)) {
  inc  <- taxa_class[[tx]]$inc
  abun <- taxa_class[[tx]]$abun
  ymax <- taxa_class[[tx]]$ymax_switch

  # Make a table that counts how many times each species was classified a particular way
  abun_class_summary <- abun %>%
    group_by(Species, classification) %>%
    summarize(total_count = n(), .groups = 'drop')
  assign(paste0(tx, "_abun_class_summary"), abun_class_summary)

  # Make a table that counts how many classifications each species had (across 6 watersheds)
  # first, remove classification for no-change-absent (only count classifications when present)
  switch1 <- inc %>%
    filter(classification != "No_change-absent") %>%
    group_by(Species) %>%
    summarize(count = n_distinct(classification))
  assign(paste0(tx, "_class.switch1"), switch1)

  numsp    <- nrow(switch1)
  switchsp <- nrow(switch1[switch1$count > 1, ])

  # plot as a bar graph
  print(
    switch1 %>% ggplot(aes(x = count)) +
      geom_bar() +
      theme_bw() +
      geom_text(x = 3.5, y = ymax,
                label = paste0("Switch class = ",
                               switchsp, "/", numsp,
                               " (", round(switchsp/numsp, 2)*100, "%)")) +
      labs(x = "Number of classifications per species",
           title = paste0(tx, ": incidence classification switches"))
  )

  # Make a table that shows how each species was categorized, and how many times
  # first, remove classification for no-change-absent (only count classifications when present)
  switch2 <- inc %>%
    filter(classification != "No_change-absent") %>%
    group_by(Species, classification) %>%
    summarize(count = n_distinct(Watershed_name))
  switch2_wide <- switch2 %>% pivot_wider(names_from = classification, values_from = count)
  assign(paste0(tx, "_class.switch2"),      switch2)
  assign(paste0(tx, "_class.switch2_wide"), switch2_wide)

  # Make a table that shows how many species were in each incidence classification, by watershed
  # first, remove classification for no-change-absent (only count classifications when present)
  switch3 <- inc %>%
    filter(classification != "No_change-absent") %>%
    group_by(Watershed_name, classification) %>%
    summarize(count = n_distinct(Species))
  switch3_wide <- switch3 %>% pivot_wider(names_from = classification, values_from = count)
  assign(paste0(tx, "_class.switch3"),      switch3)
  assign(paste0(tx, "_class.switch3_wide"), switch3_wide)
  #view(switch3_wide)

  # Make a table that shows how many species were in each abundance classification, by watershed
  switch4 <- abun %>%
    group_by(Watershed_name, classification) %>%
    summarize(count = n_distinct(Species))
  switch4_wide <- switch4 %>% pivot_wider(names_from = classification, values_from = count)
  assign(paste0(tx, "_class.switch4"),      switch4)
  assign(paste0(tx, "_class.switch4_wide"), switch4_wide)
  #view(switch4_wide)
}
```


#----------------------------------------------------------------------------------------------
## Dissimilarity from baseline year and between watersheds (all taxa) {.tabset}

This section is for conducting dissimilarity tests between watersheds for all four taxa.

These dissimilarity tests include Jaccard and Bray-Curtis (or Gower for plants) for:
- the first year to each subsequent year of the time series, within watersheds
- across watersheds, within each year separately

Is dissimilarity increasing through time? Is the rate of composition change different
according to treatment?

**Note on plants:** Cover class is ordinal, so Bray-Curtis is not appropriate (it assumes
counts). Gower's distance is used instead as the second dissimilarity metric for plants,
because it treats variables as ordinal without assuming equal intervals between classes
(which the cover classes are not; see cover class definitions above).
Gower 0 == identical; Gower 1 == completely different.


### Step 1: Prepare per-taxon wide count/cover tables

These prep blocks remain separate because the species column differs by taxon:
Species for smammals/grasshoppers/birds; Genus_spp for plants (required for downstream
joins to pl_info).

```{r dissim_prep}
# SMAMMALS: NAs left as-is (all years sampled on all watersheds)
smammals_count_allyears <- smammals %>%
  group_by(Watershed_name, Species, Recyear) %>%
  reframe(Total = Count) %>%
  pivot_wider(names_from = "Species", values_from = "Total",
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_")

# GRASSHOPPERS: replace NA with "0" (species absent in a year have no row)
grasshoppers_count_allyears <- grasshoppers %>%
  group_by(Watershed_name, Species, Recyear) %>%
  reframe(Total = Count) %>%
  pivot_wider(names_from = "Species", values_from = "Total",
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_") %>%
  replace(is.na(.), "0")

# BIRDS: replace NA with "0" (same reason as grasshoppers)
# NOTE: Species column holds AOU code (standardized in E0)
birds_count_allyears <- birds %>%
  group_by(Watershed_name, Species, Recyear) %>%
  reframe(Total = Count) %>%
  pivot_wider(names_from = "Species", values_from = "Total",
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_") %>%
  replace(is.na(.), "0")

# PLANTS: uses Genus_spp (not Species) as names_from so output columns match pl_info joins.
# values_fn = paste (not sum) because cover class is not additive across plots.
# mutate(Count = factor(Count, levels = 1:7, ordered = TRUE)) %>%  # optional; see FIXME
plants_count_allyears <- plants %>%
  group_by(Watershed_name, Species, Recyear) %>%
  reframe(Total = Count) %>%
  pivot_wider(names_from = "Species", values_from = "Total",
              values_fn = function(x) paste(x)) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_") %>%
  replace(is.na(.), "0")
```

### Step 2: Compute dissimilarity, fit LMs, write results (shared loop)

All four taxa run through the same pipeline. The only per-taxon differences are:
- `method2`: Bray-Curtis for smammals (sm), grasshoppers (gh), birds (bd); Gower for plants (pl)
- `baseline_year`: 1992 for smammals (hardcoded); min(Recyear) for the other three

The loop produces the several named objects for downstream plotting: `sm_baseline`, `sm_jacc_diss_lm`, `sm_bray_diss_lm`, `sm_watershed`,
`gh_baseline`, `gh_bray_diss_lm`, `gh_watershed`,
`bd_baseline`, `bd_bray_diss_lm`, `bd_watershed`,
`pl_baseline`, `pl_jacc_diss_lm`, `pl_gower_diss_lm`, `pl_watershed`.

Baseline tables are renamed from `baseline.Watershed2` to `Watershed_name` so downstream code is cleaner.

```{r dissim_loop}

taxa_diss <- list(
  sm = list(
    data          = smammals_count_allyears,
    baseline_year = 1992,                    # smammals: bison present from 1992
    method2       = "bray",
    method2_col   = "baseline.bray_diss",
    prefix        = "sm",
    out           = "Datasets/E1_output_data/E1_smammals/E1_smammals"
  ),
  gh = list(
    data          = grasshoppers_count_allyears,
    baseline_year = min(grasshoppers$Recyear),
    method2       = "bray",
    method2_col   = "baseline.bray_diss",
    prefix        = "gh",
    out           = "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers"
  ),
  bd = list(
    data          = birds_count_allyears,
    baseline_year = min(birds$Recyear),
    method2       = "bray",
    method2_col   = "baseline.bray_diss",
    prefix        = "bd",
    out           = "Datasets/E1_output_data/E1_birds/E1_birds"
  ),
  pl = list(
    data          = plants_count_allyears,
    baseline_year = min(plants$Recyear),
    method2       = "gower",                 # plants: Gower instead of Bray (ordinal cover)
    method2_col   = "baseline.gower_diss",
    prefix        = "pl",
    out           = "Datasets/E1_output_data/E1_plants/E1_plants"
  )
)

for (tx in names(taxa_diss)) {
  d <- taxa_diss[[tx]]

  # compute Jaccard and method2 dissimilarity
  jacc    <- compute_dissimilarity(d$data, method = "jaccard", baseline_year = d$baseline_year)
  method2 <- compute_dissimilarity(d$data, method = d$method2, baseline_year = d$baseline_year)

  # merge baseline (year1-to-N) and watershed (pairwise within year) tables
  # rename baseline.Watershed2 -> Watershed_name here for cleaner downstream use
  baseline  <- merge(jacc[1], method2[1]) %>%
    rename(Watershed_name = baseline.Watershed2)
  watershed <- merge(jacc[2], method2[2])

  # fit linear model to Jaccard baseline dissimilarity vs time, per watershed
  # "Is Jaccard dissimilarity from year 1 increasing over time?"
  jacc_lm <- baseline %>%
    rename(value = baseline.jaccard_diss) %>%  # make generic for fit_model
    group_by(Watershed_name) %>%
    mutate(n = row_number()) %>%
    arrange(Watershed_name, n) %>%
    nest() %>%
    mutate(
      model    = map(data, fit_model),
      slope    = map_dbl(model, get_slope),
      p_value  = map_dbl(model, get_p_value),
      rsquared = map_dbl(model, get_rsq)
    ) %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)

  # fit linear model to method2 (Bray or Gower) baseline dissimilarity vs time
  method2_lm <- baseline %>%
    rename(value = !!d$method2_col) %>%  # make generic for fit_model
    group_by(Watershed_name) %>%
    mutate(n = row_number()) %>%
    arrange(Watershed_name, n) %>%
    nest() %>%
    mutate(
      model    = map(data, fit_model),
      slope    = map_dbl(model, get_slope),
      p_value  = map_dbl(model, get_p_value),
      rsquared = map_dbl(model, get_rsq)
    ) %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)

  # write results to file
  write.csv(baseline,   paste0(d$out, "_dissimilarity_year1toN.csv"))
  write.csv(watershed,  paste0(d$out, "_dissimilarity_watersheds.csv"))
  write.csv(jacc_lm,    paste0(d$out, "_jacc_diss_lm.csv"))
  write.csv(method2_lm, paste0(d$out, "_", d$method2, "_diss_lm.csv"))

  # assign named objects for downstream plotting code
  assign(paste0(d$prefix, "_baseline"),                    baseline)
  assign(paste0(d$prefix, "_watershed"),                   watershed)
  assign(paste0(d$prefix, "_jacc_diss_lm"),                jacc_lm)
  assign(paste0(d$prefix, "_", d$method2, "_diss_lm"),     method2_lm)
}
```

#----------------------------------------------------------------------------------------

## Species Richness change over time (all taxa) {.tabset}
This section applies a function to calculate species richness per watershed and per year. 
It then determines if richness is changing directionally over time using a simple linear model.

**input:** the presence absence matrix and the watershed-species information. Both of these come from previous code steps (using decostand and creating the env dataframe)
**returns:** a list with both dataframes to result -- contains summary richness data and the linear model results

#### Small mammals

```{r smammals richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_smammals_pa, smammals_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_smammals/E1_smammals_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_smammals/E1_smammals_richness_lm.csv")
```

#### Grasshoppers

```{r grasshopper richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_grasshoppers_pa, grasshoppers_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_richness_lm.csv")
```

#### Birds

```{r bird richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_birds_pa, birds_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_birds/E1_birds_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_birds/E1_birds_richness_lm.csv")
```

#### Plants

```{r plants richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_plants_pa, plants_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_plants/E1_plants_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_plants/E1_plants_richness_lm.csv")
```

#----------------------------------------------------------------------------------------

## Abundance change over time (all taxa) {.tabset}
This section determines if abundance is changing directionally over time using a simple linear model, using the count data for each taxon. While abundance fluctuates in a non-linear fashion for most species, this is meant to suggest which species have more apparent trends in increasing or decreasing abundance (i.e. it will not detect random or cyclical changes or particular extremes).

**For plant taxa** We applied an ordinal logistic regression instead (see separate chunk below). This model works with ordinal abundance, where we can't assume the intervals between classes are all equal. The result models the probability of being in a higher or lower cover class as a function of year. "As year increases, is the plant more likely to be in a higher cover class?"

Each section writes a csv file summarizing the abundance information and model results.

#### Small-mammals, Grasshoppers, Birds; count data

All three taxa now share the same pipeline:
- Remove env columns, gather year columns long, fit lm per Species.watershed
- `separate(Species.watershed, c("Species", "Watershed_name"))` — works for all three
  because Species.watershed is now built as `Genus_species.Watershed_name` (one `.`
  separator) for smammals/grasshoppers, and `AOUCODE.Watershed_name` for birds
- The original smammals/grasshoppers chunks used `separate` into three parts
  (`Genus`, `Spp`, `Watershed_name`) followed by `unite("Species", Genus:Spp)`.
  That was written when the species column had a space separator. Now that
  `Species = Genus_species` (underscore joined in E0), the two-part separate is
  correct and produces identical Species values.
- The original birds chunk had two bugs: `dplyr::select(-AOU_code)` (column renamed
  to Species in E0) and a dangling `unite("Species", Genus:Spp)` after a `#` comment
  that would error if uncommented. Both are fixed here.

```{r abund_lm_smammals_grasshoppers_birds}

taxa_abund <- list(
  sm = list(
    years_wide = smammals_years_wide,
    out        = "Datasets/E1_output_data/E1_smammals/E1_smammals",
    lm_obj     = "sm_spabun_lm"
  ),
  gh = list(
    years_wide = grasshoppers_years_wide,
    out        = "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers",
    lm_obj     = "gh_spabun_lm"
  ),
  bd = list(
    years_wide = birds_years_wide,
    out        = "Datasets/E1_output_data/E1_birds/E1_birds",
    lm_obj     = "bd_spabun_lm"
  )
)

for (tx in names(taxa_abund)) {
  d <- taxa_abund[[tx]]

  #make a version of the wide dataset without the extra env columns
  spabund <- d$years_wide %>%
    dplyr::select(-Watershed_name, -Species)

  #Use 'purrr' to run the model. The gather function gets the values we need, and
  #   group_by ensures values are grouped by Species.watershed.
  #   'n' is the row number, which translates to the year (year1 = 1, year2 = 2...).
  #   arrange orders the data by Species.watershed and row number.
  #   nest nests all the data (year/row number, and count) into a single column,
  #   while each row is a species-watershed combination.
  #   The rest of the code runs the linear models for each row and extracts outputs.
  spabun_lm <- spabund %>%
    gather('year', 'value', -Species.watershed) %>%
    group_by(Species.watershed) %>%
    mutate(n = row_number()) %>%
    arrange(Species.watershed, n) %>%
    nest() %>%
    mutate(model    = map(data, fit_model)) %>%
    mutate(slope    = map_dbl(model, get_slope)) %>%
    mutate(p_value  = map_dbl(model, get_p_value)) %>%
    mutate(rsquared = map_dbl(model, get_rsq)) %>%
    # remove detailed model results before writing to file
    dplyr::select(Species.watershed, slope, p_value, rsquared) %>%
    separate(Species.watershed, c("Species", "Watershed_name"), sep = "\\.")

  # Save abundance data and model results
  write.csv(d$years_wide, paste0(d$out, "_abundance.csv"))
  write.csv(spabun_lm,    paste0(d$out, "_abundance_lm.csv"))

  # assign named object for downstream use
  assign(d$lm_obj, spabun_lm)
}
```


#### Plants; percent cover data
Here, we use the `polr` function for ordinal logistic regression instead of a linear
model. Some key notes: 
* pseudo p-value calculated from t-value of the year coefficient
* McFadden's pseudo R2 calculated from null model comparison
* model convergence failure is somewhat common; handled through NULL and NA values
* low variation or no variation in ordinal response values is sometimes a problem
* if n perfectly predicts class it is a problem
* overfitting could be an issue

```{r plants abundance}
#make a version of the wide dataset without the extra columns, for the logistic model
pl_spabund <- plants_years_wide %>% 
  dplyr::select(-Watershed_name, -Species)

# Run ordinal logistic regression models for each row (Species-watershed) and save the results
pl_spabun_logit <- pl_spabund %>%
  pivot_longer(cols = -Species.watershed, 
               names_to = "year",
               values_to = 'value') %>%
  group_by(Species.watershed) %>%
  mutate(n = row_number(),
         value = factor(value, levels = 0:7, ordered = TRUE)) %>% #ensure ordered factor
  arrange(Species.watershed, n) %>%
  nest() %>%
  mutate(model = map(data, fit_ordinal_model), 
         model_null = map(data, fit_ordinal_null),
         slope = map_dbl(model, get_polr_slope),
         psuedo_pval = map_dbl(model, get_polr_p),
         pseudo_rsq = map2_dbl(model, model_null, get_polr_rsq)) %>%
  # remove detailed model results before writing to file
  dplyr::select(Species.watershed, slope, psuedo_pval, pseudo_rsq) %>%
  separate(Species.watershed, into = c("Species", "Watershed_name"), sep="\\.")
  
#Save abundance data 
write.csv(plants_years_wide, "Datasets/E1_output_data/E1_plants/E1_plants_abundance.csv")

#Save abundance model results
write.csv(pl_spabun_logit, "Datasets/E1_output_data/E1_plants/E1_plants_abundance_logit.csv")
```


#-----------------------------------------------------------------------------------------
# STATISTICAL COMPARISONS FOR INCIDENCE and ABUNDANCE CLASSIFICATION

prepare INCIDENCE data for statistical comparison
```{r}
#prepare the INCIDENCE data for analysis
smammal_class_counts <- smammals_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(smammals_classification$Species)))) %>%
  mutate(taxa = "small mammals")

grasshoppers_class_counts <- grasshoppers_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(grasshoppers_classification$Species)))) %>%
  mutate(taxa = "grasshoppers")

birds_class_counts <- birds_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(birds_classification$AOU_code)))) %>%
  mutate(taxa = "birds")

plants_class_counts <- plants_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(plants_classification$Genus_spp)))) %>%
  mutate(taxa = "plants")

#rbind all the class summary datasets together
all_class_counts <- rbind(smammal_class_counts, grasshoppers_class_counts, birds_class_counts, plants_class_counts)

# make classification, watershed, and taxa ordered factors
# remove non-repeated watersheds
# remove absent classification (won't be compared)
all_class_counts <- all_class_counts %>%
  filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B")) %>%
  filter(!classification == "No_change-absent") %>% 
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         taxa = factor(taxa, levels = c("plants", "grasshoppers", "small mammals", "birds"),
                       ordered = TRUE),
         classification = factor(classification, 
                                 levels = c("No_change-present", "Rare",
                                            "Increasing", "Decreasing",
                                            "Recurrent", "Clumped", "Random"), 
                                 ordered = TRUE))
```

prepare ABUNDANCE data for statistical comparison
```{r}
#prepare the INCIDENCE data for analysis
smammal_class_abun <- smammals_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(smammals_abun_classification$Species)))) %>%
  mutate(taxa = "small mammals")

grasshoppers_class_abun <- grasshoppers_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(grasshoppers_abun_classification$Species)))) %>%
  mutate(taxa = "grasshoppers")

birds_class_abun <- birds_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(birds_abun_classification$AOU_code)))) %>%
  mutate(taxa = "birds")

plants_class_abun <- plants_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(plants_abun_classification$Genus_spp)))) %>%
  mutate(taxa = "plants")

#rbind all the class summary datasets together
all_class_abun <- rbind(smammal_class_abun, grasshoppers_class_abun, birds_class_abun, plants_class_abun)

# make classification, watershed, and taxa ordered factors
all_class_abun <- all_class_abun %>%
  filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         taxa = factor(taxa, 
                       levels = c("plants", "grasshoppers", "small mammals", "birds"),
                       ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("Invariant", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Clumped", "Random"), 
                                 ordered = TRUE))
```

### 1. Are classifications significantly different across the watershed treatments?

Apply Chi-square INCIDENCE comparison across watersheds for each taxa
```{r}
# run chi-square comparison across watersheds for each taxa 
chisq_results <- all_class_counts %>%
  dplyr::select(-Proportion) %>%
  group_by(taxa) %>%
  nest() %>%
  mutate(chisq = map(data, run_chisq_from_nested)) %>%
  unnest(chisq)

# add Bonferroni adjustments to p-value
chisq_results <- chisq_results %>%
  mutate(p_bonf = ifelse(!is.na(p_value), p.adjust(p_value, method = "bonferroni"), NA))

write_csv(chisq_results, "Datasets/E1_output_data/chisq_results_by_taxa.csv")
```

Apply Chi-square ABUNDANCE comparison across watersheds for each taxa
```{r}
# run chi-square comparison across watersheds for each taxa 
chisq_results_abun <- all_class_abun %>%
  dplyr::select(-Proportion) %>%
  group_by(taxa) %>%
  nest() %>%
  mutate(chisq = map(data, run_chisq_from_nested)) %>%
  unnest(chisq)

# add Bonferroni adjustments to p-value
chisq_results_abun <- chisq_results_abun %>%
  mutate(p_bonf = ifelse(!is.na(p_value), p.adjust(p_value, method = "bonferroni"), NA))

write_csv(chisq_results_abun, "Datasets/E1_output_data/chisq_abun_results_by_taxa.csv")
```

**ANSWER: For most taxa NO. For plant taxa YES.** 

Plants do show some changes in classification frequency across treatments. 

The hypothesis regarding if recurrent/clumped classification is more represented in 4-year burns is not supported. Does the recurrent category represent what we hope it does, or is it too simplistic? Or are recurrent cycles not really apparent at this site/treatment?

The hypothesis regarding if press disturbance (grazing) results in more directed change (increasing or decreasing) species is somewhat supported. Sites where native grazers are absent always are underrepresented by increasing taxa and over-represented by decreasing taxa compared to sites where native grazers have been added (press treatment). The direct effect of bison is on the "directed change" category, supporting a press disturbance impact showing up in these classifications.


### 1a. For taxa that showed significant shifts in classification frequency, what is driving that change?

For incidence classification: only plants demonstrated a significant difference (small cramers V)
For abundance classification: only plants demonstrated a significant difference (small cramers V)

Look at the INCIDENCE residuals
```{r}
# prepare data for plant residuals
pl_resid <- chisq_results %>%
  filter(taxa == "plants") %>%
  pull(residuals) %>%
  .[[1]] %>%
  pivot_longer(-classification, names_to = "Watershed_name", values_to = "Residual") %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Clumped", "Random"), 
                                 ordered = TRUE))

# heatmap of the residuals
ggplot(pl_resid, aes(x = Watershed_name, y = classification, fill = Residual)) +
  #geom_point(aes(color = Residual, size = abs(Residual))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red4", mid = "white", high = "navyblue", 
                       midpoint = 0, limits = c(-max(abs(pl_resid$Residual)), 
                                                max(abs(pl_resid$Residual))),
                       name = "Std Residual") +
  labs(title = "Plant standardized residuals", 
       x = "Watershed", 
       y = "Incidence Classification") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plant_incidence_chisq_residuals.png", dpi=300, width=5, height=4, bg="white")
```

The same plot for chisq residuals on comparing incidence classification, but using points instead of filled cells. More extreme residuals >|2| are shown with a black outline to help them stand out. 
```{r}
ggplot(pl_resid, aes(x = Watershed_name, y = classification)) +
  
  # Main points: size by abs(Residual), color by Residual
  geom_point(aes(size = abs(Residual), color = Residual)) +
  
  # Border points: only where abs(Residual) > 2
  geom_point(
    data = pl_resid %>% filter(abs(Residual) > 2),
    shape = 21, stroke = 1, fill = NA, color = "black",
    aes(size = abs(Residual))
  ) +

  scale_color_gradient2(
    low = "red4", mid = "white", high = "navyblue",
    midpoint = 0,
    limits = c(-max(abs(pl_resid$Residual)), max(abs(pl_resid$Residual))),
    name = "Std Residual"
  ) +
  scale_size_continuous(range = c(1, 10), name = "Abs(Residual)") +
  
  labs(
    title = "Plant standardized residuals",
    x = "Watershed",
    y = "Incidence Classification"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Figures/plant_incidence_chisq_residuals_points_2plus.png", dpi=300, width=5, height=4, bg="white")

```



Look at the ABUNDANCE residuals
```{r}
# prepare data for plant residuals
pl_resid_abun <- chisq_results_abun %>%
  filter(taxa == "plants") %>%
  pull(residuals) %>%
  .[[1]] %>%
  pivot_longer(-classification, names_to = "Watershed_name", values_to = "Residual") %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("Invariant", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Clumped", "Random"), 
                                 ordered = TRUE))

# heatmap of the residuals
ggplot(pl_resid_abun, aes(x = Watershed_name, y = classification, fill = Residual)) +
  #geom_point(aes(color = Residual, size = abs(Residual))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red4", mid = "white", high = "navyblue", 
                       midpoint = 0, limits = c(-max(abs(pl_resid_abun$Residual)), 
                                                max(abs(pl_resid_abun$Residual))),
                       name = "Std Residual") +
  labs(title = "Plant standardized residuals", 
       x = "Watershed", 
       y = "Abundance Classification") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plant_abundance_chisq_residuals.png", dpi=300, width=4, height=4)
```

#-----------------------------------------------------------------------------------------
# DATA VISUALIZATIONS

Let's make some descriptive plots of the species classifications. We can make violin plots that will mirror the ones in Gotelli et al. 2022. 

### Incidence Classification
```{r}
ggplot(all_class_counts, aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") + 
  labs(title = "Frequency across watersheds", y = "Incidence Classification") + 
  facet_wrap(~taxa, ncol=1)

ggsave("Figures/all_taxa_class_incid_freq.png", dpi=300, width=5, height=10)
```

Just for plants, since they were the only ones that were significantly different across watersheds
```{r}
pl <- all_class_counts %>%
  filter(taxa == "plants")

#lollipop chart (probably won't use this one in favor of bar chart)
# ggplot(pl, aes(x = Proportion, y = classification)) +
#   geom_segment(aes(x = 0, xend = Proportion, 
#                    y = classification, yend = classification),
#                color = "grey50", linewidth = 0.8) +
#   geom_point(aes(color=classification), size = 3) +
#   facet_wrap(~Watershed_name, ncol=2) +
#   labs(title = "Plant incidence",
#        x = "Proportion of species",
#        y = "Classification") +
#   theme_minimal() +
#   theme(strip.text = element_text(face = "bold"),
#         axis.text.y = element_text(size = 10),
#         legend.position = "none")

#bar graph
ggplot(pl, aes(x = Proportion, y = classification)) +
  geom_col(aes(fill = classification)) + 
  facet_wrap(~Watershed_name, ncol=2) +
  labs(title = "Plant incidence",
       x = "Proportion of species",
       y = "Classification") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

ggsave("Figures/plant_class_incid_freq.png", dpi=300, width=4, height=6)
```


### Abundance Classification
```{r}
ggplot(all_class_abun, aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") + 
  labs(title = "Frequency across watersheds", y = "Abundance Classification") + 
  facet_wrap(~taxa, ncol=1)

ggsave("Figures/all_taxa_class_abun_freq.png", dpi=300, width=5, height=10)
```

Just for plants, since they were the only ones that were significantly different across watersheds
```{r}
pl_abun <- all_class_abun %>%
  filter(taxa == "plants")

#bar graph
ggplot(pl_abun, aes(x = Proportion, y = classification)) +
  geom_col(aes(fill = classification)) + 
  facet_wrap(~Watershed_name, ncol=2) +
  labs(title = "Plant abundance class",
       x = "Proportion of species",
       y = "Classification") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

ggsave("Figures/plant_class_abun_freq.png", dpi=300, width=4, height=6)
```

### Composition
Since incidence change is necessarily impacted by species entering and exiting a community in different years, let's look at the underlying species composition patterns at the site. 

Let's focus on plants, since that's where the incidence classes were most different
```{r}
## PLANTS Jaccard

# process relevant data frames
pl_baseline <- pl_baseline %>% rename(Watershed_name=baseline.Watershed2)

pl_diss <- left_join(pl_baseline, pl_jacc_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(pl_diss, aes(x = baseline.Year2, y = baseline.jaccard_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=pl_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Jaccard dissimilarity to Year 1") +
  labs(title = "plant Jaccard trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_jacc_years.png", dpi=300, width=4, height=6)
```

#FIXME: baseline.Watershed2 doesn't exist. fix
```{r}
## PLANTS Gower

# process relevant data frames
pl_baseline <- pl_baseline %>% rename(Watershed_name=baseline.Watershed2)

pl_diss <- left_join(pl_baseline, pl_gower_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(pl_diss, aes(x = baseline.Year2, y = baseline.gower_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=pl_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Gower dissimilarity to Year 1") +
  labs(title = "plant Gower trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_gower_years.png", dpi=300, width=4, height=6)
```

Let's dig a bit deeper into the directed change category. 
**Which plant species are being classified as directed change?**
```{r}
pl_class <- plants_classification %>%
  #ensure just the watersheds for this study are kept
    filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B"))

pl_directed <- pl_class %>%
  filter(classification %in% c("Decreasing", "Increasing")) %>%
  dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification)

directed_spp <- unique(pl_directed$Genus_spp)

# How many species were increasing in at least one watershed?
pl_directed %>%
  filter(classification == "Increasing") %>%
  summarize(n_distinct(Genus_spp))

#How many species were decreasing in at least one watershed? 
pl_directed %>%
  filter(classification == "Decreasing") %>%
  summarize(n_distinct(Genus_spp))
```

**Which species are increasing at all watersheds in which they occur?**
```{r}
# Species consistently increasing
inc_species <- pl_class %>%
  group_by(Genus_spp) %>%
  filter(all(classification == "Increasing")) %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    Watershed_count = n_distinct(Watershed_name),
    .groups = "drop"
  )

#View(inc_species)
all_inc = unique(inc_species$Genus_spp)

nrow(inc_species)
unique(inc_species$Genus_spp)
```
There are 22 species that are increasing at all watersheds in which they occur. 15/22 of these only occur in grazed sites, indicating a colonization effect for these sites, potentially in response to the grazing treatment. 

**Which species are decreasing at all watersheds in which they occur?**
```{r}
# Species consistently decreasing
dec_species <- pl_class %>%
  group_by(Genus_spp) %>%
  filter(all(classification == "Decreasing")) %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    Watershed_count = n_distinct(Watershed_name),
    .groups = "drop"
  )

all_dec = unique(dec_species$Genus_spp)

dim(dec_species)
```
Answer: There are NO species that were decreasing at all watersheds in which they occur. 

**Which are increasing or decreasing only in the grazed watersheds, but not in the ungrazed?**
-- I also need to make a new subset of the data that includes the other classifications for comparison
INCREASING
```{r}
# Species increasing across grazed/ungrazed treatments
inc_species <- pl_class %>%
    mutate(treatment = case_when(
    Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed",
    TRUE ~ "ungrazed")) %>%
  group_by(Genus_spp) %>%
  filter(classification == "Increasing") %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    trmts = paste(sort(unique(treatment)), collapse = ", "),
    trmt_count = n_distinct(treatment),
    .groups = "drop"
  ) %>%
  arrange(trmt_count)
  
View(inc_species)
```
DECREASING
```{r}
# Species decreasing across grazed/ungrazed treatments
dec_species <- pl_class %>%
    mutate(treatment = case_when(
    Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed",
    TRUE ~ "ungrazed")) %>%
  group_by(Genus_spp) %>%
  filter(classification == "Decreasing") %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    trmts = paste(sort(unique(treatment)), collapse = ", "),
    trmt_count = n_distinct(treatment),
    .groups = "drop"
  ) %>%
  arrange(trmt_count)
  
View(dec_species)
```

**What else did they switch from in the watersheds where they were not in the directed change classes?**
```{r}
# Species increasing across grazed/ungrazed treatments
dir_species <- pl_class %>%
    mutate(treatment = case_when(
      Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed", TRUE ~ "ungrazed")) %>%
  filter(Genus_spp %in% directed_spp) %>%
    filter(!Genus_spp %in% all_inc) %>%
    dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    arrange(Genus_spp, Watershed_name)

View(dir_species)

```


**Let's look a little deeper at the plant species traits across categories**
How do growth form(a,b,p), life form (g,f,m), and photosynthesis (c3, c4) vary across the categories?
```{r}
# prepare the species info
pl_info <- pl_info %>%
  mutate(Genus_spp = paste(gen, spec, sep = "_")) %>%
  distinct()

#add traits back into the plant classification dataset
pl_class <- left_join(pl_class, pl_info, by="Genus_spp")
```

**Does origin (native, invasive) differ across the classifications?**
```{r}
# make a contingency table of origin and classification
table_or <- table(pl_class$classification, pl_class$origin)

#also look at the results separate by watershed -- 
#     probably grazed 4-year and grazed 20-year burns most invaded w/ rare and increasing i spp
table_or_w <- table(pl_class$classification, pl_class$origin, pl_class$Watershed_name)

chisq_result <- chisq.test(table_or)

#look at residuals
chisq_result$stdres

corrplot(chisq_result$stdres, is.cor = FALSE, tl.col = "black")

#post hoc test to see what is differing; Pairwise Fishers exact tests
# Get all unique pairs of classification categories
categories <- rownames(table_or)
pair_comparisons <- combn(categories, 2, simplify = FALSE)

# Run Fisher's exact test for each pairwise row comparison
posthoc_results <- map_dfr(pair_comparisons, function(pair) {
  sub_table <- table_or[pair, ]  # 2xN table
  test <- fisher.test(sub_table)
  
  tibble(
    group1 = pair[1],
    group2 = pair[2],
    p_value = test$p.value
  )
})

# adjust for multiple tests
posthoc_results <- posthoc_results %>%
  mutate(p_adj = p.adjust(p_value, method = "bonferroni")) %>%
  filter(p_adj < 0.05) %>%
  arrange(p_adj)
```

Plot the residuals for the growth form comparison across classifications
```{r}
#plot a different way
resid_df <- melt(chisq_result$stdres, varnames = c("classification", "Origin"), value.name = "Residual") %>%
  mutate(classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE))

ggplot(resid_df, aes(x = Origin, y = classification)) +
  # Main points: size and color based on residual
  geom_point(aes(size = abs(Residual), color = Residual)) +
    # Color scale for positive/negative residuals
  scale_color_gradient2(low = "red4", mid = "white", high = "navyblue", midpoint = 0,
    limits = c(-max(abs(resid_df$Residual)), max(abs(resid_df$Residual))),
    name = "Std Residual") +
  # Black border for large residuals (>2)
  geom_point(data = resid_df %>% filter(abs(Residual) > 2),
    aes(size = abs(Residual)),
    shape = 21, stroke = 1, fill = NA, color = "black") +
  # Size scale
  scale_size_continuous(range = c(1, 10), name = "Abs(Residual)") +
  # Labels and theme
  labs(title = "Plant standardized residuals",
    x = "Origin", y = "") +
  theme_minimal() #+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plants_incidence_origin-resid.png", dpi=300, width=4, height=4, bg="white")
```

**Does growth form (annual, biannual, perennial) differ across the classifications?**
```{r}
# make a contingency table of growth form and classification
table_cf <- table(pl_class$classification, pl_class$`growth form`)

chisq_result <- chisq.test(table_cf)

#look at residuals
chisq_result$stdres

corrplot(chisq_result$stdres, is.cor = FALSE, tl.col = "black")

#post hoc test to see what is differing; Pairwise Fishers exact tests
# Get all unique pairs of classification categories
categories <- rownames(table_cf)
pair_comparisons <- combn(categories, 2, simplify = FALSE)

# Run Fisher's exact test for each pairwise row comparison
posthoc_results <- map_dfr(pair_comparisons, function(pair) {
  sub_table <- table_cf[pair, ]  # 2xN table
  test <- fisher.test(sub_table)
  
  tibble(
    group1 = pair[1],
    group2 = pair[2],
    p_value = test$p.value
  )
})

# adjust for multiple tests
posthoc_results <- posthoc_results %>%
  mutate(p_adj = p.adjust(p_value, method = "bonferroni")) %>%
  filter(p_adj < 0.05) %>%
  arrange(p_adj)
```

Plot the residuals for the growth form comparison across classifications
```{r}
#plot a different way
resid_df <- melt(chisq_result$stdres, varnames = c("classification", "GrowthForm"), value.name = "Residual") %>%
  mutate(classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE))

ggplot(resid_df, aes(x = GrowthForm, y = classification)) +
  # Main points: size and color based on residual
  geom_point(aes(size = abs(Residual), color = Residual)) +
    # Color scale for positive/negative residuals
  scale_color_gradient2(low = "red4", mid = "white", high = "navyblue", midpoint = 0,
    limits = c(-max(abs(resid_df$Residual)), max(abs(resid_df$Residual))),
    name = "Std Residual") +
  # Black border for large residuals (>2)
  geom_point(data = resid_df %>% filter(abs(Residual) > 2),
    aes(size = abs(Residual)),
    shape = 21, stroke = 1, fill = NA, color = "black") +
  # Size scale
  scale_size_continuous(range = c(1, 10), name = "Abs(Residual)") +
  # Labels and theme
  labs(title = "Plant standardized residuals",
    x = "Growth Form", y = "") +
  theme_minimal() #+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plants_incidence_growth-form-resid.png", dpi=300, width=4, height=4, bg="white")
```

**Does growth form differ across the watersheds by classification?**
```{r}
watersheds <- unique(pl_class$Watershed_name)

# Run chi-squared + post hoc test per watershed
results_by_watershed <- map(watersheds, function(ws) {
  # Subset data
  df_ws <- pl_class %>% filter(Watershed_name == ws)
  
  # Create contingency table
  tab <- table(df_ws$classification, df_ws$`growth form`)
  
  # Skip if table is too small
  if (nrow(tab) < 2 || ncol(tab) < 2) return(NULL)
  
  # Chi-squared test
  chi <- chisq.test(tab)
  
  # Pairwise Fisher tests
  classes <- rownames(tab)
  pairs <- combn(classes, 2, simplify = FALSE)
  
  posthoc <- map_dfr(pairs, function(pair) {
    test <- fisher.test(tab[pair, ])
    tibble(
      group1 = pair[1],
      group2 = pair[2],
      p_value = test$p.value
    )
  }) %>%
    mutate(p_adj = p.adjust(p_value, method = "bonferroni")) %>%
    arrange(p_adj)
  
  # Store results
  list(
    watershed = ws,
    chisq_p = chi$p.value,
    chisq_resid = chi$stdres,
    posthoc = posthoc
  )
})

#view top significant post hoc comparisons
map_dfr(results_by_watershed, function(res) {
  if (is.null(res)) return(NULL)
  res$posthoc %>%
    filter(p_adj < 0.05) %>%
    mutate(watershed = res$watershed)
})

#view residuals
for (res in results_by_watershed) {
  if (is.null(res)) next
  corrplot(res$chisq_resid, is.cor = FALSE,
           tl.col = "black",
           title = paste("Std Residuals:", res$watershed),
           mar = c(0, 0, 2, 0))
}
```





Can we annotate directed species with known traits like growth form, photosynthesis, and annual/perennial?
```{r}
#add details back in for plants that are directed change
dir_species <- left_join(dir_species, pl_info, by="Genus_spp")

dir_spp_info <- dir_species %>%
  dplyr::select(Genus_spp, genus, species, family, `growth form`, `life form`, origin) %>%
  distinct()

#check out some summaries (note that if it doesn't add up to 146, is NA)
table(dir_spp_info$origin)
table(dir_spp_info$`growth form`)
table(dir_spp_info$`life form`)
```
#FIXME: I am not sure what to do with the above results... How to summarize better without getting into the weeds?


**how does classification relate to percentage of years present?**
```{r}
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  ggplot(aes(x = classification, y = percyears)) +
  geom_jitter(width = 0.2, height = 0, color = "steelblue", alpha = 0.6) +
  geom_violin(alpha=0.7, aes(fill=classification)) +
  labs(title = "Percent of Years by Classification",
       x = "Classification",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() + 
  guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_class_percyears.png", dpi=300, width=4, height=4, bg="white")
```

**how does proportion of years present relate to watershed?**
```{r}
#boxplot
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  mutate(treatment = case_when(
      Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed", TRUE ~ "ungrazed")) %>%
  ggplot(aes(x = Watershed_name, y = percyears, fill=Watershed_name)) +
  geom_boxplot(alpha=0.7, notch=TRUE) +
  scale_fill_manual(values = c("darkolivegreen", "sienna4", "darkolivegreen4", 
                               "sienna3", "darkolivegreen3", "sienna2")) +
  labs(title = "Percent of Years by Watershed",
       x = "Watershed",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() +
  guides(fill="none") +
  theme(legend.position="none")

ggsave("Figures/plants_watershed_percyears_bxp.png", dpi=300, width=4, height=4, bg="white")
```

```{r}
#violin + jitter plot
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  ggplot(aes(x = Watershed_name, y = percyears, group=Watershed_name)) +
  geom_violin(alpha=0.7) +
  geom_jitter(aes(col=classification), width = 0.2, height = 0, alpha = 0.8) +
  labs(title = "Percent of Years by Watershed",
       x = "Watershed",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() 

ggsave("Figures/plants_watershed_percyears.png", dpi=300, width=4, height=4, bg="white")
```

**how does life form relate to percentage of years present?**
#FIXME: This plot only shows those that are increasing or decreasing on at least some plots; not all species
```{r}
dir_species %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE),
           growth_form = factor(`growth form`, 
                                levels = c("a", "b", "p"),
                                ordered = TRUE)) %>%
  ggplot(aes(x = growth_form, y = percyears)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6) +
  geom_boxplot(alpha=0.5, aes(fill=growth_form), notch=TRUE) +
  labs(title = "Percent of Years by Growth Form",
       x = "Growth Form",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() + 
  guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_dir-change-ann-per_percyears.png", dpi=300, width=4, height=4, bg="white")
```

**for species that were directed change in a particular watershed, how are they divided out across watershed and across growth form?**
```{r}
dir_species %>%
  filter(classification %in% c("Increasing", "Decreasing")) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("Decreasing", "Increasing"), 
                                 ordered = TRUE),
           growth_form = factor(`growth form`, 
                                levels = c("a", "b", "p"),
                                ordered = TRUE)) %>%
  ggplot(aes(x = growth_form, y = percyears)) +
  geom_boxplot(alpha=0.5) +
  #geom_violin() +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6, aes(col=classification)) +
  labs(title = "Directed change populations",
       x = "Growth Form",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  facet_wrap(~Watershed_name, ncol=2) +
  theme_bw()# + 
  #guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_dir-change-ann-per__watershed_percyears.png", dpi=300, width=4, height=6)
```

#----------------------------------------------------------

```{r}
## SMAMMALS Jaccard

# process relevant data frames
sm_baseline <- sm_baseline %>% rename(Watershed_name=baseline.Watershed2)

sm_diss <- left_join(sm_baseline, sm_jacc_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(sm_diss, aes(x = baseline.Year2, y = baseline.jaccard_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=sm_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Jaccard dissimilarity to Year 1") +
  labs(title = "small mammal Jaccard trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/smammals_jacc_years.png", dpi=300, width=4, height=6)
```

```{r}
## SMAMMALS Bray

# process relevant data frames
sm_diss <- left_join(sm_baseline, sm_bray_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(sm_diss, aes(x = baseline.Year2, y = baseline.bray_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=sm_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Bray-Curtis dissimilarity to Year 1") +
  labs(title = "small mammal Bray trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/smammals_bray_years.png", dpi=300, width=4, height=6)
```
#TODO: START HERE, make composition plots like this for all taxa; will need to jump up to composition section and add lm models and writecsv to all other taxa too. 


#TODO: START HERE AGAIN

# TODO: START IN THE PLOTTING CODE AND ADD THIS TYPE OF PLOT FOR RICHNESS AND ABUNDANCE
#     LINES ONLY SHOW WHEN THE PVALUE IS SIGNIFICANT, MAKES IT EASIER TO SCAN; 
#     filter out extra watersheds
#     add label for trend direction

###Plant richness
```{r}
# process relevant data frames
rich_p <- left_join(result$richness, result$model, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))

  
# make panel plot of results 
#(check that plants were the most recent richness result run)
ggplot(rich_p, aes(x = Recyear, y = Richness)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=rich_p %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  labs(title = "Plant richness trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_richness_years.png", dpi=300, width=4, height=6)
```

###Plant abundance
```{r}
# identify the plant species that had significant cover class trends, using pseudo pvalue
#   and add trend direction (positive or negative)
pl_spabun_logit_sig <- pl_spabun_logit %>% 
  filter(psuedo_pval <= 0.05) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))

#show results divided by positive and negative significant trends
table(pl_spabun_logit_sig$Watershed_name, pl_spabun_logit_sig$trend_dir)

# count number of significant increasing and decreasing trends per watershed
trend_counts <- pl_spabun_logit_sig %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  count(Watershed_name, trend_dir) %>%
  pivot_wider(
    names_from = trend_dir,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(label = paste0("↑ ", positive, "  ↓ ", negative))

#annotate abundance df with logit model results for significant species-watersheds
pl_abun <- left_join(pl_spabun_logit_sig, plants_years_wide, 
                     by=c("Species" = "Genus_spp", "Watershed_name" = "Watershed_name")) %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE))

#pivot longer for plotting
# Pivot all year columns into long format
pl_abun_sig_long <- pl_abun %>%
  pivot_longer(
    cols = matches("^\\d{4}$"),  # columns named like 1992, 1993...
    names_to = "Year",
    values_to = "Cover_class"
  ) %>%
  mutate(Year = as.numeric(Year))  # make Year numeric if needed

#add the trend info in for plotting
pl_abun_sig_long <- pl_abun_sig_long %>%
  left_join(trend_counts, by = "Watershed_name")


#make plot of data
ggplot(pl_abun_sig_long, aes(x = Year, y = Cover_class, group = Species, color = trend_dir)) +
  geom_smooth(method = "lm", se = FALSE, size = 0.4, alpha=0.5) +  # no points, just lines
    # Add annotation per facet
  geom_text(data = trend_counts, aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.1, vjust = 1.1, size = 3) +
  facet_wrap(~Watershed_name, ncol=2) +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +
  labs(
    title = "Significant species Cover class Trends",
    x = "Year",
    y = "Cover Class",
    color = "Trend") +
  theme_bw() +
  theme(strip.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none")

ggsave("Figures/plants_signif_abundance_years.png", bg="white", dpi=300, width=4, height=6)

```



