---
title: "E1_AllTaxa_Analysis"
author: "Maya Parker-Smith"
date: "2023-11-13"
output: 
 html_document:
    smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---

Creates tabs in the HTML document \# {.tabset .tabset-pills .tabset-fade}

```{r setup, include=FALSE}
#Clear environment
rm(list=ls())

#Read in necessary libraries
library(tidyverse)
library(boot)
library(broom)
library(corrplot)
library(cowplot)
library(ggpubr)
library(lme4)
library(MASS)
library(multcomp)
library(purrr)
library(readxl)
library(tseries)
library(vegan)

#Call our custom functions for classification and analysis
source("classify_functions.R")
```

#Read in Data Tables This includes the watersheds tables and the species-level data for the 4 taxa.

```{r}
watershed_info <- read_excel("Datasets/Watershed_data/Watershed_info.xlsx")

#Reading in the cleaned taxa data sets
E0_smammals <- read_csv("Datasets/E0_cleaned_data/E0_smammals.csv")
E0_grasshoppers <- read_csv("Datasets/E0_cleaned_data/E0_grasshoppers.csv")
E0_birds<- read_csv("Datasets/E0_cleaned_data/E0_birds.csv")
E0_plants <- read_csv("Datasets/E0_cleaned_data/E0_plants.csv")

#read in plant species list
pl_info <- read_csv("Datasets/Raw_data/Plants/plant_sp_list.csv")
```

# Clean Data to Classified Incidence for All Taxa

This Rmarkdown is for taking the clean datasets for all 4 taxa (small-mammals, grasshoppers, birds, and plants) and using the incidence classification method to categorize each species. This file also estimates composition change using presence-absence information (Jaccard dissimilarity) and abundance information (Bray-Curtis dissimilarity), richness change (linear model), and abundance change (linear model) over time.

## Brief overview of Classification

Incidence Classification categories include:

-   **"No_change-absent"** \<- Species was detected at other surrounding watersheds, but it was absent at specified watershed for the duration of the time series.
-   **"Rare"** \<- Species was present at specified watershed for between 1-10% of years in the duration of the time series.
-   **"No_change-present"** \<- Species at watershed is present in \>=90% of years in which data was collected.
-   **"Increasing"** \<- The presence of the species increases throughout the duration of the study (it is absent during the beginning of the study, but appears towards the end of the study)
-   **"Decreasing"** \<- The presence of the species decreases throughout the duration of the study (it is present during the beginning of the study, but then disappears towards the end of the study)
-   **"Recurrent"** \<- Species is detected in consecutive years and is followed by the absence of the species for a period of time, in a distinct pattern. Species must at least two instances of presences followed by absences.
-   **"Random"** \<- Species is detected sporadically for the duration of the study at watershed

Abundance Classification categories include:

-   **Invariant** \<- Species abundance did not vary across the time series. Cannot be analyzed.
-   **"No_change-absent"** \<- Species was detected at other surrounding watersheds, but it was absent at specified watershed for the duration of the time series (\# years present = 0).
-   **"Rare"** \<- Species was present at specified watershed for between 1-10% of years in the duration of the time series. Abundance will not be classified.
-   **"No_change-increasing"** \<- Species at watershed is above average abundance in \>=90% of years in which data was collected. #CHECKME: is this even possible?
-   **"No_change-decreasing"** \<- Species at watershed is below average abundance in \>=90% of years in which data was collected. #CHECKME: is this even possible?
-   **"Increasing"** \<- The second half of the study contains most of the above average records of the species (it is mostly below average during the beginning of the study)
-   **"Decreasing"** \<- The second half of the study contains most of the below average records of the species (it is mostly above average during the beginning of the study)
-   **"Recurrent"** \<- Species abundance fluctuations appear to be clumped into consecutive years, with distinct periods of above average and below average abundances.
-   **"Random"** \<- Species abundance fluctuates in an unpredictable pattern of above vs below average for the duration of the study.

#----------------------------------------------------------------------------------

# Determining Species Incidence Classification

## Small-mammals (1992 - 2013) {.tabset}

### Small-mammals: Presence/ Absence & Incidence Classification

For this section, I am reorganizing the cleaned small-mammal data and reformatting it into a presence/absence matrix. This matrix will then run through the getTrends3.0 function to get the incidence classifications for each species at each watershed.

Conduct the **incidence classification** method
```{r smammals incidence classification}

# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
smammals <- E0_smammals %>% 
  group_by(Recyear, Watershed_name, Species) %>% 
  reframe(Count = sum(Count))

#Combine the species & watershed variables to create a unique identifier
smammals <- smammals %>% unite("Species.watershed", Species:Watershed_name, sep = ".", remove = FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
smammals_years_wide <- smammals %>% 
  pivot_wider(names_from = "Recyear", values_from = "Count")

#Create a separate species.watershed dataframe to add back to the dataset after standardization
smammals_env <- smammals_years_wide[, 1:3]

#Standardize by presence/absence
decostand_smammals_pa <- decostand(smammals_years_wide[, -(1:3)], "pa", na.rm = TRUE )

#Transform the presence/absence data into a matrix so I can feed it to the function
smammals_pa_matrix <- as.matrix(decostand_smammals_pa)

#Apply the getTrends3.0 function to each row of the presence/absence matrix
smammals_trends <- apply(smammals_pa_matrix, 1, getTrends3.0) 

# Create a table with the species.watershed and the getTrends info. 
#   puts everything back into a dataframe format
#   1-10 stands for the number of rows in the output tibble from getTrends3.0
smammals_classification <- tibble((smammals_env),                            do.call(rbind,lapply(smammals_trends,(function(v){v[c(1:10)]}))))

#Add back in the presence/absence per year data
smammals_classification <- smammals_classification %>% add_column(decostand_smammals_pa)

# Write results to file
write_csv(smammals_classification, "Datasets/E1_output_data/E1_smammals/E1_smammals_classified.csv")
```

Conduct the **abundance classification** method
```{r smammals abundance classification}
#use the wide format count data as input; convert to matrix format and remove labels
smammals_years_matrix <- as.matrix(smammals_years_wide[, -(1:3)])

# input matrix into label_abundance method
# returns a new matrix where 0=below average, 1=above or equal to average
#   final column = proportion of years the species was absence (proportion of zeros)
abund_matrix <- smammals_years_matrix %>%
  as.data.frame() %>%
  mutate(row_id = row_number()) %>%
  split(.$row_id) %>%
  map(~ label_abundance(as.numeric(.x[ , -ncol(.x)]))) %>%
  do.call(rbind, .)

#input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
smammals_abun_trends <- apply(abund_matrix, 1, getAbundTrends) 

#Create table with the getAbunTrends results annotated with species & watershed labels
#   labels from smammals_env, created earlier
smammals_abun_classification <- tibble((smammals_env), do.call(rbind,lapply(smammals_abun_trends,(function(v){v[c(1:9)]}))))

# Write to file
write_csv(smammals_abun_classification, "Datasets/E1_output_data/E1_smammals/E1_smammals_classified_abund.csv")
```

## Grasshoppers (2002 - 2020) {.tabset}

### Grasshoppers: Presence/ Absence & Incidence Classification

For this section, I am reorganizing the grasshopper data and reformatting it into a presence/absence matrix. This matrix will then be ran through the getTrends3.0 function to get the incidence classifications for each species at each watershed.

Conduct the **incidence classification** method
```{r grasshoppers incidence classification}
# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
grasshoppers <- E0_grasshoppers %>% 
  group_by(Recyear, Watershed_name, Species) %>% 
  reframe(Count = sum(Total)) %>% 
  unite("Species.watershed", Species:Watershed_name, sep = ".", remove = FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
grasshoppers_years_wide <- grasshoppers %>% 
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:22, ~replace_na(.,0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
grasshoppers_env <- grasshoppers_years_wide[, 1:3]

#Standardize by presence/absence
decostand_grasshoppers_pa <- decostand(grasshoppers_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

#Transform the presence/absence data into a matrix so I can feed it to the function
grasshoppers_pa_matrix <- as.matrix(decostand_grasshoppers_pa)

#Apply the getTrends3.0 function to each row of the presence/absence matrix
grasshoppers_trends <- apply(grasshoppers_pa_matrix, 1, getTrends3.0)

# Create a table with the species.watershed and the getTrends info. 
#   puts everything back into a dataframe format
#   1-10 stands for the number of rows in the output tibble from getTrends3.0
grasshoppers_classification <- tibble((grasshoppers_env), do.call(rbind,lapply(grasshoppers_trends,(function(v){v[c(1:10)]}))))

#Add back in the presence/absence per year data
grasshoppers_classification <- grasshoppers_classification %>% add_column(decostand_grasshoppers_pa)

# Write to file
write_csv(grasshoppers_classification, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_classified.csv")
```

Conduct the **abundance classification** method
```{r grasshoppers abundance classification}
#use the wide format count data as input; convert to matrix format and remove labels
grasshoppers_years_matrix <- as.matrix(grasshoppers_years_wide[, -(1:3)])

# input matrix into label_abundance method
# returns a new matrix where 0=below average, 1=above or equal to average
#   final column = proportion of years the species was absence (proportion of zeros)
abund_matrix <- grasshoppers_years_matrix %>%
  as.data.frame() %>%
  mutate(row_id = row_number()) %>%
  split(.$row_id) %>%
  map(~ label_abundance(as.numeric(.x[ , -ncol(.x)]))) %>%
  do.call(rbind, .)

#input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
grasshoppers_abun_trends <- apply(abund_matrix, 1, getAbundTrends) 

#Create table with the getAbunTrends results annotated with species & watershed labels
#   labels from grasshoppers_env, created earlier
grasshoppers_abun_classification <- tibble((grasshoppers_env), do.call(rbind,lapply(grasshoppers_abun_trends,(function(v){v[c(1:9)]}))))

# Write to file
write_csv(grasshoppers_abun_classification, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_classified_abund.csv")
```

## Birds (1992 - 2009) {.tabset}

### Birds: Presence/ Absence & Incidence Classification

For this section, I am reorganizing the bird data and reformatting it into a presence/absence matrix. This matrix will then be ran through the getTrends3.0 function to get the incidence classifications for each species at each watershed.

Conduct the **incidence classification** method
```{r birds incidence classification}
# Sum the data based on watershed
# Here, we won't differentiate the data based on the trap lines or season
birds <- E0_birds %>% 
  group_by(Recyear, Watershed_name, AOU_code) %>% 
  reframe(Count = sum(Count)) %>%
  unite("Species.watershed", AOU_code:Watershed_name, sep=".", remove=FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
birds_years_wide <- birds %>% 
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:21, ~replace_na(.,0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
birds_env <- birds_years_wide[, 1:3]

#Standardize by presence/absence
decostand_birds_pa <- decostand(birds_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

#Transform the presence/absence data into a matrix so I can feed it to the function
birds_pa_matrix <- as.matrix(decostand_birds_pa)

#Apply the getTrends3.0 function to each row of the presence/absence matrix
birds_trends <- apply(birds_pa_matrix, 1, getTrends3.0)

# Create a table with the species.watershed and the getTrends info. 
#   puts everything back into a dataframe format
#   1-10 stands for the number of rows in the output tibble from getTrends3.0
birds_classification <- tibble((birds_env), do.call(rbind,lapply(birds_trends,(function(v){v[c(1:10)]}))))

#Add back in the presence/absence per year data
birds_classification <- birds_classification %>% add_column(decostand_birds_pa)

# Write to file
write_csv(birds_classification, "Datasets/E1_output_data/E1_birds/E1_birds_classified.csv")
```

Conduct the **abundance classification** method
```{r birds abundance classification}
#use the wide format count data as input; convert to matrix format and remove labels
birds_years_matrix <- as.matrix(birds_years_wide[, -(1:3)])

# input matrix into label_abundance method
# returns a new matrix where 0=below average, 1=above or equal to average
#   final column = proportion of years the species was absence (proportion of zeros)
abund_matrix <- birds_years_matrix %>%
  as.data.frame() %>%
  mutate(row_id = row_number()) %>%
  split(.$row_id) %>%
  map(~ label_abundance(as.numeric(.x[ , -ncol(.x)]))) %>%
  do.call(rbind, .)

#input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
birds_abun_trends <- apply(abund_matrix, 1, getAbundTrends) 

#Create table with the getAbunTrends results annotated with species & watershed labels
#   labels from birds_env, created earlier
birds_abun_classification <- tibble((birds_env), do.call(rbind,lapply(birds_abun_trends,(function(v){v[c(1:9)]}))))

# Write to file
write_csv(birds_abun_classification,
          "Datasets/E1_output_data/E1_birds/E1_birds_classified_abund.csv")
```


## Plants (1992 - 2022) {.tabset}

### Plants: Presence/ Absence & Incidence Classification

For this section, I am reorganizing the plant data and reformatting it into a presence/absence matrix. This matrix will then be ran through the getTrends3.0 function to get the classifications for each species at each watershed.

**Note**: Plant data does not collect abundance as counts, but uses cover instead. 
It is recorded  on a scale of 1-7 where: 
  - 1 = <1% cover
  - 2 = 1-5% cover
  - 3 = 5-25% cover
  - 4 = 25-50% cover
  - 5 = 50-75% cover
  - 6 = 75-95% cover
  - 7 = 95-100% cover

Conduct the **incidence classification** method
*Note: Here, the column named 'Count' represents median Cover class (range 1-7) for plants*
```{r plants incidence classification}

# Average the data based on watershed (not using SUM because it's cover, not count, data)
# Here, we won't differentiate the data based on the trap lines or season
# NOTE: Count=median(cover) here, but it can be converted into pres/abs later
plants <- E0_plants %>% 
  group_by(Recyear, Watershed_name, Genus_spp) %>% 
  reframe(Count = median(Cover)) %>%
  unite("Species.watershed", Genus_spp:Watershed_name, sep=".", remove=FALSE)

#Pivot the table so that each year is a column and each row is the 'species.watershed' variable
# years where the species was not detected should appear as NA, and are converted to zero
plants_years_wide <- plants %>% 
  pivot_wider(names_from = "Recyear", values_from = "Count") %>%
  mutate(across(4:34, ~replace_na(.,0)))

#Create a separate species.watershed dataframe to add back to the dataset after standardization
plants_env <- plants_years_wide[, 1:3]

#Standardize by presence/absence
decostand_plants_pa <- decostand(plants_years_wide[, -c(1:3)], "pa", na.rm = FALSE)

#Transform the presence/absence data into a matrix so I can feed it to the function
plants_pa_matrix <- as.matrix(decostand_plants_pa)

#Apply the getTrends3.0 function to each row of the presence/absence matrix
plants_trends <- apply(plants_pa_matrix, 1, getTrends3.0)

# Create a table with the species.watershed and the getTrends info. 
#   puts everything back into a dataframe format
#   1-10 stands for the number of rows in the output tibble from getTrends3.0
plants_classification <- tibble((plants_env), do.call(rbind,lapply(plants_trends,(function(v){v[c(1:10)]}))))

#Add back in the presence/absence per year data
plants_classification <- plants_classification %>% add_column(decostand_plants_pa)

# Write to file
write_csv(plants_classification, "Datasets/E1_output_data/E1_plants/E1_plants_classified.csv")
```

Conduct the **abundance classification** method
*Note: Here, the column named 'Count' represents median Cover class (range 1-7) for plants*
```{r plants abundance classification}
#use the wide format count data as input; convert to matrix format and remove labels
plants_years_matrix <- as.matrix(plants_years_wide[, -(1:3)])

# input matrix into label_abundance method
# returns a new matrix where 0=below average, 1=above or equal to average
#   final column = proportion of years the species was absence (proportion of zeros)
abund_matrix <- plants_years_matrix %>%
  as.data.frame() %>%
  mutate(row_id = row_number()) %>%
  split(.$row_id) %>%
  map(~ label_abundance(as.numeric(.x[ , -ncol(.x)]))) %>%
  do.call(rbind, .)

#input abund_matrix into getAbundTrends function, apply to each row of the converted matrix
plants_abun_trends <- apply(abund_matrix, 1, getAbundTrends) 

#Create table with the getAbunTrends results annotated with species & watershed labels
#   labels from birds_env, created earlier
plants_abun_classification <- tibble((plants_env), do.call(rbind,lapply(plants_abun_trends,(function(v){v[c(1:9)]}))))

# Write to file
write_csv(plants_abun_classification,
          "Datasets/E1_output_data/E1_plants/E1_plants_classified_abund.csv")

```


#----------------------------------------------------------------------------------------------
## Summary statistics on classification (all taxa) {.tabset}
This section makes some basic summaries of the incidence classification results.

#### Small mammals

**Compare the incidence classification results across species and watersheds**

How often did species "switch" their classifications versus stay consistent across sites?
```{r smammals_summstat}

# Make a table that counts how many times each species was classified  particular way
smammals_abun_class_summary <- smammals_abun_classification %>% 
  group_by(Species, classification) %>% 
  summarize(total_count=n(),.groups = 'drop')

# Make a table that counts how many classifications each species had (across 6 watersheds) 
smammals_class.switch1 <- smammals_classification %>% 
  group_by(Species) %>%
  summarize(count = n_distinct(classification))

# plot as a bar graph
smammals_class.switch1 %>% ggplot(aes(x = count)) + 
  #geom_histogram(binwidth = 0.5) + 
  geom_bar() +
  theme_bw() +
  geom_text(x=3.5, y=7, label= 
              "Switch classification at least once = 12/14 (86%)") +
  labs(x = "Number of classifications per species", 
       title = "Small-mammals: incidence classification switches")
```

Were species classified differently across the 6 watersheds?
```{r}
# Make a table that shows how each species was categorized, and how many times (across 6 watersheds) 
smammals_class.switch2 <- smammals_classification %>%
  group_by(Species, classification) %>%
  summarize(count = n_distinct(Watershed_name))
#view(smammals_class.switch2)

smammals_class.switch2_wide <- smammals_class.switch2 %>% pivot_wider(names_from = classification, values_from = count)
#view(smammals_class.switch2_wide)
```

How frequent were the each of the incidence classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
smammals_class.switch3 <- smammals_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Species))
#view(smammals_class.switch3)

smammals_class.switch3_wide <- smammals_class.switch3 %>% pivot_wider(names_from = classification, values_from = count)
view(smammals_class.switch3_wide)
```

How frequent were the each of the abundance classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
smammals_class.switch4 <- smammals_abun_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Species))
#view(smammals_class.switch3)

smammals_class.switch4_wide <- smammals_class.switch4 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(smammals_class.switch4_wide)
```

#### Grasshoppers

**Compare the incidence classification results across species and watersheds**

How often did species "switch" their classifications versus stay consistent across sites?
```{r grasshoppers_summstat}

# Make a table that counts how many times each species was classified  particular way
grasshoppers_abun_class_summary <- grasshoppers_abun_classification %>% 
  group_by(Species, classification) %>% 
  summarize(total_count=n(),.groups = 'drop')

# Make a table that counts how many classifications each species had (across 6 watersheds) 
grasshoppers_class.switch1 <- grasshoppers_classification %>% 
  group_by(Species) %>%
  summarize(count = n_distinct(classification))

# plot as a bar graph
grasshoppers_class.switch1 %>% ggplot(aes(x = count)) + 
  #geom_histogram(binwidth = 0.5) + 
  geom_bar() +
  theme_bw() +
  geom_text(x=3.5, y=7, label= 
              "Switch classification at least once = 30/44 (68%)") +
  labs(x = "Number of classifications per species", 
       title = "Grasshoppers: incidence classification switches")
```

Were species classified differently across the 6 watersheds?
```{r}
# Make a table that shows how each species was categorized, and how many times (across 6 watersheds) 
grasshoppers_class.switch2 <- grasshoppers_classification %>%
  group_by(Species, classification) %>%
  summarize(count = n_distinct(Watershed_name))
#view(grasshoppers_class.switch2)

grasshoppers_class.switch2_wide <- grasshoppers_class.switch2 %>% 
  pivot_wider(names_from = classification, values_from = count)
#view(grasshoppers_class.switch2_wide)
```

How frequent were the each of the incidence classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
grasshoppers_class.switch3 <- grasshoppers_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Species))
#view(grasshoppers_class.switch3)

grasshoppers_class.switch3_wide <- grasshoppers_class.switch3 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(grasshoppers_class.switch3_wide)
```

How frequent were the each of the abundance classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
grasshoppers_class.switch4 <- grasshoppers_abun_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Species))

grasshoppers_class.switch4_wide <- grasshoppers_class.switch4 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(grasshoppers_class.switch4_wide)
```


#### Birds

**Compare the incidence classification results across species and watersheds**

How often did species "switch" their classifications versus stay consistent across sites?
```{r birds_summstat}

# Make a table that counts how many times each species was classified a particular way
birds_abun_class_summary <- birds_abun_classification %>% 
  group_by(AOU_code, classification) %>% 
  summarize(total_count=n(),.groups = 'drop')

# Make a table that counts how many classifications each species had (across 6 watersheds) 
birds_class.switch1 <- birds_classification %>% 
  group_by(AOU_code) %>%
  summarize(count = n_distinct(classification))

# plot as a bar graph
birds_class.switch1 %>% ggplot(aes(x = count)) + 
  #geom_histogram(binwidth = 0.5) + 
  geom_bar() +
  theme_bw() +
  geom_text(x=3.5, y=7, label= 
              "Switch classification at least once = 44/81 (54%)") +
  labs(x = "Number of classifications per species", 
       title = "Birds: incidence classification switches")
```

Were species classified differently across the 6 watersheds?
```{r}
# Make a table that shows how each species was categorized, and how many times (across 6 watersheds) 
birds_class.switch2 <- birds_classification %>%
  group_by(AOU_code, classification) %>%
  summarize(count = n_distinct(Watershed_name))
#view(grasshoppers_class.switch2)

birds_class.switch2_wide <- birds_class.switch2 %>% 
  pivot_wider(names_from = classification, values_from = count)
#view(grasshoppers_class.switch2_wide)
```

How frequent were the each of the incidence classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
birds_class.switch3 <- birds_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(AOU_code))
#view(grasshoppers_class.switch3)

birds_class.switch3_wide <- birds_class.switch3 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(birds_class.switch3_wide)
```

How frequent were the each of the abundance classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
birds_class.switch4 <- birds_abun_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(AOU_code))

birds_class.switch4_wide <- birds_class.switch4 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(birds_class.switch4_wide)
```


#### Plants

**Compare the incidence classification results across species and watersheds**

How often did species "switch" their classifications versus stay consistent across sites?
```{r plants_summstat}

# Make a table that counts how many times each species was classified a particular way
plants_abun_class_summary <- plants_abun_classification %>% 
  group_by(Genus_spp, classification) %>% 
  summarize(total_count=n(),.groups = 'drop')

# Make a table that counts how many classifications each species had (across 6 watersheds) 
plants_class.switch1 <- plants_classification %>% 
  group_by(Genus_spp) %>%
  summarize(count = n_distinct(classification))

# plot as a bar graph
plants_class.switch1 %>% ggplot(aes(x = count)) + 
  #geom_histogram(binwidth = 0.5) + 
  geom_bar() +
  theme_bw() +
  geom_text(x=3.5, y=7, label= 
              "Switch classification at least once = 207/325 (64%)") +
  labs(x = "Number of classifications per species", 
       title = "Plants: incidence classification switches")
```

Were species classified differently across the 6 watersheds?
```{r}
# Make a table that shows how each species was categorized, and how many times (across 6 watersheds) 
plants_class.switch2 <- plants_classification %>%
  group_by(Genus_spp, classification) %>%
  summarize(count = n_distinct(Watershed_name))

plants_class.switch2_wide <- plants_class.switch2 %>% 
  pivot_wider(names_from = classification, values_from = count)
```

How frequent were the each of the classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
plants_class.switch3 <- plants_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Genus_spp))
#view(plants_class.switch3)

plants_class.switch3_wide <- plants_class.switch3 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(plants_class.switch3_wide)
```

How frequent were the each of the abundance classifications across the watersheds?
```{r}
# Make a table that shows how many species were in each classification, by watershed 
plants_class.switch4 <- plants_abun_classification %>%
  group_by(Watershed_name, classification) %>%
  summarize(count = n_distinct(Genus_spp))

plants_class.switch4_wide <- plants_class.switch4 %>% 
  pivot_wider(names_from = classification, values_from = count)
view(plants_class.switch4_wide)
```

#----------------------------------------------------------------------------------------------
## Dissimilarity from baseline year and between watersheds (all taxa) {.tabset}

This section is for conducting dissimilarity tests between watersheds for the small-mammal dataset.

These dissimilarity tests will include Jaccard's and Bray comparison for: - the first year to each subsequent year of the time series, within watersheds - across watersheds, within each year separately

Is dissimilarity increasing through time? Is the rate of composition change different according to treatment?

#### Small mammal Composition

Save results for **Small mammal composition dissimilarity**
```{r}
# prepare the count data into wide format with merged column for watershed_year;
# Adds NAs for missing species since they aren't all present on every watershed
smammals_count_allyears <- smammals %>% 
  group_by(Watershed_name, Species, Recyear) %>% 
  reframe(Total = (Count)) %>% 
  pivot_wider(names_from = "Species", values_from = "Total", 
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_")

# small mammal dissimilarity estimates
sm_jacc <- compute_dissimilarity(smammals_count_allyears, method="jaccard", baseline_year = 1992)
sm_bray <- compute_dissimilarity(smammals_count_allyears, method="bray", baseline_year = 1992)

# merge and write results
sm_baseline <- merge(sm_jacc[1], sm_bray[1])
sm_watershed <- merge(sm_jacc[2], sm_bray[2])

# for comparison to baseline year, add linear model trend results
# Fit trend model across time for each watershed
sm_jacc_diss_lm <- sm_baseline %>%
  rename(Watershed_name = baseline.Watershed2) %>%
  rename(value = baseline.jaccard_diss) %>% #make this generic so it works with the function
  group_by(Watershed_name) %>%
  mutate(n = row_number()) %>%
  arrange(Watershed_name, n) %>%
  nest() %>%
  mutate(
    model = map(data, fit_model),
    slope = map_dbl(model, get_slope),
    p_value = map_dbl(model, get_p_value),
    rsquared = map_dbl(model, get_rsq)
  )  %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)


sm_bray_diss_lm <- sm_baseline %>%
  rename(Watershed_name = baseline.Watershed2) %>%
  rename(value = baseline.bray_diss) %>% #make this generic so it works with the function
  group_by(Watershed_name) %>%
  mutate(n = row_number()) %>%
  arrange(Watershed_name, n) %>%
  nest() %>%
  mutate(
    model = map(data, fit_model),
    slope = map_dbl(model, get_slope),
    p_value = map_dbl(model, get_p_value),
    rsquared = map_dbl(model, get_rsq)
  ) %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)

##Save to file
write.csv(sm_baseline, "Datasets/E1_output_data/E1_smammals/E1_smammals_dissimilarity_year1toN.csv")

write.csv(sm_jacc_diss_lm, "Datasets/E1_output_data/E1_smammals/E1_smammals_jacc_diss_lm.csv")
write.csv(sm_bray_diss_lm, "Datasets/E1_output_data/E1_smammals/E1_smammals_bray_diss_lm.csv")

write.csv(sm_watershed, "Datasets/E1_output_data/E1_smammals/E1_smammals_dissimilarity_watersheds.csv")
```

#### Grasshopper Composition

Save results for **Grasshopper composition dissimilarity**
```{r}
# prepare the count data into wide format with merged column for watershed_year
grasshoppers_count_allyears <- grasshoppers %>% 
  group_by(Watershed_name, Species, Recyear) %>% 
  reframe(Total = (Count)) %>% 
  pivot_wider(names_from = "Species", values_from = "Total", 
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep = "_") %>%
  replace(is.na(.), "0")

# find the first year of the study
year1 <- min(grasshoppers$Recyear)

# grasshopper dissimilarity estimates
gh_jacc <- compute_dissimilarity(grasshoppers_count_allyears, method="jaccard", baseline_year = year1)
gh_bray <- compute_dissimilarity(grasshoppers_count_allyears, method="bray", baseline_year = year1)

# merge and write results
gh_baseline <- merge(gh_jacc[1], gh_bray[1])
gh_watershed <- merge(gh_jacc[2], gh_bray[2])

##Save to file
write.csv(gh_baseline, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_dissimilarity_year1toN.csv")

write.csv(gh_watershed, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_dissimilarity_watersheds.csv")
```

#### Bird Composition

Save results for **Bird composition dissimilarity**
```{r}
# prepare the count data into wide format with merged column for watershed_year
birds_count_allyears <- birds %>% 
  group_by(Watershed_name, AOU_code, Recyear) %>% 
  reframe(Total = (Count)) %>% 
  pivot_wider(names_from = "AOU_code", values_from = "Total",
              values_fn = function(x) paste(sum(x))) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep="_") %>%
  replace(is.na(.),"0")

# find the first year of the study
year1 <- min(birds$Recyear)

# bird dissimilarity estimates
bd_jacc <- compute_dissimilarity(birds_count_allyears, method="jaccard", baseline_year = year1)
bd_bray <- compute_dissimilarity(birds_count_allyears, method="bray", baseline_year = year1)

# merge and write results
bd_baseline <- merge(bd_jacc[1], bd_bray[1])
bd_watershed <- merge(bd_jacc[2], bd_bray[2])

##Save to file
write.csv(bd_baseline, "Datasets/E1_output_data/E1_birds/E1_birds_dissimilarity_year1toN.csv")

write.csv(bd_watershed, "Datasets/E1_output_data/E1_birds/E1_birds_dissimilarity_watersheds.csv")
```

#### Plant Composition

**Note:** The plant dataset uses an ordinal variable for cover class instead of abundance, or count. This is fine to covert to presence absence for Jaccard dissimilarity but would be flawed to use in Bray-Curtis directly, in place of abundance. There are two potential options, and we can calculate both: 

1. Convert cover class to midpoints (this still would represent percent cover, not abundance)
2. Use an alternative method that works with ordinal variables, like Gower's distance

I have a preference for using Gower's distance here because we can treat the variables as ordinal, without assuming equal intervals between the classes (they aren't; see label descriptions above). The first option of converting to midpoints implicitly suggests continuous or equal interval data, which it isn't really. Gower 0 == the same; Gower 1 == completely different.

Save results for **Plant composition dissimilarity**
```{r}
# prepare the cover data into wide format with merged column for watershed_year
plants_count_allyears <- plants %>% 
  #add a line to ensure Count is an ordered factor (plant cover class)
#  mutate(Count = factor(Count, levels = 1:7, ordered = TRUE)) %>%
  group_by(Watershed_name, Genus_spp, Recyear) %>% 
  reframe(Total = (Count)) %>% 
  pivot_wider(names_from = "Genus_spp", values_from = "Total",
              values_fn = function(x) paste(x)) %>%
  unite("Watershed_year", Watershed_name:Recyear, sep="_") %>%
  replace(is.na(.),"0")

# find the first year of the study
year1 <- min(plants$Recyear)

# bird dissimilarity estimates
pl_jacc <- compute_dissimilarity(plants_count_allyears, method="jaccard", baseline_year = year1)
pl_gower <- compute_dissimilarity(plants_count_allyears, method="gower", baseline_year = year1)

# merge and write results
pl_baseline <- merge(pl_jacc[1], pl_gower[1])
pl_watershed <- merge(pl_jacc[2], pl_gower[2])

# for comparison to baseline year, add linear model trend results
# Fit trend model across time for each watershed
pl_jacc_diss_lm <- pl_baseline %>%
  rename(Watershed_name = baseline.Watershed2) %>%
  rename(value = baseline.jaccard_diss) %>% #make this generic so it works with the function
  group_by(Watershed_name) %>%
  mutate(n = row_number()) %>%
  arrange(Watershed_name, n) %>%
  nest() %>%
  mutate(
    model = map(data, fit_model),
    slope = map_dbl(model, get_slope),
    p_value = map_dbl(model, get_p_value),
    rsquared = map_dbl(model, get_rsq)
  )  %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)

# for comparison to baseline year, add linear model trend results
# Fit trend model across time for each watershed
pl_gower_diss_lm <- pl_baseline %>%
  rename(Watershed_name = baseline.Watershed2) %>%
  rename(value = baseline.gower_diss) %>% #make this generic so it works with the function
  group_by(Watershed_name) %>%
  mutate(n = row_number()) %>%
  arrange(Watershed_name, n) %>%
  nest() %>%
  mutate(
    model = map(data, fit_model),
    slope = map_dbl(model, get_slope),
    p_value = map_dbl(model, get_p_value),
    rsquared = map_dbl(model, get_rsq)
  )  %>%
    dplyr::select(Watershed_name, slope, p_value, rsquared)


##Save to file
write.csv(pl_baseline, "Datasets/E1_output_data/E1_plants/E1_plants_dissimilarity_year1toN.csv")
write.csv(pl_jacc_diss_lm, "Datasets/E1_output_data/E1_plants/E1_plants_jacc_diss_lm.csv")
write.csv(pl_gower_diss_lm, "Datasets/E1_output_data/E1_plants/E1_plants_gower_diss_lm.csv")

write.csv(pl_watershed, "Datasets/E1_output_data/E1_plants/E1_plants_dissimilarity_watersheds.csv")
```



#----------------------------------------------------------------------------------------

## Species Richness change over time (all taxa) {.tabset}
This section applies a function to calculate species richness per watershed and per year. 
It then determines if richness is changing directionally over time using a simple linear model.

**input:** the presence absence matrix and the watershed-species information. Both of these come from previous code steps (using decostand and creating the env dataframe)
**returns:** a list with both dataframes to result -- contains summary richness data and the linear model results

#### Small mammals

```{r smammals richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_smammals_pa, smammals_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_smammals/E1_smammals_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_smammals/E1_smammals_richness_lm.csv")
```

#### Grasshoppers

```{r grasshopper richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_grasshoppers_pa, grasshoppers_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_richness_lm.csv")
```

#### Birds

```{r bird richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_birds_pa, birds_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_birds/E1_birds_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_birds/E1_birds_richness_lm.csv")
```

#### Plants

```{r plants richness}

# Apply function to calculate species richness per watershed and per year
result <- richness_trend(decostand_plants_pa, plants_env)

#Save richness data 
write.csv(result$richness, "Datasets/E1_output_data/E1_plants/E1_plants_richness.csv")

#Save richness model results
write.csv(result$model, "Datasets/E1_output_data/E1_plants/E1_plants_richness_lm.csv")
```



## Abundance change over time (all taxa) {.tabset}
This section determines if abundance is changing directionally over time using a simple linear model, using the count data for each taxonomic. While abundance fluctuates in a non-linar fashion for most species, this is meant to suggest which species have more apparent trends in increasing or decreasing abundance (i.e. it will not detect random or cyclical changes or particular extremes.

**For plant taxa** We applied an ordinal logistic regression instead. This model works with ordinal abundance instead, where we can't assume the intervals between classes are all equal. The result models the probability of being in a higher or lower cover class as a function of year. "As year increases, is the plant more likely to be in a higher cover class?"

Each section writes a csv file summarizing the abundance information and model results.

#### Small-mammals; count data

```{r small mammals abundance}
#make a version of the wide dataset without the extra columns, for the linear model analysis
sm_spabund <- smammals_years_wide %>% 
  dplyr::select(-Watershed_name, -Species)

#Use 'purr' to run the model. The gather function get the values we need, and the group_by function makes sure the values are grouped by 'Watershed_name'. 
#   'n' is the row number, which translates to the year (1992 = 1, 1993 = 2... 2022 = 31). 
#   The arrange function orders the data by 'Watershed_name' and row number. 
#   The nest function nests all the data (year/row number, and richness) into a single column, while each row is a watershed. 
#   The rest of the code runs the linear models for each row, and creates new columns with the outputs.
sm_spabun_lm <- sm_spabund %>%
  gather('year', 'value', -Species.watershed) %>%
  group_by(Species.watershed) %>%
  mutate(n = row_number()) %>%
  arrange(Species.watershed, n) %>%
  nest() %>%
  mutate(model = map(data, fit_model)) %>%
  mutate(slope = map_dbl(model, get_slope)) %>%
  mutate(p_value = map_dbl(model, get_p_value)) %>%
  mutate(rsquared = map_dbl(model, get_rsq)) %>%
  # remove detailed model results before writing to file
  dplyr::select(Species.watershed, slope, p_value, rsquared) %>%
  separate(Species.watershed, c("Genus", "Spp", "Watershed_name")) %>%
  unite("Species", Genus:Spp, sep = "_") 

#Save abundance data 
write.csv(smammals_years_wide, "Datasets/E1_output_data/E1_smammals/E1_smammals_abundance.csv")

#Save abundance model results
write.csv(sm_spabun_lm, "Datasets/E1_output_data/E1_smammals/E1_smammals_abundance_lm.csv")
```

#### Grasshoppers; count data

```{r grasshoppers abundance}
#make a version of the wide dataset without the extra columns, for the linear model analysis
gh_spabund <- grasshoppers_years_wide %>% 
  dplyr::select(-Watershed_name, -Species)

# Run linear models for each row (species-watershed) and save the results
gh_spabun_lm <- gh_spabund %>%
  gather('year', 'value', -Species.watershed) %>%
  group_by(Species.watershed) %>%
  mutate(n = row_number()) %>%
  arrange(Species.watershed, n) %>%
  nest() %>%
  mutate(model = map(data, fit_model)) %>%
  mutate(slope = map_dbl(model, get_slope)) %>%
  mutate(p_value = map_dbl(model, get_p_value)) %>%
  mutate(rsquared = map_dbl(model, get_rsq)) %>%
  # remove detailed model results before writing to file
  dplyr::select(Species.watershed, slope, p_value, rsquared) %>%
  separate(Species.watershed, c("Genus", "Spp", "Watershed_name")) %>%
  unite("Species", Genus:Spp, sep = "_") 

#Save abundance data 
write.csv(grasshoppers_years_wide, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_abundance.csv")

#Save abundance model results
write.csv(gh_spabun_lm, "Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_abundance_lm.csv")
```


#### Birds; count data

```{r birds abundance}
#make a version of the wide dataset without the extra columns, for the linear model analysis
bd_spabund <- birds_years_wide %>% 
  dplyr::select(-Watershed_name, -AOU_code)

# Run linear models for each row (Species-watershed) and save the results
bd_spabun_lm <- bd_spabund %>%
  gather('year', 'value', -Species.watershed) %>%
  group_by(Species.watershed) %>%
  mutate(n = row_number()) %>%
  arrange(Species.watershed, n) %>%
  nest() %>%
  mutate(model = map(data, fit_model)) %>%
  mutate(slope = map_dbl(model, get_slope)) %>%
  mutate(p_value = map_dbl(model, get_p_value)) %>%
  mutate(rsquared = map_dbl(model, get_rsq)) %>%
  # remove detailed model results before writing to file
  dplyr::select(Species.watershed, slope, p_value, rsquared) %>%
  separate(Species.watershed, c("AOU_code", "Watershed_name")) # %>%
  unite("Species", Genus:Spp, sep = "_") 

#Save abundance data 
write.csv(birds_years_wide, "Datasets/E1_output_data/E1_birds/E1_birds_abundance.csv")

#Save abundance model results
write.csv(bd_spabun_lm, "Datasets/E1_output_data/E1_birds/E1_birds_abundance_lm.csv")
```


#### Plants; percent cover data
Here, we use the `polr` function for ordinal logistic regression instead of a linear model. Some key notes: 
* psuedo pvalue calculated
* McFadden's psuedo R2 calculated from null model comparison
* model convergence failure is somewhat common; handled through NULL and NA values
* low variation or no variation in ordinal response values is sometimes a problem
* if n perfectly predicts class it is a problem
* overfitting could be an issue

```{r plants abundance}
#make a version of the wide dataset without the extra columns, for the linear model analysis
pl_spabund <- plants_years_wide %>% 
  dplyr::select(-Watershed_name, -Genus_spp)

# Run ordinal logistic regression models for each row (Species-watershed) and save the results
pl_spabun_logit <- pl_spabund %>%
  pivot_longer(cols = -Species.watershed, 
               names_to = "year",
               values_to = 'value') %>%
  group_by(Species.watershed) %>%
  mutate(n = row_number(),
         value = factor(value, levels = 0:7, ordered = TRUE)) %>% #ensure ordered factor
  arrange(Species.watershed, n) %>%
  nest() %>%
  mutate(model = map(data, fit_ordinal_model), 
         model_null = map(data, fit_ordinal_null),
         slope = map_dbl(model, get_polr_slope),
         psuedo_pval = map_dbl(model, get_polr_p),
         pseudo_rsq = map2_dbl(model, model_null, get_polr_rsq)) %>%
  # remove detailed model results before writing to file
  dplyr::select(Species.watershed, slope, psuedo_pval, pseudo_rsq) %>%
  separate(Species.watershed, into = c("Species", "Watershed_name"), sep="\\.")
  
#Save abundance data 
write.csv(plants_years_wide, "Datasets/E1_output_data/E1_plants/E1_plants_abundance.csv")

#Save abundance model results
write.csv(pl_spabun_logit, "Datasets/E1_output_data/E1_plants/E1_plants_abundance_logit.csv")
```

#-----------------------------------------------------------------------------------------
# STATISTICAL COMPARISONS FOR INCIDENCE and ABUNDANCE CLASSIFICATION

prepare INCIDENCE data for statistical comparison
```{r}
#prepare the INCIDENCE data for analysis
smammal_class_counts <- smammals_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(smammals_classification$Species)))) %>%
  mutate(taxa = "small mammals")

grasshoppers_class_counts <- grasshoppers_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(grasshoppers_classification$Species)))) %>%
  mutate(taxa = "grasshoppers")

birds_class_counts <- birds_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(birds_classification$AOU_code)))) %>%
  mutate(taxa = "birds")

plants_class_counts <- plants_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(plants_classification$Genus_spp)))) %>%
  mutate(taxa = "plants")

#rbind all the class summary datasets together
all_class_counts <- rbind(smammal_class_counts, grasshoppers_class_counts, birds_class_counts, plants_class_counts)

# make classification, watershed, and taxa ordered factors
# remove non-repeated watersheds
# remove absent classification (won't be compared)
all_class_counts <- all_class_counts %>%
  filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B")) %>%
  filter(!classification == "No_change-absent") %>% 
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         taxa = factor(taxa, levels = c("plants", "grasshoppers", "small mammals", "birds"),
                       ordered = TRUE),
         classification = factor(classification, 
                                 levels = c("No_change-present", "Rare",
                                            "Increasing", "Decreasing",
                                            "Recurrent", "Random"), 
                                 ordered = TRUE))
```

prepare ABUNDANCE data for statistical comparison
```{r}
#prepare the INCIDENCE data for analysis
smammal_class_abun <- smammals_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(smammals_abun_classification$Species)))) %>%
  mutate(taxa = "small mammals")

grasshoppers_class_abun <- grasshoppers_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(Proportion = (total_count/length(unique(grasshoppers_abun_classification$Species)))) %>%
  mutate(taxa = "grasshoppers")

birds_class_abun <- birds_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(birds_abun_classification$AOU_code)))) %>%
  mutate(taxa = "birds")

plants_class_abun <- plants_abun_classification %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(Proportion = (total_count/length(unique(plants_abun_classification$Genus_spp)))) %>%
  mutate(taxa = "plants")

#rbind all the class summary datasets together
all_class_abun <- rbind(smammal_class_abun, grasshoppers_class_abun, birds_class_abun, plants_class_abun)

# make classification, watershed, and taxa ordered factors
all_class_abun <- all_class_abun %>%
  filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         taxa = factor(taxa, 
                       levels = c("plants", "grasshoppers", "small mammals", "birds"),
                       ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("Invariant", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE))
```

### 1. Are classifications significantly different across the watershed treatments?

Apply Chi-square INCIDENCE comparison across watersheds for each taxa
```{r}
# run chi-square comparison across watersheds for each taxa 
chisq_results <- all_class_counts %>%
  dplyr::select(-Proportion) %>%
  group_by(taxa) %>%
  nest() %>%
  mutate(chisq = map(data, run_chisq_from_nested)) %>%
  unnest(chisq)

# add Bonferroni adjustments to p-value
chisq_results <- chisq_results %>%
  mutate(p_bonf = ifelse(!is.na(p_value), p.adjust(p_value, method = "bonferroni"), NA))

write_csv(chisq_results, "Datasets/E1_output_data/chisq_results_by_taxa.csv")
```

Apply Chi-square ABUNDANCE comparison across watersheds for each taxa
```{r}
# run chi-square comparison across watersheds for each taxa 
chisq_results_abun <- all_class_abun %>%
  dplyr::select(-Proportion) %>%
  group_by(taxa) %>%
  nest() %>%
  mutate(chisq = map(data, run_chisq_from_nested)) %>%
  unnest(chisq)

# add Bonferroni adjustments to p-value
chisq_results_abun <- chisq_results_abun %>%
  mutate(p_bonf = ifelse(!is.na(p_value), p.adjust(p_value, method = "bonferroni"), NA))

write_csv(chisq_results_abun, "Datasets/E1_output_data/chisq_abun_results_by_taxa.csv")
```

**ANSWER: For most taxa NO. For plant taxa YES.** 

Plants do show some changes in classification frequency across treatments. 

The hypothesis regarding if recurrent classification is more represented in 4-year burns is not supported. Does the recurrent category represent what we hope it does, or is it too simplistic? Or are recurrent cycles not really apparent at this site/treatment?

The hypothesis regarding if press disturbance (grazing) results in more increasing/decreasing species is somewhat supported. Sites where native grazers are absent always are underrepresented by increasing taxa and over-represented by decreasing taxa compared to sites where native grazers have been added (press treatment). The direct effect of bison is on the "directed change" category, supporting a press disturbance impact showing up in these classifications.


### 1a. For taxa that showed significant shifts in classification frequency, what is driving that change?

For incidence classification: only plants demonstrated a significant difference (small cramers V)
For abundance classification: only plants demonstrated a significant difference (small cramers V)

Look at the INCIDENCE residuals
```{r}
# prepare data for plant residuals
pl_resid <- chisq_results %>%
  filter(taxa == "plants") %>%
  pull(residuals) %>%
  .[[1]] %>%
  pivot_longer(-classification, names_to = "Watershed_name", values_to = "Residual") %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE))

# heatmap of the residuals
ggplot(pl_resid, aes(x = Watershed_name, y = classification, fill = Residual)) +
  #geom_point(aes(color = Residual, size = abs(Residual))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red4", mid = "white", high = "navyblue", 
                       midpoint = 0, limits = c(-max(abs(pl_resid$Residual)), 
                                                max(abs(pl_resid$Residual))),
                       name = "Std Residual") +
  labs(title = "Plant standardized residuals", 
       x = "Watershed", 
       y = "Incidence Classification") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plant_incidence_chisq_residuals.png", dpi=300, width=5, height=4, bg="white")
```

The same plot for chisq residuals on comparing incidence classification, but using points instead of filled cells
```{r}
# plot of the residuals
ggplot(pl_resid, aes(x = Watershed_name, y = classification)) +
  geom_point(aes(size = abs(Residual), color = Residual)) +
  scale_color_gradient2(
    low = "red4", mid = "white", high = "navyblue", 
    midpoint = 0,
    limits = c(-max(abs(pl_resid$Residual)), max(abs(pl_resid$Residual))),
    name = "Std Residual"
  ) +
  scale_size_continuous(range = c(1, 10), name = "Abs(Residual)") +
  labs(
    title = "Plant standardized residuals",
    x = "Watershed", 
    y = "Incidence Classification"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plant_incidence_chisq_residuals_points.png", dpi=300, width=5, height=4, bg="white")
```

We can also try a version that only shows the  more extreme residuals, >|2|
```{r}
ggplot(pl_resid, aes(x = Watershed_name, y = classification)) +
  
  # Main points: size by abs(Residual), color by Residual
  geom_point(aes(size = abs(Residual), color = Residual)) +
  
  # Border points: only where abs(Residual) > 2
  geom_point(
    data = pl_resid %>% filter(abs(Residual) > 2),
    shape = 21, stroke = 1, fill = NA, color = "black",
    aes(size = abs(Residual))
  ) +

  scale_color_gradient2(
    low = "red4", mid = "white", high = "navyblue",
    midpoint = 0,
    limits = c(-max(abs(pl_resid$Residual)), max(abs(pl_resid$Residual))),
    name = "Std Residual"
  ) +
  scale_size_continuous(range = c(1, 10), name = "Abs(Residual)") +
  
  labs(
    title = "Plant standardized residuals",
    x = "Watershed",
    y = "Incidence Classification"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Figures/plant_incidence_chisq_residuals_points_2plus.png", dpi=300, width=5, height=4, bg="white")

```



Look at the ABUNDANCE residuals
```{r}
# prepare data for plant residuals
pl_resid_abun <- chisq_results_abun %>%
  filter(taxa == "plants") %>%
  pull(residuals) %>%
  .[[1]] %>%
  pivot_longer(-classification, names_to = "Watershed_name", values_to = "Residual") %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
         classification = factor(classification, 
                                 levels=c("Invariant", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE))

# heatmap of the residuals
ggplot(pl_resid_abun, aes(x = Watershed_name, y = classification, fill = Residual)) +
  #geom_point(aes(color = Residual, size = abs(Residual))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red4", mid = "white", high = "navyblue", 
                       midpoint = 0, limits = c(-max(abs(pl_resid_abun$Residual)), 
                                                max(abs(pl_resid_abun$Residual))),
                       name = "Std Residual") +
  labs(title = "Plant standardized residuals", 
       x = "Watershed", 
       y = "Abundance Classification") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figures/plant_abundance_chisq_residuals.png", dpi=300, width=4, height=4)
```

#-----------------------------------------------------------------------------------------
# DATA VISUALIZATIONS

Let's make some descriptive plots of the species classifications. We can make violin plots that will mirror the ones in Gotelli et al. 2022. 

### Incidence Classification
```{r}
ggplot(all_class_counts, aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") + 
  labs(title = "Frequency across watersheds", y = "Incidence Classification") + 
  facet_wrap(~taxa)

ggsave("Figures/all_taxa_class_incid_freq.png", dpi=300, width=6, height=6)
```

Just for plants, since they were the only ones that were significantly different across watersheds
```{r}
pl <- all_class_counts %>%
  filter(taxa == "plants")

#lollipop chart
ggplot(pl, aes(x = Proportion, y = classification)) +
  geom_segment(aes(x = 0, xend = Proportion, 
                   y = classification, yend = classification),
               color = "grey50", linewidth = 0.8) +
  geom_point(aes(color=classification), size = 3) +
  facet_wrap(~Watershed_name, ncol=2) +
  labs(title = "Plant incidence",
       x = "Proportion of species",
       y = "Classification") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

#bar graph
ggplot(pl, aes(x = Proportion, y = classification)) +
  geom_col(aes(fill = classification)) + 
  facet_wrap(~Watershed_name, ncol=2) +
  labs(title = "Plant incidence",
       x = "Proportion of species",
       y = "Classification") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

ggsave("Figures/plant_class_incid_freq.png", dpi=300, width=4, height=6)
```


### Abundance Classification
```{r}
ggplot(all_class_abun, aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  theme_bw() +
  theme(legend.position = "none") + 
  labs(title = "Frequency across watersheds", y = "Abundance Classification") + 
  facet_wrap(~taxa)

ggsave("Figures/all_taxa_class_abun_freq.png", dpi=300, width=6, height=6)
```

Just for plants, since they were the only ones that were significantly different across watersheds
```{r}
pl_abun <- all_class_abun %>%
  filter(taxa == "plants")

#bar graph
ggplot(pl_abun, aes(x = Proportion, y = classification)) +
  geom_col(aes(fill = classification)) + 
  facet_wrap(~Watershed_name, ncol=2) +
  labs(title = "Plant abundance class",
       x = "Proportion of species",
       y = "Classification") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

ggsave("Figures/plant_class_abun_freq.png", dpi=300, width=4, height=6)
```

### Composition
Since incidence change is necessarily impacted by species entering and exiting a community in different years, let's look at the underlying species composition patterns at the site. 

Let's focus on plants, since that's where the incidence classes were most different
```{r}
## PLANTS Jaccard

# process relevant data frames
pl_baseline <- pl_baseline %>% rename(Watershed_name=baseline.Watershed2)

pl_diss <- left_join(pl_baseline, pl_jacc_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(pl_diss, aes(x = baseline.Year2, y = baseline.jaccard_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=pl_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Jaccard dissimilarity to Year 1") +
  labs(title = "plant Jaccard trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_jacc_years.png", dpi=300, width=4, height=6)
```

```{r}
## PLANTS Gower

# process relevant data frames
pl_baseline <- pl_baseline %>% rename(Watershed_name=baseline.Watershed2)

pl_diss <- left_join(pl_baseline, pl_gower_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(pl_diss, aes(x = baseline.Year2, y = baseline.gower_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=pl_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Gower dissimilarity to Year 1") +
  labs(title = "plant Gower trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_gower_years.png", dpi=300, width=4, height=6)
```

Let's dig a bit deeper into the directed change category. 
**Which plant species are being classified as directed change?**
```{r}
pl_class <- plants_classification %>%
  #ensure just the watersheds for this study are kept
    filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B"))

pl_directed <- pl_class %>%
  filter(classification %in% c("Decreasing", "Increasing")) %>%
  dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification)

directed_spp <- unique(pl_directed$Genus_spp)

# How many species were increasing in at least one watershed?
pl_directed %>%
  filter(classification == "Increasing") %>%
  summarize(n_distinct(Genus_spp))

#How many species were decreasing in at least one watershed? 
pl_directed %>%
  filter(classification == "Decreasing") %>%
  summarize(n_distinct(Genus_spp))
```

**Which species are increasing at all watersheds in which they occur?**
```{r}
# Species consistently increasing
inc_species <- pl_class %>%
  group_by(Genus_spp) %>%
  filter(all(classification == "Increasing")) %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    Watershed_count = n_distinct(Watershed_name),
    .groups = "drop"
  )

#View(inc_species)
all_inc = unique(inc_species$Genus_spp)

nrow(inc_species)
unique(inc_species$Genus_spp)
```
There are 22 species that are increasing at all watersheds in which they occur. 15/22 of these only occur in grazed sites, indicating a colonization effect for these sites, potentially in response to the grazing treatment. 

**Which species are decreasing at all watersheds in which they occur?**
```{r}
# Species consistently decreasing
dec_species <- pl_class %>%
  group_by(Genus_spp) %>%
  filter(all(classification == "Decreasing")) %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    Watershed_count = n_distinct(Watershed_name),
    .groups = "drop"
  )

all_dec = unique(dec_species$Genus_spp)

dim(dec_species)
```
Answer: There are NO species that were decreasing at all watersheds in which they occur. 

**Which are increasing or decreasing only in the grazed watersheds, but not in the ungrazed?**
-- I also need to make a new subset of the data that includes the other classifications for comparison
INCREASING
```{r}
# Species increasing across grazed/ungrazed treatments
inc_species <- pl_class %>%
    mutate(treatment = case_when(
    Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed",
    TRUE ~ "ungrazed")) %>%
  group_by(Genus_spp) %>%
  filter(classification == "Increasing") %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    trmts = paste(sort(unique(treatment)), collapse = ", "),
    trmt_count = n_distinct(treatment),
    .groups = "drop"
  ) %>%
  arrange(trmt_count)
  
View(inc_species)
```
DECREASING
```{r}
# Species decreasing across grazed/ungrazed treatments
dec_species <- pl_class %>%
    mutate(treatment = case_when(
    Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed",
    TRUE ~ "ungrazed")) %>%
  group_by(Genus_spp) %>%
  filter(classification == "Decreasing") %>%
  summarise(
    Watersheds = paste(sort(unique(Watershed_name)), collapse = ", "),
    trmts = paste(sort(unique(treatment)), collapse = ", "),
    trmt_count = n_distinct(treatment),
    .groups = "drop"
  ) %>%
  arrange(trmt_count)
  
View(dec_species)
```

**What else did they switch from in the watersheds where they were not in the directed change classes?**
```{r}
# Species increasing across grazed/ungrazed treatments
dir_species <- pl_class %>%
    mutate(treatment = case_when(
      Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed", TRUE ~ "ungrazed")) %>%
  filter(Genus_spp %in% directed_spp) %>%
    filter(!Genus_spp %in% all_inc) %>%
    dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    arrange(Genus_spp, Watershed_name)

View(dir_species)

```

Can we annotate directed species with known traits like growth form, photosynthesis, and annual/perennial?
```{r}
# species info
pl_info <- pl_info %>%
  mutate(Genus_spp = paste(gen, spec, sep = "_")) %>%
  distinct()

#add details back in for plants that are directed change
dir_species <- left_join(dir_species, pl_info, by="Genus_spp")

dir_spp_info <- dir_species %>%
  dplyr::select(Genus_spp, genus, species, family, `growth form`, `life form`, origin) %>%
  distinct()

#check out some summaries (note that if it doesn't add up to 146, is NA)
table(dir_spp_info$origin)
table(dir_spp_info$`growth form`)
table(dir_spp_info$`life form`)
```
#FIXME: I am not sure what to do with the above results... How to summarize better without getting into the weeds?


**how does classification relate to percentage of years present?**
```{r}
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  ggplot(aes(x = classification, y = percyears)) +
  geom_jitter(width = 0.2, height = 0, color = "steelblue", alpha = 0.6) +
  geom_violin(alpha=0.7, aes(fill=classification)) +
  labs(title = "Percent of Years by Classification",
       x = "Classification",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() + 
  guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_class_percyears.png", dpi=300, width=4, height=4)
```

**how does proportion of years present relate to watershed?**
```{r}
#boxplot
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  mutate(treatment = case_when(
      Watershed_name %in% c("N1B", "N4D", "N20B") ~ "grazed", TRUE ~ "ungrazed")) %>%
  ggplot(aes(x = Watershed_name, y = percyears, fill=Watershed_name)) +
  geom_boxplot(alpha=0.7, notch=TRUE) +
  scale_fill_manual(values = c("darkolivegreen", "sienna4", "darkolivegreen4", 
                               "sienna3", "darkolivegreen3", "sienna2")) +
  labs(title = "Percent of Years by Watershed",
       x = "Watershed",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() +
  guides(fill="none") +
  theme(legend.position="none")

ggsave("Figures/plants_watershed_percyears_bxp.png", dpi=300, width=4, height=4)
```

```{r}
#violin + jitter plot
pl_class %>%
   dplyr::select(Watershed_name, Genus_spp, numyears, numpresent, percyears, classification) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE)) %>%
  ggplot(aes(x = Watershed_name, y = percyears, group=Watershed_name)) +
  geom_violin(alpha=0.7) +
  geom_jitter(aes(col=classification), width = 0.2, height = 0, alpha = 0.8) +
  labs(title = "Percent of Years by Watershed",
       x = "Watershed",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() 

ggsave("Figures/plants_watershed_percyears.png", dpi=300, width=4, height=4)
```

**how does life form relate to percentage of years present?**
#FIXME: This plot only shows those that are increasing or decreasing on at least some plots; not all species
```{r}
dir_species %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("No_change-present", "Rare","Increasing", "Decreasing",
                                          "Recurrent", "Random"), 
                                 ordered = TRUE),
           growth_form = factor(`growth form`, 
                                levels = c("a", "b", "p"),
                                ordered = TRUE)) %>%
  ggplot(aes(x = growth_form, y = percyears)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6) +
  geom_boxplot(alpha=0.5, aes(fill=growth_form), notch=TRUE) +
  labs(title = "Percent of Years by Growth Form",
       x = "Growth Form",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  theme_minimal() + 
  guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_dir-change-ann-per_percyears.png", dpi=300, width=4, height=4)
```

**for species that were directed change in a particular watershed, how are they divided out across watershed and across growth form?**
```{r}
dir_species %>%
  filter(classification %in% c("Increasing", "Decreasing")) %>%
    mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE),
           classification = factor(classification, 
                                 levels=c("Decreasing", "Increasing"), 
                                 ordered = TRUE),
           growth_form = factor(`growth form`, 
                                levels = c("a", "b", "p"),
                                ordered = TRUE)) %>%
  ggplot(aes(x = growth_form, y = percyears)) +
  geom_boxplot(alpha=0.5) +
  #geom_violin() +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6, aes(col=classification)) +
  labs(title = "Directed change populations",
       x = "Growth Form",
       y = "Percent of Years Present") +
  geom_hline(yintercept = c(0.33, 0.67), linetype="dashed") +
  facet_wrap(~Watershed_name, ncol=2) +
  theme_bw()# + 
  #guides(fill="none") + 
  theme(legend.position="none")

ggsave("Figures/plants_dir-change-ann-per__watershed_percyears.png", dpi=300, width=4, height=6)
```

#----------------------------------------------------------

```{r}
## SMAMMALS Jaccard

# process relevant data frames
sm_baseline <- sm_baseline %>% rename(Watershed_name=baseline.Watershed2)

sm_diss <- left_join(sm_baseline, sm_jacc_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(sm_diss, aes(x = baseline.Year2, y = baseline.jaccard_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=sm_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +    
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Jaccard dissimilarity to Year 1") +
  labs(title = "small mammal Jaccard trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/smammals_jacc_years.png", dpi=300, width=4, height=6)
```

```{r}
## SMAMMALS Bray

# process relevant data frames
sm_diss <- left_join(sm_baseline, sm_bray_diss_lm, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))


# make panel plot of results
ggplot(sm_diss, aes(x = baseline.Year2, y = baseline.bray_diss)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=sm_diss %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  xlab("Year") +
  ylab("Bray-Curtis dissimilarity to Year 1") +
  labs(title = "small mammal Bray trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/smammals_bray_years.png", dpi=300, width=4, height=6)
```
#TODO: START HERE, make composition plots like this for all taxa; will need to jump up to composition section and add lm models and writecsv to all other taxa too. 


#TODO: START HERE AGAIN

# TODO: START IN THE PLOTTING CODE AND ADD THIS TYPE OF PLOT FOR RICHNESS AND ABUNDANCE
#     LINES ONLY SHOW WHEN THE PVALUE IS SIGNIFICANT, MAKES IT EASIER TO SCAN; 
#     filter out extra watersheds
#     add label for trend direction

###Plant richness
```{r}
# process relevant data frames
rich_p <- left_join(result$richness, result$model, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE)) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))

  
# make panel plot of results 
(check that plants were the most recent richness result run)
ggplot(rich_p, aes(x = Recyear, y = Richness)) + 
  geom_point(size=0.75) + 
  facet_wrap(.~Watershed_name, ncol=2) + 
  geom_smooth(data=rich_p %>% filter(p_value <= 0.05), 
              aes(color = trend_dir),
              method = "lm") +
  scale_color_manual(values = c("positive" = "cadetblue4", "negative" = "indianred2")) +
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10),
        legend.position = "none") +
  labs(title = "Plant richness trends",
       color = NULL)  +
  ggpubr::stat_cor()

ggsave("Figures/plants_richness_years.png", dpi=300, width=4, height=6)
```

###Plant abundance
```{r}
# identify the plant species that had significant cover class trends, using pseudo pvalue
#   and add trend direction (positive or negative)
pl_spabun_logit_sig <- pl_spabun_logit %>% 
  filter(psuedo_pval <= 0.05) %>%
  mutate(trend_dir = ifelse(slope >= 0, "positive", "negative"))

#show results divided by positive and negative significant trends
table(pl_spabun_logit_sig$Watershed_name, pl_spabun_logit_sig$trend_dir)

# process relevant data frames
#TODO: NEED TO ADD WATERSHED BACK INTO ABUNDANCE DF
pl_abun <- left_join(pl_spabund, pl_spabun_logit_sig, by="Watershed_name") %>%
  filter(!Watershed_name %in% c("4A", "N4A")) %>%
  mutate(Watershed_name = factor(Watershed_name, 
                                 levels = c("1D", "N1B", "4B", "N4D", "20B", "N20B"), 
                                 ordered = TRUE))
  
#TODO: NOT SURE BEST WAY TO SHOW THE RESULTS. MAYBE EACH WATERSHED WITH THE SPECIES LINES SHOWN IF SIGNIF INCREASING OR DECREASING.THEN IT'S STILL 6 PANELS, BUT MIGHT BE MESSY?Thee are 206 unique species :O
```



#leaving this here as an inspo note for E2 file data viz
```{r}
## Creating a plot to show how dissimilarity in species presence compared to the baseline year for watershed 4F in the small mammals dataset
plants_jaccdiss_4A_plot <- 
  plants_jaccdiss_4A_df[-(1), ] %>% 
  ggplot(aes(x = Year2, y = Jacc_diss)) + 
  geom_point(alpha = 0.5, size = 4, color = "#4A148C") +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme_bw() +
  geom_label(label="4A", 
    x=30,
    y=1.00,
    label.padding = unit(0.55, "lines"), # Rectangle size around label
    label.size = 0.6,
    color = "white",
    fill="#4A148C") +
    ylim(0, 1) + 
  labs(y = "Jaccard's Dissimilarity", x = "Year from Baseline")

#Putting all the plots together (used to be other plots, just here for quick reminder for now...)
plot_grid(
  plants_jaccdiss_1D_plot, 
  #plants_jaccdiss_4A_plot, #not included for other taxa
  plants_jaccdiss_4B_plot,
  plants_jaccdiss_20B_plot, 
  plants_jaccdiss_N1B_plot,
  #plants_jaccdiss_N4A_plot, #not included for other taxa
  plants_jaccdiss_N4D_plot,
  plants_jaccdiss_N20B_plot,
nrow = 2,
ncol = 3
)

ggsave("Figures/plant_jaccard-diss_3xfire.png", dpi=300, width=10, height=6)
```


#FIXME: delete when ready, old code from 2024 poster plots
Chi square tests for significance
```{r}
#---------plants
# set up each taxa as a contingency table
p_chsq <- plants_class_counts %>%
  filter(classification != "No_change-absent") %>%
  filter(Watershed_name %in% c("1D", "N1B", "4B", "N4D", "20B", "N20B")) %>%
  dplyr::select(Watershed_name, classification, total_count) %>%
  pivot_wider(names_from = Watershed_name, values_from = total_count) %>% 
  replace(is.na(.), 0)

#make it be a matrix for the ChiSquare test
p_chsq2 <- as.matrix(p_chsq[,2:7])
  
plant_chisq <- chisq.test(p_chsq2) #Categories ARE SIG DIFF for plants

#calculate and round Pearson residuals from Observed & Expected values
round(plant_chisq$residuals, 3)
  # Positive residuals specify an attraction (positive association) between the corresponding row and column variables.
  # Negative residuals imply a repulsion (negative association)

#show as a corrplot
rownames(plant_chisq$residuals) <- c("Rare", "No change", "recurrent", "random", "increasing", "decreasing")
corrplot(plant_chisq$residuals, is.cor = FALSE)

# Contibution in percentage (%)
contrib <- 100*plant_chisq$residuals^2/plant_chisq$statistic
round(contrib, 3)

# Visualize the contribution
corrplot(contrib, is.cor = FALSE)
```
