---
title: "E2_AllTaxa_Plots"
author: "Maya Parker-Smith"
date: "2023-12-06"
output:
  html_document:
    smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---


```{r setup, include=FALSE}

library(tidyverse)
library(corrplot)
```

Read in all relevant datasets for plotting
```{r}
# small mammal data and results
E1_smammals_classified <- read_csv("Datasets/E1_output_data/E1_smammals/E1_smammals_classified.csv")
E1_smammals_classified_abund <-
  read_csv("Datasets/E1_output_data/E1_smammals/E1_smammals_classified_abund.csv")
E1_smammals_dissimilarity_year1toN <- 
  ("read_csvDatasets/E1_output_data/E1_smammals/E1_smammals_dissimilarity_year1toN.csv")
E1_smammals_richness <- 
  read_csv("Datasets/E1_output_data/E1_smammals/E1_smammals_richness.csv")
E1_smammals_abundance <- read.csv("Datasets/E1_output_data/E1_smammals/E1_smammals_abundance.csv")
E1_smammals_abundance_lm <- read.csv("Datasets/E1_output_data/E1_smammals/E1_smammals_abundance_lm.csv")

# grasshopper data and results
E1_grasshoppers_classified <-
  read_csv("Datasets/E1_output_data/E1_grasshoppers/E1_grasshoppers_classified.csv")

# bird data and results
E1_birds_classified <- read_csv("Datasets/E1_output_data/E1_birds/E1_birds_classified.csv")

# plant data and results
E1_plants_classified <- read_csv("Datasets/E1_output_data/E1_plants/E1_plants_classified.csv")


```


## Small Mammals

### Set-up
```{r smammals set up}
smammals_classified_long <- E1_smammals_classified %>% 
  pivot_longer(cols = "1992":"2013", names_to = "Recyear", values_to = "Presence")

smammals_graphing <- smammals_classified_long %>% 
  dplyr::select(Recyear, Species, Common_name, 
                classification, Presence, Watershed_name)


#Making the zeroes in the presence column into NAs for plotting (that way they don't show up on the graph)
smammals_graphing$Presence[smammals_graphing$Presence == 0] <- NA

#Making sure the Recyear is coded as numeric 
smammals_graphing <- smammals_graphing %>% mutate(Recyear = as.numeric(Recyear))


E1_smammals_classified$Watershed_name <- factor(E1_smammals_classified$Watershed_name,
                                        levels = c("1D", "4B", "20B", "N1B", "N4D", "N20B"))

E1_smammals_classified$classification <- factor(E1_smammals_classified$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

smammals_graphing$Watershed_name <- factor(smammals_graphing$Watershed_name,
                                        levels = c("1D", "4B", "20B", "N1B", "N4D", "N20B"))

smammals_graphing$classification <- factor(smammals_graphing$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

#unique(smammals_classified$Species)
```


```{r}
E1_smammals_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/14)) %>%
  
  ggplot(aes(x = classification, y = Proportion, fill = classification)) + 
  geom_col() + 
  scale_fill_manual(values = c("lightgray", "darkgray","#000000", "#CC79A7", "#0072B2", "#009E73")) +
  facet_wrap(.~Watershed_name) +
  theme_bw() +
  theme(text = element_text(size=14), 
        axis.text.x = element_text(angle=90, hjust=1)) +
  labs(title= "Proportion of Small Mammals Species in Each Incidence Classification for Each Watershed", 
       x = "Classification", 
       y = "Proportion of Species", 
       fill = "Classification")

  


smammals_violin <- E1_smammals_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/14))

smammals_violin <- smammals_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


smammals_violin[is.na(smammals_violin)] <- 0

smammals_violin <- smammals_violin %>% 
  pivot_longer(cols = "No_change-absent":"Increasing", 
               names_to = "classification", values_to = "Proportion")
  
smammals_violin$classification <- factor(smammals_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing"))

smammals_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("lightgray","darkgray","#000000", "#CC79A7", "#0072B2", "#009E73")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Small mammals: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/smammals_overall_classes.png", dpi=300, width=6, height=4)
```

```{r}
smammals_graphing %>% 
  ggplot(aes(x = Recyear, y = Presence, fill = classification)) + 
  geom_tile(width = .6, height = .6) + 
  facet_grid(Common_name~Watershed_name, switch = "y") +
  theme(text = element_text(size=9),
        axis.text.x = element_text(angle=45, hjust=1, size =9),
        axis.text.y = element_text(size=12),
        strip.text.y.left = element_text(angle = 0, size = 9),
        strip.text.x.top = element_text(size = 9)) +
  geom_vline(xintercept = 2002.5, linetype= 2) +
  coord_fixed(ratio = 5) +
  labs(x = "Year", y = "Presence") +
  scale_fill_manual(values = c("white","darkgray" ,"black", "#CC79A7", "#0072B2","#009E73")) +
  labs(title = "Small Mammal Classifcation by Species",
       fill = "Classification")

```

**Small Mammal Dissimilarity plots**
Visualize Jaccard's Dissimilarity compared to baseline year (year 1)
```{r}
smammals_jacc_plot <- 
  ggplot(data=E1_smammals_dissimilarity_year1toN, aes(x=Year2, y=Jacc_diss)) + 
  geom_point() +
  geom_smooth(method="lm") +
  ggpubr::stat_cor() +
  ylim(0,1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x="Year compared to baseline", y="Jaccard's dissimilarity") + 
  facet_wrap(~Watershed1)
  
  smammals_jacc_plot
  
  ggsave("Figures/smammals_jacc_years.png", dpi=300, width=6, height=4)
```

Visualize the Bray-Curtis Dissimilarity compared to baseline year (year 1)
```{r}
smammals_bray_plot <- 
  ggplot(data=E1_smammals_dissimilarity_year1toN, aes(x=Year2, y=Bray_diss)) + 
  geom_point() +
  geom_smooth(method="lm") +
  ggpubr::stat_cor() +
  ylim(0,1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x="Year compared to baseline", y="Bray-Curtis dissimilarity") + 
  facet_wrap(~Watershed1)
  
  smammals_bray_plot
  
  ggsave("Figures/smammals_bray_years.png", dpi=300, width=6, height=4)
```

**Small Mammal Species Richness plots**
Visualize the richness trends over time across the watersheds.
All are statistically insignificant.
```{r}
ggplot(data=E1_smammals_richness, aes(x = Recyear, y = Richness)) + 
  geom_point() + 
  facet_wrap(.~Watershed_name) + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw() +
  theme(text = element_text(size=12),
        axis.text.x = element_text(angle=45, hjust=1, size =12)) +
  labs(title = "Small Mammal Species Richness trend") +
  xlab("Year") +
  ggpubr::stat_cor()

ggsave("Figures/small_mammals_richness.png", dpi=300, width=6, height=4)
```

Visualize the species abundance trends
```{r}
# Take the table of species counts, then 
#   pivot the data into a long format,
#   where 'Recyear' column is a numeric column, and 
#   the 'Watershed_name', 'Species', and 'Common_name' columns into a factor
smammals_spabund_long <- E1_smammals_abundance %>%
  pivot_longer(cols = "1992":"2013",
               names_to = "Recyear",
               values_to = "Count") %>%
  mutate(Recyear = as.numeric(Recyear),
         Watershed_name = factor(Watershed_name),
         Species = factor(Species),
         Common_name = factor(Common_name))%>%
  dplyr::select(-Species.watershed)

# show abundance trends by species, comparing across watersheds
ggplot(data=smammals_spabund_long, aes(x = Recyear, y = Count, group = Watershed_name)) + 
  geom_point(alpha=0.50, aes(col=Watershed_name)) + 
  geom_smooth(se = FALSE, aes(col=Watershed_name)) + 
  theme_bw() +
  theme(text = element_text(size=12),
        axis.text.x = element_text(angle=45, hjust=1, size =12)) +
  labs(title = "Small Mammal Species Abundance trends") +
  xlab("Year") +
  scale_color_viridis(discrete = TRUE, option = "D") +  # options: "A" to "E"
  #ggpubr::stat_cor() + 
  facet_wrap(.~Species, scales='free')

ggsave("Figures/small_mammals_abundance.png", dpi=300, width=10, height=10)
```


## Grasshoppers

### Set-up
```{r}
grasshoppers_classified_long <- E1_grasshoppers_classified %>% 
  pivot_longer(cols = "2002":"2020", names_to = "Recyear", values_to = "Presence")

grasshoppers_graphing <- grasshoppers_classified_long %>% 
  dplyr::select(Recyear, Species,classification, Presence, Watershed_name)


#Making the zeroes in the presence column into NAs for plotting (that way they don't show up on the graph)
grasshoppers_graphing$Presence[grasshoppers_graphing$Presence == 0] <- NA

#Making sure the Recyear is coded as numeric 
grasshoppers_graphing <- grasshoppers_graphing %>% mutate(Recyear = as.numeric(Recyear))


E1_grasshoppers_classified$Watershed_name <- factor(E1_grasshoppers_classified$Watershed_name,
                                        levels = c("1D", "N1B", "4B", 
                                           "4F", "N4A", "N4D", 
                                           "20B", "N20B"))

E1_grasshoppers_classified$classification <- factor(E1_grasshoppers_classified$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

grasshoppers_graphing$Watershed_name <- factor(grasshoppers_graphing$Watershed_name,
                                        levels = c("1D", "N1B", "4B", 
                                           "4F", "N4A", "N4D", 
                                           "20B", "N20B"))

grasshoppers_graphing$classification <- factor(grasshoppers_graphing$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))


```


```{r}
grasshoppers_violin <- E1_grasshoppers_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/44))

grasshoppers_violin <- grasshoppers_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


grasshoppers_violin[is.na(grasshoppers_violin)] <- 0

grasshoppers_violin <- grasshoppers_violin %>% 
  pivot_longer(cols = "No_change-absent":"Decreasing", 
               names_to = "classification", values_to = "Proportion") 
  
grasshoppers_violin$classification <- factor(grasshoppers_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Decreasing"))

grasshoppers_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("lightgray", "darkgray", "#000000", "#CC79A7", "#0072B2", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Grasshoppers: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/grasshoppers_overall_classes.png", dpi=300, width=6, height=4)

```



## Birds
```{r}
birds_classified_long <- E1_birds_classified %>% 
  pivot_longer(cols = "1992":"2009", names_to = "Recyear", values_to = "Presence")

birds_graphing <- birds_classified_long %>% 
  dplyr::select(Recyear, Species, classification, Presence, Watershed_name)


#Making the zeroes in the presence column into NAs for plotting (that way they don't show up on the graph)
birds_graphing$Presence[birds_graphing$Presence == 0] <- NA

#Making sure the Recyear is coded as numeric 
birds_graphing <- birds_graphing %>% mutate(Recyear = as.numeric(Recyear))


E1_birds_classified$Watershed_name <- factor(E1_birds_classified$Watershed_name,
                                        levels = c("1D", "N1B", "4A", 
                                           "4B", "N4B", "N4D", 
                                           "20B", "N20B"))

E1_birds_classified$classification <- factor(E1_birds_classified$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

birds_graphing$Watershed_name <- factor(birds_graphing$Watershed_name,
                                        levels = c("1D", "N1B", "4A", 
                                           "4B", "N4B", "N4D", 
                                           "20B", "N20B"))

birds_graphing$classification <- factor(birds_graphing$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))


```


```{r}
birds_violin <- E1_birds_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/81))

birds_violin <- birds_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


birds_violin[is.na(birds_violin)] <- 0

birds_violin <- birds_violin %>% 
  pivot_longer(cols = "No_change-absent":"Increasing", 
               names_to = "classification", values_to = "Proportion") 
  
birds_violin$classification <- factor(birds_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing", "Decreasing"))

birds_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("lightgray","darkgray", "#000000", "#CC79A7", "#0072B2", "#009E73", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Birds: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/birds_overall_classes.png", dpi=300, width=6, height=4)


```

## Plants

```{r}
unique(E1_plants_classified$Watershed_name)

#FIXME: clean up the increasing/decreasing stuff, or put that as a separate analysis/visual in a different markdown chunk.
plants_increasing <- E1_plants_classified %>% subset(classification == "Increasing")
length(unique(plants_increasing$Species))

plants_increasing_firstfourth <- plants_increasing[1:28, ] 
plants_increasing_secondfourth <- plants_increasing[29:56,] 
plants_increasing_thirdfourth <- plants_increasing[57:84,] 
plants_increasing_fourthfourth <- plants_increasing[85:112,] 

plants_decreasing <- E1_plants_classified %>% subset(classification == "Decreasing")
length(unique(plants_decreasing$Species))

plants_classified_increasing_firstfourth <- E1_plants_classified %>% subset(Species %in% plants_increasing_firstfourth$Species)



plants_classified_increasing_firstfourth_long <- plants_classified_increasing_firstfourth %>% pivot_longer(cols = "1992":"2022", names_to = "Recyear", values_to = "Presence")

plants_classified_increasing_firstfourth_long$Presence[plants_classified_increasing_firstfourth_long$Presence == 0] <- NA
plants_classified_increasing_firstfourth_long <- plants_classified_increasing_firstfourth_long %>% mutate(Recyear = as.numeric(Recyear))

plants_classified_increasing_firstfourth_long %>% ggplot(aes(x = Recyear, y = Presence, fill = classification)) + geom_tile(width = .6, height = .6) + 
  facet_grid(Species~Watershed_name, switch = "y") +
  theme(text = element_text(size=9),
        axis.text.x = element_text(angle=45, hjust=1, size =9),
        axis.text.y = element_text(size=12),
        strip.text.y.left = element_text(angle = 0, size = 9),
        strip.text.x.top = element_text(size = 9)) +
  labs(x = "Year", y = "Presence") +
  coord_fixed(ratio = 5) +
  geom_vline(xintercept = 2008, linetype= 2)+
  scale_fill_manual(values = c("gray","darkgray", "#000000", "#CC79A7", "#0072B2","#009E73", "#D55E00")) +
  labs(title = "Plant classifcations with increasing incidence (1/4)", 
       fill = "Classification")


plants_classified_long <- E1_plants_classified %>% 
  pivot_longer(cols = "1992":"2022", names_to = "Recyear", values_to = "Presence")



plants_graphing <- plants_classified_long %>% 
  dplyr::select(Recyear, Species,classification, Presence, Watershed_name)


#Making the zeroes in the presence column into NAs for plotting (that way they don't show up on the graph)
plants_graphing$Presence[plants_graphing$Presence == 0] <- NA

#Making sure the Recyear is coded as numeric 
plants_graphing <- plants_graphing %>% mutate(Recyear = as.numeric(Recyear))


E1_plants_classified$Watershed_name <- factor(E1_plants_classified$Watershed_name,
                                        levels = c("1D", "N1B", "4A", 
                                           "4B", "N4A", "N4D", 
                                           "20B", "N20B"))

E1_plants_classified$classification <- factor(E1_plants_classified$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

E1_plants_graphing$Watershed_name <- factor(E1_plants_graphing$Watershed_name,
                                        levels = c("1D", "N1B", "4A", 
                                           "4B", "N4A", "N4D", 
                                           "20B", "N20B"))

plants_graphing$classification <- factor(plants_graphing$classification,
                                     levels = c("No_change-absent", "Rare", "No_change-present",
                                                "Recurrent", "Random", "Increasing", "Decreasing"))

plants_graphing_increasing <- plants_graphing %>% subset(classification == "Increasing")


plants_graphing_increasing <- plants_graphing %>% subset(classification == "Increasing")
length(unique(plants_graphing_increasing$Species))

```


```{r}
plants_violin <- E1_plants_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/325))

plants_violin <- plants_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


plants_violin[is.na(plants_violin)] <- 0

plants_violin <- plants_violin %>% 
  pivot_longer(cols = "No_change-absent":"Decreasing", 
               names_to = "classification", values_to = "Proportion") 
  
plants_violin$classification <- factor(plants_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing", "Decreasing"))

plants_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("lightgray","darkgray", "#000000", "#CC79A7", "#0072B2", "#009E73", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Plants: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/plants_overall_classes.png", dpi=300, width=6, height=4)

```


##Violin plots without the "no change absent category"
In other words, this tracks the prorportions of what is actually there.
```{r}
#SMALL MAMMALS
smammals_violin <- E1_smammals_classified %>% 
  filter(classification != "No_change-absent") %>%
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/14))

smammals_violin <- smammals_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


smammals_violin[is.na(smammals_violin)] <- 0

smammals_violin <- smammals_violin %>% 
  pivot_longer(cols = "Rare":"Increasing", 
               names_to = "classification", values_to = "Proportion")
  
smammals_violin$classification <- factor(smammals_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing"))
smammals_violin$taxa="small mammals"

smammals_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("darkgray","#000000", "#CC79A7", "#0072B2", "#009E73")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Small mammals: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/smammals_overall_classes_NOABS.png", dpi=300, width=6, height=4)

# GRASSHOPPERS
grasshoppers_violin <- E1_grasshoppers_classified %>% 
  filter(classification != "No_change-absent") %>%
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/44))

grasshoppers_violin <- grasshoppers_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


grasshoppers_violin[is.na(grasshoppers_violin)] <- 0

grasshoppers_violin <- grasshoppers_violin %>% 
  pivot_longer(cols = "Rare":"Decreasing", 
               names_to = "classification", values_to = "Proportion") 
  
grasshoppers_violin$classification <- factor(grasshoppers_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Decreasing"))
grasshoppers_violin$taxa = "grasshoppers"

grasshoppers_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("darkgray", "#000000", "#CC79A7", "#0072B2", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Grasshoppers: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/grasshoppers_overall_classes_NOABS.png", dpi=300, width=6, height=4)


# BIRDS
birds_violin <- E1_birds_classified %>% 
  filter(classification != "No_change-absent") %>%
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/81))

birds_violin <- birds_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


birds_violin[is.na(birds_violin)] <- 0

birds_violin <- birds_violin %>% 
  pivot_longer(cols = "Rare":"Increasing", 
               names_to = "classification", values_to = "Proportion") 
  
birds_violin$classification <- factor(birds_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing", "Decreasing"))
birds_violin$taxa = "birds"

birds_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("darkgray", "#000000", "#CC79A7", "#0072B2", "#009E73", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Birds: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/birds_overall_classes_NOABS.png", dpi=300, width=6, height=4)


# PLANTS
plants_violin <- E1_plants_classified %>% 
  filter(classification != "No_change-absent") %>%
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>% 
  mutate(Proportion = (total_count/325))

plants_violin <- plants_violin %>% 
  dplyr::select(Watershed_name, classification, Proportion) %>% 
  pivot_wider(names_from = classification, values_from = Proportion) 


plants_violin[is.na(plants_violin)] <- 0

plants_violin <- plants_violin %>% 
  pivot_longer(cols = "Rare":"Decreasing", 
               names_to = "classification", values_to = "Proportion") 
  
plants_violin$classification <- factor(plants_violin$classification,
         levels = c("No_change-absent", "Rare", "No_change-present",
                    "Recurrent", "Random", "Increasing", "Decreasing"))
plants_violin$taxa = "plants"

plants_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("darkgray", "#000000", "#CC79A7", "#0072B2", "#009E73", "#D55E00")) +
  theme_bw() +
  labs(x = "Proportion of species in watersheds", 
       y = "Classification", 
       title = "Plants: proportion of species in each classification category",
       fill = "Classification")

ggsave("Figures/plants_overall_classes_NOABS.png", dpi=300, width=6, height=4)


alltaxa_violin <- rbind(smammals_violin, grasshoppers_violin, birds_violin, plants_violin)
alltaxa_violin$taxa <- factor(alltaxa_violin$taxa,
         levels = c("plants", "grasshoppers", "birds", "small mammals"))

alltaxa_violin %>% ggplot(aes(x = Proportion, y = classification, fill = classification)) + 
  geom_violin() +
  scale_fill_manual(values = c("darkgray", "#000000", "#CC79A7", "#0072B2", "#009E73", "#D55E00")) +
  facet_wrap(~taxa, ncol=1) +
  theme_bw() +
  labs(x = "Proportion species in watersheds", 
       y = "Classification", 
       title = "Classification by taxa",
       fill = "Classification") +
    theme(legend.position="bottom") +
  theme(legend.title=element_blank())

ggsave("Figures/ALLTAXA_overall_classes_NOABS.png", dpi=300, width=5.5, height=10)

```


Chi square tests for significance
```{r}
#prepare the data
smammal_class_counts <- E1_smammals_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(taxa = "small mammals")

grasshoppers_class_counts <- E1_grasshoppers_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop') %>%
  mutate(taxa = "grasshoppers")

birds_class_counts <- E1_birds_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(taxa = "birds")

plants_class_counts <- E1_plants_classified %>% 
  group_by(Watershed_name, classification) %>% 
  dplyr::summarise(total_count=n(),.groups = 'drop')  %>%
  mutate(taxa = "plants")

#rbind all the count datasets together
alltaxa_counts <- rbind(smammal_class_counts, grasshoppers_class_counts, birds_class_counts, plants_class_counts)
alltaxa_counts$taxa <- factor(alltaxa_counts$taxa,
         levels = c("plants", "grasshoppers", "birds", "small mammals"))

#------------smammals
# set up each taxa as a contingency table, removing the absent ones
smamm_chsq <- smammal_class_counts %>%
  filter(classification != "No_change-absent") %>%
  dplyr::select(Watershed_name, classification, total_count) %>%
  pivot_wider(names_from = Watershed_name, values_from = total_count) %>% 
  replace(is.na(.), 0)

#make it be a matrix for the ChiSquare test
smamm_chsq2 <- as.matrix(smamm_chsq[,2:7])
  
chisq.test(smamm_chsq2) #Categories are NS different for small mammals

#---------hoppers
# set up each taxa as a contingency table
grasshop_chsq <- grasshoppers_class_counts %>%
  filter(classification != "No_change-absent") %>%
  dplyr::select(Watershed_name, classification, total_count) %>%
  pivot_wider(names_from = Watershed_name, values_from = total_count) %>% 
  replace(is.na(.), 0)

#make it be a matrix for the ChiSquare test
grasshop_chsq2 <- as.matrix(grasshop_chsq[,2:9])
  
chisq.test(grasshop_chsq2) #Categories are NS for grasshoppers

#---------birds
# set up each taxa as a contingency table
b_chsq <- birds_class_counts %>%
  filter(classification != "No_change-absent") %>%
  dplyr::select(Watershed_name, classification, total_count) %>%
  pivot_wider(names_from = Watershed_name, values_from = total_count) %>% 
  replace(is.na(.), 0)

#make it be a matrix for the ChiSquare test
b_chsq2 <- as.matrix(b_chsq[,2:9])
  
chisq.test(b_chsq2) #Categories are NS for birds

#---------plants
# set up each taxa as a contingency table
p_chsq <- plants_class_counts %>%
  filter(classification != "No_change-absent") %>%
  dplyr::select(Watershed_name, classification, total_count) %>%
  pivot_wider(names_from = Watershed_name, values_from = total_count) %>% 
  replace(is.na(.), 0)

#make it be a matrix for the ChiSquare test
p_chsq2 <- as.matrix(p_chsq[,2:9])
  
plant_chisq <- chisq.test(p_chsq2) #Categories ARE SIG DIFF for plants

#calculate and round Pearson residuals from Observed & Expected values
round(plant_chisq$residuals, 3)
  # Positive residuals specify an attraction (positive association) between the corresponding row and column variables.
  # Negative residuals imply a repulsion (negative association)

#show as a corrplot
rownames(plant_chisq$residuals) <- c("Rare", "No change", "recurrent", "random", "increasing", "decreasing")
corrplot(plant_chisq$residuals, is.cor = FALSE)

# Contibution in percentage (%)
contrib <- 100*plant_chisq$residuals^2/plant_chisq$statistic
round(contrib, 3)

# Visualize the contribution
corrplot(contrib, is.cor = FALSE)
```

