---
title: "E3_Test_Method"
author: "Sarah Supp"
date: "2026-06-30"
output: html_document
---

This code tests the incidence classification method, by simulating different incidence time series comprised of 0 (not present) and 1 (present). We test different total time series lengths (20, 30, 40), different numbers of incidences (minimum present at least once).  For each different number of incidences, we resample 300 random orderings of the 0s and 1s within the timeseries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tseries)

#Call our custom functions for classification and analysis
source("classify_functions.R")
```

For a given time series length and number of incidences, we generate random orderings to show what the distribution of classifications looks like for different combinations.

Function to generate random arrangements of 0s and 1s
```{r simulation-functions}
# generate a random arrangement of 'npresent' 1's among 'tslen' slots
make_random <- function(tslen, npresent) {
  x <- rep(0, tslen)
  if (npresent > 0) x[sample(1:tslen, npresent)] <- 1
  x
}

# runs one simulation
run_one <- function(tslen, npresent, arrangement) {
  x <- switch(arrangement,
    random       = make_random(tslen, npresent),
    clumped      = make_clumped(tslen, npresent),
    front_loaded = make_front_loaded(tslen, npresent),
    back_loaded  = make_back_loaded(tslen, npresent)
  )
  getTrends3.0(x)
}

```

Now, run the random simulation
```{r run-simulation}
#set seed so results are repeatable
set.seed(41) 

# set values to test 
lengths    <- c(20, 30, 40)
n_runs     <- 300   # random orderings per (length, npresent) combo
arrangement_types <- c("random")

# make a simulation grid that will be used to hold results
sim_grid <- expand_grid(
  tslen       = lengths,
  npresent    = NA,           # filled in below per tslen
  arrangement = arrangement_types,
  run         = 1:n_runs
)

# expand npresent properly per length (0:tslen)
sim_grid <- map_dfr(lengths, function(L) {
  expand_grid(
    tslen       = L,
    npresent    = 1:L,
    arrangement = arrangement_types,
    run         = 1:n_runs
  )
})

# run the test and record the results
results <- sim_grid %>%
  mutate(res = pmap(list(tslen, npresent, arrangement), run_one)) %>%
  unnest(res, names_repair = "minimal")
```

Add colors for plotting. Group colors reflect three main groupings:
   - Undetermined (stable or uninformative): Common, Rare, Random, No_change-absent
   - Directional trend: Increasing, Decreasing
   - Non-directional trend (episodic or patchy presence): Clumped, Recurrent

These colors match a flowchart designed in Canva for the manuscript.

```{r color-palette}
#colors for the overall groups
group_colors <- c(
  "Undetermined"      = "#705e63",
  "Directional"       = "#e1be6a",
  "Non-directional"   = "#40b0a6"
)

# colors for the classifications using shading
classification_colors <- c(
  "Common"           = "#493D40",
  "Random"           = "#816F74",
  "Rare"             = "#A49297",
  "No_change-absent" = "#615457",
  "Increasing"       = "#947603",
  "Decreasing"       = "#F7D381",
  "Clumped"          = "#197169",
  "Recurrent"        = "#6BD1C7"
)
```


Summarize the results by classification proportion
```{r}
# summarize using the raw values for number of presences
summary_tbl <- results %>%
  count(tslen, npresent, arrangement, group, classification) %>%
  group_by(tslen, npresent, arrangement) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

summary_tbl

# Binned summary: collapse presence proportion into 10 bins (deciles) per
# series length, so plots show ~10 bars/bands instead of one per possible
# npresent value. This is much easier to read and the underlying pattern
# doesn't meaningfully change at this resolution.
summary_binned <- results %>%
  mutate(
    prop_present = npresent / tslen,
    prop_bin = cut(prop_present, breaks = seq(0, 1, by = 0.1),
                   include.lowest = TRUE, right = TRUE)
  ) %>%
  count(tslen, prop_bin, arrangement, group, classification) %>%
  group_by(tslen, prop_bin, arrangement) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

summary_binned
```

One more summary, to look at the total proportions for those present an intermediate amount of time (10%-90%).
```{r}
# keep only those who are not absent, common, or rare
# calculate proportion of total
summary_intermed <- results %>%
  filter(!classification %in% c("Common", "Rare", "No_change-absent")) %>%
  count(tslen, arrangement, group, classification) %>%   # creates the n column
  group_by(tslen, arrangement) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

summary_intermed
```


Now, let's show the results by classification and presence level
```{r plot-random-binned, fig.width=9, fig.height=6}
summary_binned %>%
  filter(arrangement == "random") %>%
  ggplot(aes(x = prop_bin, y = prop, fill = classification)) +
  geom_col(position = "stack") +
  facet_wrap(~ tslen, labeller = label_both) +
  scale_fill_manual(values = classification_colors) +
  labs(
    x = "Proportion of years present (binned)",
    y = "Proportion of runs classified as category",
    title = "Classification outcomes under random ordering of presences"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom") # Places the legend below the plot


ggsave("Figures/simulation-figs/barplot-random-ordering.png", dpi=300, width=9, height=6, bg="white")
```


```{r violin-presence-proportion, fig.width=10, fig.height=6}
results %>%
  mutate(prop_present = npresent / tslen) %>%
  ggplot(aes(x = classification, y = prop_present, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.6, scale = "width") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.05, size = 0.6) +
  facet_wrap(~ tslen, labeller = label_both) +
  scale_fill_manual(values = group_colors) +
  labs(
    x = NULL, y = "Proportion of years present",
    title = "Presence proportion by assigned classification"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

ggsave("Figures/simulation-figs/violin-random-prop-years.png", dpi=300, width=9, height=5, bg="white")
```

```{r violin-transitions, fig.width=10, fig.height=6}
results %>%
  ggplot(aes(x = classification, y = numtransitions, fill = group)) +
  geom_violin(trim = TRUE, alpha = 0.6, scale = "width") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.05, size = 0.6) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "hotpink") +
  facet_wrap(~ tslen, labeller = label_both) +
  scale_fill_manual(values = group_colors) +
  labs(
    x = NULL, y = "Number of incidence blocks",
    title = "Incidence blocks by assigned classification",
    subtitle = "Dashed line marks the Recurrent vs Clumped cutoff"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

ggsave("Figures/simulation-figs/violin-random-nblocks.png", dpi=300, width=9, height=5, bg="white")
```


Now calculate and show the overall proportion of each classification across the different runs
```{r}

prop_by_tslen <- results %>%
  group_by(tslen) %>%
  mutate(total_runs = n()) %>%
  ungroup() %>%
  group_by(tslen, group, classification) %>%
  summarise(
    count      = n(),
    total_runs = first(total_runs),
    .groups    = "drop"
  ) %>%
  mutate(Proportion = count / total_runs)

#bar graph
ggplot(prop_by_tslen, aes(x = Proportion, y = classification)) +
  geom_col(aes(fill = classification)) + 
  facet_wrap(~tslen, labeller = label_both) +
  labs(title = "Random incidence",
       x = "Proportion of timeseries",
       y = "Classification") +
  scale_fill_manual(values = classification_colors) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 10),
        legend.position = "none")

ggsave("Figures/simulation-figs/random_class_incid_bar.png", dpi=300, width=9, height=5, bg="white")
```


