---
title: "EAGER-Plants"
author: "Maya Parker-Smith"
date: "2023-05-15"
output: html_document
 smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---

```{r setup, include=FALSE}
## Clear environment 
rm(list=ls())


##Libraries
library(tidyverse)
library(cowplot)
library(bbmle)
library(lmerTest)
library(car)
library(readxl)
library(multcomp)
library(readr)
library(incidence)
library(vegan)
library(goeveg)
library(randtests)
library(reshape2)
library(lubridate)

```

##Plant Species Composition
This section is dedicated to cleaning the plant species composition/cover dataset. Data was collected from 1983 to 2022. Watersheds that are included: FA, SuB, N4A, R1A, 2D, WB, N20B, 1D, N1A, R1B, SpA, SpB, WA, 20B, 4A, 4F, SuA, N1B, N20A, N4D.

Author: David Hartnett, Scott Collins, Zak Ratajczak

ZIP file: knb.lter.knz.69.21

Data code: PVC02

File(s): PVC021.csv
 Pasture c01b was called Texas Hog (thp) in 2009. 
Vegetation species composition from 1983: The transects were permanently laid out in the current format of 4 transects (A-D), each with 5 plots. 
Transect E only occurred on watershed N20B florence in 1986 and 1987. This transect is the same as the current transect D for this watershed and soil type. 
The old transect D was abandoned in 1987 prior to bison reintroduction. 
Seven cover classes were used to estimate species canopy coverage. 1 - 0-1% cover; 2 - 2-5% cover; 3 - 5-25% cover; 4 - 25-50% cover; 5 - 50-75%; 6 - 75-95% cover; 7 - 95-100% cover. (Note: for the watershed r20b, no data were collected in the transect A and B in the fall of 2011 due to wildfire occurred.) 
Transect 'E' occurred only on the N20b florence site in 1986 and 1987 and was renamed the new 'transect' in 1988. 
The old 'D' transect was sampled from 1983 to 1987 and then abandoned when the bison fence was constructed. 
Thus, plots d1-d5 from 1983-1987 are NOT the same d1-d5 plots in subsequent years.

```{r}

plantcomp_raw <- read_csv("Datasets_EAGER/Plants/knb-lter-knz.69.21_SpeciesComposition/PVC021.csv")

watershed_info <- read_excel("Datasets_EAGER/Other/Watershed Info.xlsx")

unique(plantcomp_raw$WaterShed)

plantcomp <- plantcomp_raw %>% mutate(RecType = factor(RecType),
                                        RecMonth = factor(RecMonth),
                                        RecDay = factor(RecDay),
                                        WaterShed = factor(WaterShed),
                                        SoilType = factor(SoilType),
                                        Transect = factor(Transect),
                                        Plot = factor(Plot),
                                        SpeCode = factor(SpeCode))

#Combining the genus and species column and creating a new column. 
plantcomp1 <- unite(plantcomp, "Genus_spp", AB_genus:AB_species, sep = "_", remove = FALSE)


plantcomp1$WaterShed <- dplyr::recode(plantcomp1$WaterShed,
                                "001c" = "1C",
                                "001d" = "1D",
                                "004b" = "4B",
                                "020b" = "20B",
                                "n01b" = "N1B",
                                "n04d" = "N4D",
                                "n20b" = "N20B",
                                "002c" = "2C",
                                "002d" = "2D",
                                "004a" = "4A",
                                "004f" = "4F",
                                "n04a" = "N4A",
                                "n20a" = "N20A",
                                "020d" = "20D",
                                "n01a" = "N1A",
                                "00wb" = "WB",
                                "0spa" = "SpA",
                                "0spb" = "SpB",
                                "00fa" = "FA",
                                "00fb" = "FB",
                                "00wa" = "WA",
                                "0sua" = "SuA",
                                "0sub" = "SuB",
                                "001a" = "1A",
                                "020a" = "20A",
                                "r01a" = "R1A",
                                "r01b" = "R1B",
                                "r20a" = "R20A",
                                "r20b" = "R20B",
                                "n02a" = "N2A",
                                "n02b" = "N2B")

unique(plantcomp1$WaterShed)

plantcomp1 <- plantcomp1 %>% rename("Watershed_name" = "WaterShed",
                              "Recyear" = "RecYear",
                              "Recmonth" = "RecMonth",
                              "Recday" = "RecDay")


plantcomp2 <- plantcomp1
plantcomp2 <- plantcomp2 %>% subset(.$Recyear >= 1993)


#Removing watersheds and transects I won't use for further analysis
plantcomp3 <- plantcomp2

plantcomp3 <- plantcomp3 %>%
  subset(Watershed_name %in% c("1A", "1C", "1D",
                               "2C", "2D",
                              "4A", "4B", "4F",
                              "20A", "20B", "20D",
                              "N1A", "N1B",
                              "N2A", "N2B",
                              "N4A", "N4D",
                              "N20A", "N20B")) 

plantcomp3 <- left_join(plantcomp3, watershed_info, by = "Watershed_name")
sort(unique(plantcomp3$Genus_spp))
length(unique(plantcomp3$SpeCode))


plantcomp4 <- plantcomp3 

plantcomp4 <- subset(plantcomp4,!(Genus_spp %in% c("annual_forb", "carex_spp.", "cyperu_spp.", 
                                                   "euphor_spp.", "symphy_spp.")))


plantcomp4 <- plantcomp4 %>% unite("Month_Day", Recmonth:Recday, sep = "-")

plantcomp4 <- plantcomp4 %>% unite("Date", Recyear:Month_Day, sep = "-", remove = FALSE)

plantcomp4$Date <- ymd(plantcomp4$Date)
datecheck <- plantcomp4 %>% group_by(Date)%>% reframe(Cover = sum(Cover))

```


```{r}
#This section is for checking the sampling amounts for certain variables/parameters

plantcomp_group1 <- plantcomp3 %>% group_by(Recyear, Watershed_name, Genus_spp) %>% 
  reframe(Cover = sum(Cover))

plantcomp_wide1 <- plantcomp_group1 %>% pivot_wider(names_from = Genus_spp, values_from = Cover)

plantcomp_count1 <- plantcomp_wide1 %>% group_by(Watershed_name) %>% 
  summarise(total_count=n(),.groups = 'drop')


#After looking at the total number of times each watershed is sampled, I decided to cut watersheds 1A, 1C, 20A, 20D, 4F, N1A, and N20A. And keep watersheds 1D, 20B, 4A, 4B, N1A, N1B, N20A, N20B, N4A, and N4D

plantcomp4 <- plantcomp4 %>% subset(Watershed_name %in% c("1D", "20B", "4A", 
                                                          "4B", "N1A", "N1B", 
                                                          "N20A", "N20B", "N4A", 
                                                          "N4D"))

#Checking the months the watershed were sampled in
# plantcomp_group2 <- plantcomp4 %>% group_by(Recyear, Watershed_name, Genus_spp) %>% 
#   reframe(Cover = sum(Cover))
# 
# plantcomp_wide2 <- plantcomp_group2 %>% pivot_wider(names_from = Genus_spp, values_from = Cover)
# 
# plantcomp_count2 <- plantcomp_wide2 %>% group_by(Recyear, Watershed_name) %>% 
#   summarise(total_count=n(),.groups = 'drop')
# 
# 
# plantcomp_count2 %>% ggplot(aes(x = Recyear, y = total_count)) + 
#   geom_point() + 
#   facet_wrap(.~Watershed_name) + 
#   theme(text = element_text(size=12),
#         axis.text.x = element_text(angle=45, hjust=1, size =7))
# 
# plantcomp_count2 %>% ggplot(aes(x = Recyear, y = total_count, color = Recmonth)) + 
#   geom_point(position = "stack") + 
#   facet_wrap(.~Watershed_name) + 
#   theme(text = element_text(size=12),
#         axis.text.x = element_text(angle=45, hjust=1, size =7))
# 
# 
# plantcomp_group3 <- plantcomp5 %>% group_by(Recyear, Recmonth, Watershed_name, Genus_spp) %>% 
#   reframe(Cover = sum(Cover))
# 
# plantcomp_wide3 <- plantcomp_group3 %>% pivot_wider(names_from = Genus_spp, values_from = Cover)
# 
# plantcomp_count3 <- plantcomp_wide3 %>% group_by(Watershed_name) %>% 
#   summarise(total_count=n(),.groups = 'drop')


```

```{r}

# 
# spp_grouped <- plantcomp4 %>% group_by(SpeCode, AB_genus, AB_species, Genus_spp) %>% summarise(total_count=n(),.groups = 'drop')
# 
# spp_grouped1 <- spp_grouped[order(spp_grouped$Genus_spp),]
# 
# length(unique(spp_grouped$Genus_spp))
# 
# species_tbl <- plantcomp4 %>% dplyr::select(AB_genus, AB_species, Genus_spp)


```

```{r}
#Dissimilarity matrix for watersheds
plantcomp5 <- plantcomp4 %>% 
  group_by(Watershed_name, Fire_interval, Grazing, Genus_spp) %>% 
  reframe(Cover = sum(Cover))

plantcomp5_wide <- plantcomp5 %>% 
  pivot_wider(names_from = "Genus_spp", 
              values_from = "Cover", 
              values_fn = function(x) paste(sum(x)))

plantcomp5_wide_env <- plantcomp5_wide[, c(1:3)]
plantcomp5_wide1 <- plantcomp5_wide[, -c(1:3)]
plantcomp5_wide1 <- plantcomp5_wide1 %>% mutate_if(is.character, as.numeric)
decostand_tot_plants <- decostand(plantcomp5_wide1, method = "total")
plantcomp5_wide1[is.na(plantcomp5_wide1)] <- 0

diss_matrix <- vegdist(decostand_tot_plants, method = "bray")

```


```{r}

plantcomp_short <- plantcomp4 %>% group_by(Recyear, Watershed_name, Fire_interval, 
                                                Grazing, Genus_spp) %>%  dplyr::reframe(Cover = sum(Cover))


#Pivoting the the dataset wider, so that each species is its own column
plantcomp_wide <- plantcomp_short %>% 
  pivot_wider(names_from = "Genus_spp", values_from = "Cover", 
              values_fn = function(x) paste(sum(x)))

#Making an environment table to add back onto the presence/absence data later
plant_env <- plantcomp_wide[, 1:4]

#Standardizing by presence/absence
decostand_pa_plant <- decostand(plantcomp_wide[, -c(1:4)], "pa")
decostand_pa_plant[is.na(decostand_pa_plant)] <- 0


#Checking the stress of this presence/absence data
dimcheckMDS(decostand_pa_plant, distance = "jaccard", k=5)


#Adding the presence/absence dataset back with the environment dataset
plantcomp_wide2 <- decostand_pa_plant %>% add_column(plant_env, .before = "achill_mille")

#Pivoting the presence/absence back to a longer format
plantcomp_long <- plantcomp_wide2 %>% pivot_longer(cols = 5:352,
                                           names_to = "Species", 
                                           values_to = "Presence")

#Making sure the species are coded as factors
plantcomp_long <- plantcomp_long %>% mutate(Species = factor(Species))


```


```{r}
#Creating another dataset by species and years to calculate persistence
plantcomp_years_wide <- plantcomp_long %>% pivot_wider(names_from = "Recyear", values_from = "Presence")


#Summing the persistence of all years into a new column
plantcomp_years_wide2 <- plantcomp_years_wide %>% mutate(Total_Persistence = 
                                                                rowSums(plantcomp_years_wide[, 5:34]))


plantcomp_years_wide2 <- plantcomp_years_wide2 %>% mutate(Persistence = (.$Total_Persistence/30))

plantcomp_years_wide3 <- plantcomp_years_wide2 %>% 
  mutate(Early_pers = rowSums(plantcomp_years_wide2[,c(5:19)], na.rm = TRUE)) %>%
  mutate(Late_pers = rowSums(plantcomp_years_wide2[,c(20:34)], na.rm = TRUE))

plantcomp_long2 <- pivot_longer(plantcomp_years_wide3, cols = 5:34,
                     names_to = "Recyear",
                     values_to = "Presence")

```


```{r}

#Making a new dataset just in case
plantcomp_cont <- plantcomp_long2

#Adding empty columns onto new dataset
plantcomp_cont <- plantcomp_cont %>% mutate(Presence_cont = NA,
                                            Early_late = NA,
                                            Presence = factor(Presence))



#Filling presence contingency column based on presence/absence
 for(j in 1:nrow(plantcomp_cont)){
   if (plantcomp_cont[j,10] == "1"){plantcomp_cont[j,11] <- "Present"}
   if (plantcomp_cont[j,10] == "0"){plantcomp_cont[j,11] <- "Absent"}
 } 

#Filling early/late contingency based on year
for(i in 1:nrow(plantcomp_cont)){
   if (plantcomp_cont[i,9] < 2008){plantcomp_cont[i,12] <- "Early"}
   if (plantcomp_cont[i,9] >= 2008){plantcomp_cont[i,12] <- "Late"}
}

plantcomp_cont <- plantcomp_cont %>% mutate(Presence_cont = factor(Presence_cont),
                                                  Early_late = factor(Early_late))


summary_tbl <- plantcomp_cont %>%
  group_by(Species, Watershed_name, Fire_interval, 
           Grazing, Persistence, Early_pers, Late_pers)%>%
    dplyr::count(Presence_cont, Early_late)

summary_tbl <- summary_tbl %>% unite("Species_Watershed", Species:Watershed_name, sep = "_") 
summary_tbl <- summary_tbl %>% mutate(Species_Watershed = factor(Species_Watershed))

contabs <- xtabs(n ~ Presence_cont + Early_late + Species_Watershed, summary_tbl)
chisqtests <- apply(contabs, 3, chisq.test)



chisq_p.values <- tibble(names(chisqtests), do.call(rbind,lapply(chisqtests,function(v){v$p.value})))

chisq_p.values <- chisq_p.values %>% rename("chisq_p.values" = "do.call(...)",
                                            "Species_Watershed" = "names(chisqtests)")
chisq_p.values <- chisq_p.values %>% mutate(Species_Watershed = factor(Species_Watershed))




plantcomp_long3 <- plantcomp_long2 %>% 
  group_by(Recyear, Species, Watershed_name, Persistence, Early_pers, Late_pers) %>% reframe(Presence = Presence)


plantcomp_long3 <- plantcomp_long3 %>% unite("Species_Watershed", Species:Watershed_name, sep = "_")

testfun <- function(x){
  runs.test(x, alternative = "left.sided")
}

test.runs <- plantcomp_long3 %>% subset(Species_Watershed == "aristi_oliga_N20A")
runs.test(test.runs$Presence, alternative = "left.sided", plot = TRUE)



presence.tabs <- xtabs(Presence ~ Recyear + Species_Watershed, plantcomp_long3)
runstests <- apply(presence.tabs, 2, testfun)
runstest_p.values <- tibble(names(runstests), do.call(rbind,lapply(runstests,function(v){v$p.value})))
runstest_p.values <- runstest_p.values %>% rename("runstest_p.values" = "do.call(...)",
                                                  "Species_Watershed" = "names(runstests)")
runstest_p.values <- runstest_p.values %>% mutate(Species_Watershed = factor(Species_Watershed))


merged_p.values <- merge(chisq_p.values,runstest_p.values, by = "Species_Watershed", all = TRUE)



```


```{r}

merged_p.values2 <- merged_p.values %>% mutate(Incidence_change = NA,
                                          Sequential_patterns = NA)
group <- summary_tbl %>% dplyr::select(Species_Watershed, Fire_interval, Grazing, Persistence, Early_pers, Late_pers) %>% distinct()

merged_p.values2 <- merge(merged_p.values2, group, by = "Species_Watershed")

for (k in 1:nrow(merged_p.values2)){
  if (is.nan(merged_p.values2$chisq_p.values[k])) {merged_p.values2$chisq_p.values[k] <- 2}
  if (is.na(merged_p.values2$runstest_p.values[k])) {merged_p.values2$runstest_p.values[k] <- 2}
}

merged_p.values3 <- merged_p.values2

for (kk in 1:nrow(merged_p.values3)){
  if (merged_p.values3[kk, 2] == 2){merged_p.values3[kk, 4] <- "-"}
  if (merged_p.values3[kk, 2] <= 0.05){merged_p.values3[kk, 4] <- "Significant"}
  if (merged_p.values3[kk, 2] > 0.05 & merged_p.values3[kk, 2] < 2){merged_p.values3[kk, 4] <- "NS"}
  if (merged_p.values3[kk, 3] == 2){merged_p.values3[kk, 5] <- "-"}
  if (merged_p.values3[kk, 3] <= 0.05){merged_p.values3[kk, 5] <- "Significant"}
  if (merged_p.values3[kk, 3] > 0.05 & merged_p.values3[kk, 3] < 2){merged_p.values3[kk, 5] <- "NS"}
}


merged_p.values4 <- merged_p.values3 %>% mutate(Classification = NA)

for (jj in 1:nrow(merged_p.values4)){
  if (merged_p.values4[jj, 4] == "-" & merged_p.values4[jj, 5] == "-" & merged_p.values4[jj,8] == 0){merged_p.values4[jj, 11] <- "No_change.always_absent"}
  if (merged_p.values4[jj, 4] == "-" & merged_p.values4[jj, 5] == "-" & merged_p.values4[jj,8] == 1){merged_p.values4[jj, 11] <- "No_change.always_present"}
  if (merged_p.values4[jj, 4] == "NS" & merged_p.values4[jj, 5] == "NS"){merged_p.values4[jj, 11] <- "Random"}
  if (merged_p.values4[jj, 4] == "NS" & merged_p.values4[jj, 5] == "Significant"){merged_p.values4[jj, 11] <- "Recurrent"}
  if (merged_p.values4[jj, 4] == "Significant" & merged_p.values4[jj, 5] == "-"){merged_p.values4[jj, 11] <- "Significant"}
  if (merged_p.values4[jj, 4] == "NS" & merged_p.values4[jj, 5] == "-"){merged_p.values4[jj, 11] <- "Random"}
  if (merged_p.values4[jj, 4] == "Significant" & merged_p.values4[jj, 5] == "Significant"){merged_p.values4[jj, 11] <- "Significant"}
  if (merged_p.values4[jj, 4] == "Significant" & merged_p.values4[jj, 5] == "NS"){merged_p.values4[jj, 11] <- "Significant"}
}

merged_p.values5 <- merged_p.values4  

for (xx in 1:nrow(merged_p.values5)){
  if (merged_p.values5[xx, 11] == "Significant" & ((merged_p.values5[xx, 9]) > (merged_p.values5[xx, 10]))){merged_p.values5[xx, 11] <- "Decreasing"}
  if (merged_p.values5[xx, 11] == "Significant" & ((merged_p.values5[xx, 10]) > (merged_p.values5[xx, 9]))){merged_p.values5[xx, 11] <- "Increasing"}
}


for (ii in 1:nrow(merged_p.values5)){
  if (merged_p.values5[ii, 2] == 2) {merged_p.values5$chisq_p.values[ii] <- NA}
  if (merged_p.values5[ii, 3] == 2) {merged_p.values5$runstest_p.values[ii] <- NA}
}

unique(merged_p.values5$Classification)
```


```{r}

plantcomp_long4 <- left_join(plantcomp_long3, merged_p.values5)

plantcomp_long4 <- plantcomp_long4 %>% separate(Species_Watershed, into = c("Genus", "spp", "Watershed_name")) %>% 
  unite("Species", Genus, spp)

plantcomp_long4 <- plantcomp_long4 %>% mutate(Species = factor(Species), Watershed_name = factor(Watershed_name), Recyear = as.numeric(Recyear))

plantcomp_long4$Classification <- factor(plantcomp_long4$Classification, 
                               levels = c("No_change.always_absent", "No_change.always_present", "Random", "Recurrent", "Increasing", "Decreasing"))


#Trying to plot persistence
plantcomp_long4 %>% ggplot(aes(x = Persistence, fill = Watershed_name)) + geom_histogram(stat = "count") + facet_grid(.~Watershed_name) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =10))


plantcomp_long4 %>% ggplot(aes(x = Classification, fill = Watershed_name)) + geom_histogram(stat = "count") + facet_grid(.~Watershed_name) +
  facet_grid(.~Watershed_name) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1, size =10))


plantcomp_long4_short <- plantcomp_long4 %>% subset(Classification %in% c("Recurrent", "Increasing", "Decreasing"))

change_species <- list(plantcomp_long4_short$Species)



```

