---
title: "KNZ-incidence: Cleaning Biotic Datasets"
author: "Maya Parker-Smith"
date: "2023-03-01"
output: html_document
 smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---


# {.tabset .tabset-pills .tabset-fade}

#KNZ-Incidence EAGER Project

## Setup and Overview

### **Setting up data for analysis**
#### This RMarkdown will be used for the inital data cleaning of the four biotic datasets in the EAGER Konza Temporal Incidence project. 

####The four datasets include bird sampling (knb-lter-knz.26.12 -> CB011.csv), grasshopper sweeping (knb-lter-knz.29.20 -> CGR021.csv & CGR022), small-mammal trapping (knb-lter-knz.88.9 -> CSM011.csv & CSM012.csv), and plant data- which includes biomass and species composition (knb-lter-knz.55.14 -> PAB011, and knb-lter-knz.69.21 -> PVC021.csv). All data were taken from the EDI Data Portal Repository. All are LTER datasets from Konza Prairie. Each dataset will have its own section for cleaning

###Libraries and Set-up 
```{r setup, include=FALSE}

## Clear environment 
rm(list=ls())


##Libraries
library(tidyverse)
library(cowplot)
library(bbmle)
library(multcomp)
library(mosaic)
library(readr)

```

## Bird Sampling
This section is dedicated to cleaning the bird-sampling dataset. Data was collected from 1981 to 2009. Watersheds that are supposedly included: N4B, N4D, 4A, N1B, 1D, R20A, R1B, 20C, 20B, N20B.

Author: Alice Boyle

ZIP file: knb-lter-knz.26.12

Data code: CBP01

File(s): CB011.csv

Summary of All Changes Up to 1993: 
1981: Transects were as presently located except no transect in N20B and there were transects in  N20C (now N01A), N20D (now part of N01B), N01C (now N02B), N01D (now part of N01B), and N04C (now N04B). 
1982: Transects in N20C (N01A) and N01C (N02B) were dropped, N20D was maintained in what is now N01B. N01D became the second transect in what is now N01B, and N04C transect was maintained in what is now N04B. N20B was added. 
1990: 004d became SA and transect was dropped.

```{r}
#Read in and view bird data
birds_raw <- read_csv("Datasets_EAGER/Birds/knb-lter-knz.26.12/CBP011.csv")
View(birds_raw)

#Changing classifications and names of columns
birds <- birds_raw %>% mutate(RECYEAR = factor(RECYEAR),
                          RECMONTH = factor(RECMONTH),
                          RECDAY = factor(RECDAY),
                          SEASON = factor(SEASON),
                          TRANSNUM = factor(TRANSNUM),
                          WATERSHED = factor(WATERSHED),
                          OBSNUM = factor(OBSNUM),
                          SPECNAME = factor(SPECNAME),
                          AOUCODE = factor(AOUCODE),
                          COMMONNAME = factor(COMMONNAME),
                          DISTANCE = as.numeric(DISTANCE),
                          COUNT = as.numeric(COUNT), 
                          SEX = factor(SEX),
                          STATUS = factor(STATUS),
                          COMMENTS = as.character(COMMENTS))
glimpse(birds)


#Shortening the dataset to only have the columns we are interested in
birds_1 <- birds %>% dplyr::select(RECYEAR, 
                                       RECMONTH, 
                                       RECDAY, 
                                       SEASON, 
                                       TRANSNUM,
                                       DISTANCE,
                                       WATERSHED, 
                                       OBSNUM, 
                                       SPECNAME,
                                       COMMONNAME,
                                       AOUCODE, 
                                       COUNT,
                                       COMMENTS)


#Checking the species names to make sure there are any weird misspellings
unique(birds_1$SPECNAME)
unique(birds_1$COMMONNAME)
##There are 204 unique species names, but 138 unique common names. I'll probably have to go through the species names to their full species names because I don't know the shortened ones.



#Now checking the watersheds to see exactly what is here compared to what is said in the metadata
unique(birds_1$WATERSHED)
birds_1$WATERSHED <- recode(birds_1$WATERSHED, 
                            "N04D" = "N4D",
                            "N04B" = "N4B",
                            "004B" = "4B",
                            "004A" = "4A",
                            "004D" = "4D",
                            "N01B" = "N1B",
                            "001D" = "1D",
                            "R20A" = "R20A",
                            "R01B" = "R1B",
                            "020C" = "20C",
                            "020B" = "20B",
                            "L00A" = "LA",
                            "S00A" = "SA",
                            "G00A" = "GA",
                            "N20A" = "N20A")



unique(birds_1$COUNT)
#Okay so the count is always "1"... they recorded the data so that every individual sighting has its own row and count number
unique(birds_1$RECMONTH)

```




```{r}
#Creating a tibble that has the info for the watersheds

birds_watershed <- tibble(WATERSHED = c("N4B", "N4D", "4A", "4B",
                                       "N1B", "1D", "R20A", 
                                       "R1B", "20C", "20B", 
                                       "LA", "SA", "GA",
                                       "N20B"), 
                         FIRE_INT = c("4_years", "4_years", "4_years", "4_years",
                                           "1_year", "1_year", "20_years_reversed", 
                                           "1_year_reversed", "20_years", "20_years",
                                           "UNK", "UNK", "UNK", "20_years"),
                         GRAZING = c("Bison", "Bison", "Ungrazed", "Ungrazed",
                                     "Bison", "Ungrazed", "Ungrazed", 
                                     "Ungrazed", "Ungrazed", "Ungrazed",
                                     "UNK", "UNK", "UNK", "Bison"),
                         TYPE = c("Press", "Press", "Pulse", "Pulse",
                                  "UNK", "Pulse", "UNK",
                                  "UNK", "Control", "Control", 
                                  "Lower_Kings_Creek", "Shane_Creek", "North_Kings_Creek", "Press"))

#Combining the bird dataset with the watershed dataset
birds_2 <- left_join(birds_1, birds_watershed, by="WATERSHED")


#Removing the entries from Watershed 4D since it was converted to watershed SA in 1981
##birds_2 <- birds_2 %>% subset(birds_short1$WATERSHED != "004D")
view(birds_2)


```

Looking at the amount of time data was collected for each watershed, transect, in each year, and in each month.
```{r}
#I am going to try to tally up how many times data is collected for each watershed per year.

#Since the original dataset is looking at each individual bird as a separate row/observation, I am grouping them so that the species that have the same RECYEAR, RECMONTH, RECDAY, SEASON, TRANSNUM, and WATERSHED are grouped together
birds_2_group <- birds_2 %>% group_by(RECYEAR, TRANSNUM, RECMONTH, SEASON, WATERSHED, SPECNAME) %>% 
  dplyr::summarise(COUNT = sum(COUNT))


#Making each species a different column so then we can count the number of times data was collected at each watershed in each year
birds_2_group_wide <- birds_2_group %>% pivot_wider(names_from = SPECNAME, values_from = COUNT)

birds_group_count1 <- birds_2_group_wide %>% group_by(WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')


birds_group_count2 <- birds_2_group_wide %>% group_by(RECYEAR, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

birds_group_count2 %>% ggplot(aes(x = RECYEAR, y = total_count, color = WATERSHED)) + geom_jitter(width = 0.2, height = 0.2)


birds_group_count3 <- birds_2_group_wide %>% group_by(SEASON, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

birds_group_count4 <- birds_2_group_wide %>% group_by(SEASON, RECMONTH, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

birds_group_count5 <- birds_2_group_wide %>% group_by(TRANSNUM, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

```


### Plots

```{r}

#Looking at the distribution of bird observations by year
birds_2 %>% ggplot(aes(x = RECYEAR, y = COUNT)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


#Looking at the distribution of bird observations by watershed
birds_2 %>% ggplot(aes(x = WATERSHED, y = COUNT, fill = TYPE)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


#Looking at press vs pulse observations because why not
##Grouping by species first to get the total observations per species per year
birds_2_group <- birds_short2 %>% 
  group_by(RECYEAR, TYPE, SPECNAME) %>% 
  dplyr::summarise(COUNT = sum(COUNT))

##Summarizing the summed species observations by mean per year
birds_2_group <- birds_2_group %>%
  dplyr::summarise(MEAN_COUNT = mean(COUNT))


#Plotting the mean per year 
birds_short2_grouping %>% ggplot(aes(x = RECYEAR, y = Mean_count, group = Type)) + 
  geom_line(aes(color = Type)) + geom_point(aes(color = Type)) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


##Taking away the unknowns and forests and replotting

birds_short2_grouping1 <- subset(birds_short2_grouping, birds_short2_grouping$Type %in% c("Control","Press","Pulse"))

birds_short2_grouping1 %>% ggplot(aes(x = RECYEAR, y = Mean_count, group = Type)) + 
  geom_line(aes(color = Type)) + geom_point(aes(color = Type)) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))

#All of this needs to be redone because I need to standardize using vegan


```


##Grasshopper Sweeping
This section is dedicated to cleaning the grasshopper sweeping dataset. Data was collected from 1982 to 2020. Watersheds that are supposedly included: 2D, 1D, N20B, N1B, SuB, 4F, 20B, N4D, 2C, SpB, 4B, 4A, N1A, N20A.

Author: Anthony Joern

ZIP file: knb-lter-knz.29.20

Data code: CGR02

File(s): CGR021.csv, CGR022.csv, CGR023.csv

1982-1987: Sites were sampled at various earlier dates in addition to the late July-early August. 
1985:Sweeping was restricted to all sites being sampled (twice, on different dates) in late July and early August. 
Additional watersheds (002D, 004D (now 0SpB), 004F AND 010D (now 0SuB) were added to the sampling regime for early August to provide more long-term data on the influence of fire frequency on grasshoppers. 
Sampling on watersheds to be grazed (N01B, N04D, and N20B) was discontinued. 
1986: Watershed 004G (now 00WB) was temporarily added. 
Sampling in June and early July was reduced to watersheds 001D, 004D, 010D and 020B only; too few grasshoppers are collected by sweeping in the first half of the summer for all watersheds to merit sampling. 
1987: Sampling in June and early July was restricted to sites 001D, 002C, 004B and 004F. 
1994:Fire regime changed for 004D (became 0SpB) and 010D (became 0SuB). In the years when 0SuB is burned the sweeps are done 2 weeks earlier than normal. 
The summer burn is conducted on the first water appropriate day in late July to early August. 
1996: Wildfire in February, burned 004F, 0SuB, 001D and 002D. 
1998: 0SuB done earlier than other sites (mid-July) under the mistaken idea that the summer burns were to occur this year. 
2002: Grazed (bison) transects were added in N01A, N01B, N04A, N04D, N20A and N02B. An older version of the methods manual indicates that 3 lowland (Tully) sites were once done. Locations were: 001D A: T-28; 004B A: G-28; B: S-28; B: F-28; 020B A: N-29; B: N-29 
2011: WindMate 200 replaced previous machine used for checking wind speeds. The previous machine had limit of detection of 5mph. WindMate 200 specifications: Temperature -20 oC to 158 oF accuracy +1.8 oC to 89 mph accuracy +3%. 
2013: Beginning with this year, Oecanthinae spp., Gryllidae spp., Tettigoniidae spp., are being added to our list. These are not new species to Konza. We are adding them to the official listing because they are related and ecologically similar. 
2016: In years when SuB is burned, pre-burn and post-burn sweeps will be done. Timing for pre-burn will be mid-July and timing for post-burn will be mid-September. 
2018: March 14. 004B – A was burned in a wildfire. A third sweep, “C”, will be done in the vicinity of pab011 FC until 2021 when 004b is scheduled to burn per “normal” schedule.



```{r}

#First dataset is about the environmental factors at the sweeping sites
grasshoppers_env <- read_csv("Datasets_EAGER/Grasshoppers/knb-lter-knz.29.20/CGR021.csv")

#Second dataset is the grasshopper species counts from each set of sweeps
grasshoppers_raw <- read_csv("Datasets_EAGER/Grasshoppers/knb-lter-knz.29.20/CGR022.csv")

#Third dataset is the counts of instars and male and female grasshoppers 
grasshoppers_other <- read_csv("Datasets_EAGER/Grasshoppers/knb-lter-knz.29.20/CGR023.csv")

view(grasshoppers_raw)

```

```{r}

#Making the variables into factors for later graphing and analysis
grasshoppers <- grasshoppers_raw %>% mutate(RECTYPE = factor(RECTYPE), 
                                          RECYEAR = factor(RECYEAR),
                                          RECMONTH = factor(RECMONTH),
                                          RECDAY = factor(RECDAY),
                                          WATERSHED = factor(WATERSHED),
                                          SOILTYPE = factor(SOILTYPE),
                                          REPSITE = factor(REPSITE),
                                          SPECIES = factor(SPECIES),
                                          SPCODE = factor(SPCODE))


#Checking the number of unique names for different variables to then compare to the metadata
unique(grasshoppers$SPECIES)
##Oh lord... theres capitalizations in some but not others and there are some unknowns. I'll have to alphabetize it and go through to see if there are multiple spellings for the same species

unique(grasshoppers$WATERSHED)
#There is weird capitalizations for these too... I'll have to go through and clean and probably get rid of some


#Renaming watersheds to make it consistent
grasshoppers$WATERSHED <- recode(grasshoppers$WATERSHED, 
                                  "001d" = "1D",
                                  "004b" = "4B",
                                  "000b" = "B",
                                  "n04a" = "N4A",
                                  "n01b" = "N1B",
                                  "n04d" = "N4D",
                                  "n00b" = "NB",
                                  "002d" = "2D", 
                                  "004d" = "4D",
                                  "004f" = "4F",
                                  "010d" = "10D",
                                  "004g" = "4G",
                                  "002c" = "2C",
                                  "0sub" = "SuB",
                                  "020b" = "20B",
                                  "0spb" = "SpB",
                                  "n20a" = "N20A",
                                  "n20b" = "N20B",
                                  "n01a" = "N1A",
                                  "N20B" = "N20B",
                                  "002D" = "2D",
                                  "N01B" = "N1B",
                                  "001D" = "1D",
                                  "N04D" = "N4D",
                                  "004F" = "4F",
                                  "002C" = "2C",
                                  "020B" = "20B",
                                  "0SUB" = "SuB",
                                  "0SPB" = "SpB",
                                  "004B" = "4B",
                                  "N01A" = "N1A",
                                  "N04A" = "N4A",
                                  "N20A" = "N20A")

#Recoding some of the species to make it consistent
grasshoppers$SPECIES <- recode(grasshoppers$SPECIES,
                                "ageneotett deorum" = "Ageneotettix deorum",
                                "arphia conspersa" = "Arphia conspersa",
                                "arphia simplex" = "Arphia simplex",
                                "arphia species" = "Arphia spp.",
                                "arphia spp." = "Arphia spp.",
                                "arphia xanthopte" = "Arphia xanthoptera",
                                "boopedon auriventr" = "Boopedon auriventris",
                                "boopedon gracile" = "Boopedon gracile",
                                "boopedon nubilum" = "Boopedon nubilum",
                                "brachystol magna" = "Brachystola magna", 
                                "campylacan olivacea" = "Campylacantha olivacea",
                                "chortophag viridifas" = "Chortophaga viridifasciata",
                                "encoptolop sordidus" = "Encoptolophus sordidus",
                                "encoptolop spp." = "Encoptolphus spp.",
                                "encoptolop subgracil" = "Encoptolophus subgracilis",
                                "eritettix simplex" = "Eritettix simplex",
                                "hadrotetti trifascia" = "Hadrotettix trifasciatus",
                                "hesperotet species" = "Hesperotettix spp.",
                                "hesperotet speciosus" = "Hesperotettix speciosus",
                                "hesperotet spp." = "Hesperotettix spp.",
                                "hesperotet viridis" = "Hesperotettix viridis",
                                "hippiscus rugosus" = "Hippiscus rugosus",
                                "hypochlora alba" = "Hypochlora alba",
                                "melanoplus angustipe" = "Melanoplus angustipennis",
                                "melanoplus bivittatu" = "Melanoplus bivittatus",
                                "melanoplus confusus" = "Melanoplus confusus",
                                "melanoplus different" = "Melanoplus differentialis",
                                "melanoplus femurrubr" = "Melanoplus femurrubrum",
                                "melanoplus foedus" = "Melanoplus foedus",
                                "melanoplus keeleri" = "Melanoplus keeleri",
                                "melanoplus occidenta" = "Melanoplus occidentalis",
                                "melanoplus packardii" = "Melanoplus packardii",
                                "melanoplus sanguinip" = "Melanoplus sanguinipes",
                                "melanoplus scudderi" = "Melanoplus scudderi",
                                "melanoplus species" = "Melanoplus spp.",
                                "melanoplus spp." = "Melanoplus spp.",
                                "mermiria bivitatta" = "Mermiria bivittata",
                                "mermiria bivittata" = "Mermiria bivittata",
                                "mermiria picta" = "Mermiria picta",
                                "mermiria species" = "Mermiria spp.",
                                "mermiria spp." = "Mermiria spp.",
                                "oedipodinae" = "Oedipodinae spp.",
                                "opeia obscura" = "Opeia obscura",
                                "orphulella speciosa" = "Orphulella speciosa",
                                "orphullela speciosa" = "Orphulella speciosa",
                                "paratylota brunneri" = "Paratylotropidia brunneri",
                                "paratylotr brunneri" = "Paratylotropidia brunneri",
                                "pardalopho haldemani" = "Pardalophora haldemani",
                                "pardalopho apiculata" = "Pardalophora apiculata",
                                "pardalopho species" = "Pardalophora spp.",
                                "pardalopho spp." = "Pardalophora spp.",
                                "phoetaliot nebrascen" = "Phoetaliotes nebrascensis",
                                "pseudopoma brachypte" = "Pseuodopomala brachyptera",
                                "psoloessa delicatul" = "Psoloessa delicatula",
                                "schistocer lineata" = "Schistocerca lineata",
                                "schistocer obscura" = "Schistocerca obscura",
                                "syrbula admirabil" = "Syrbula admirabilis",
                                "unknown" = "Unknown",
                                "xanthippus corallipe" = "Xanthippus corallipes")

grasshoppers$SPECIES <- recode(grasshoppers$SPECIES,
                                " Phoetaliotes nebrascensis" = "Phoetaliotes nebrascensis",
                                " Xanthippus corallipes" = "Xanthippus corallipes")

#Making a new dataset to then condense some of the watersheds based on the metadata
grasshoppers_1 <- grasshoppers

grasshoppers_1$WATERSHED <- recode(grasshoppers_1$WATERSHED,
                                  "B" = "20B",
                                  "NB" = "N20B",
                                  "2D" = "SpB",
                                  "4D" = "SpB",
                                  "4F" = "SuB",
                                  "10D" = "SuB",
                                  "4G" = "WB")

unique(grasshoppers_1$WATERSHED)
unique(sort(grasshoppers_1$SPECIES))

watershed_grasshoppers <- tibble(WATERSHED = c("1D", "4B", "20B", 
                                               "N1B", "N4D", "N20B",
                                               "SpB", "SuB", "WB", 
                                               "2C", "N20A", "N1A", "N4A"), 
                          Fire_interal = c("1_year", "4_years", "20_years", 
                                                  "1_year", "4_year", "20_years", 
                                                  "1_year", "2_years", "1_year", 
                                                  "2_years", "20_years", "1_year", "4_years"),
                          Grazing = c("Ungrazed", "Ungrazed", "Ungrazed", 
                                             "Grazed", "Grazed", "Grazed", 
                                             "Ungrazed", "Ungrazed", "Ungrazed", 
                                             "Ungrazed", "Grazed", "Grazed", "Grazed"))

grasshoppers_2 <- left_join(grasshoppers_1, watershed_grasshoppers, by = "WATERSHED")
view(grasshoppers_2)

unique(grasshoppers_2$RECMONTH)

```

Looking at the amount of time data was collected for each watershed, in each year, and in each month.
```{r}
#I am going to try to tally up how many times data is collected for each watershed per year.

grasshoppers_2_group <- grasshoppers_2 %>% group_by(RECYEAR, RECMONTH, WATERSHED, SPECIES) %>% 
  dplyr::summarise(TOTAL = sum(TOTAL))



#Making each species a different column so then we can count the number of times data was collected at each watershed in each year
grasshoppers_2_group_wide <- grasshoppers_2_group %>% pivot_wider(names_from = SPECIES, values_from = TOTAL)


grasshoppers_group_count1 <- grasshoppers_2_group_wide %>% group_by(WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')


grasshoppers_group_count2 <- grasshoppers_2_group_wide %>% group_by(RECYEAR, WATERSHED) %>% summarise(total_count=n(),.groups = 'drop')

grasshoppers_group_count2 %>% ggplot(aes(x = RECYEAR, y = total_count, color = WATERSHED)) + geom_jitter(width = 0.2, height = 0.2)




grasshoppers_group_count3 <- grasshoppers_2_group_wide %>% group_by(RECYEAR, RECMONTH, WATERSHED) %>% summarise(total_count=n(),.groups = 'drop')

unique(grasshoppers_group_count3$total_count)

```


```{r}

#Looking at the distribution of grasshopper counts per year
grasshoppers2_2 %>% ggplot(aes(x = RECYEAR, y = TOTAL)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


#Looking at the distribution of grasshopper counts by watershed
grasshoppers2_2 %>% ggplot(aes(x = WATERSHED, y = TOTAL, fill = Type)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))

grasshoppers2_2 %>% ggplot(aes(x = RECYEAR, y = TOTAL, fill = WATERSHED)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))



#Looking at press vs pulse observations because why not
##Grouping by species first to get the total observations per species per year
grasshoppers2_2_grouping <- grasshoppers2_2 %>% 
  group_by(RECYEAR, Type, SPECIES) %>% 
  dplyr::reframe(TOTAL = sum(TOTAL))



#Summarizing the summed species observations by mean per year
grasshoppers2_2_grouping_1 <- grasshoppers2_2_grouping %>%
  dplyr::reframe(Mean_total = mean(TOTAL), .by = c(RECYEAR, Type))

grasshoppers2_2_grouping_1 <- subset(grasshoppers2_2_grouping_1, grasshoppers2_2_grouping_1$Type %in% c("Control","Press","Pulse"))



#Plotting the mean per year 
grasshoppers2_2_grouping_1 %>% ggplot(aes(x = RECYEAR, y = Mean_total, group = Type)) + 
  geom_line(aes(color = Type)) + geom_point(aes(color = Type)) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


```



##Small-mammal trapping
This section is dedicated to cleaning the small-mammal trapping dataset. Data was collected from 1981 to 2013. Watersheds that are supposedly included: 4B, 4F, N4D, N20B, 1D, 20B, N1B.

Author: Donald Kaufman

ZIP file: knb-lter-knz.88.9

Data code: CSM01

File(s): CSM011.csv, CSM012.csv

During winter 1981-1982, the Konza Prairie management committee shifted treatment boundaries, so our traplines in N00D were encompassed in the new boundaries of N01D. Therefore, we established two new traplines in another treatment unit (N00B) in spring 1982.
In autumn 1984, four new traplines were established with two traplines in 002C and two in 002D (24 total sampling lines). 
Four more traplines were added in autumn 1985 with two in 010D and two in 001A (28 total sampling lines). 
After 1987, summer sampling was discontinued because of the intensive labor required to close traps in the morning and open traps in late afternoon each day on each trapline in each trapping period.
In 1988, Konza Prairie management committee changed unburned research treatments to treatments with a 20-year frequency of occurrence of fire and, therefore, 000B and N00B became 020B and N20B, respectively. 
The treatment units (N01B and N04D) remained unburned from 1968 until spring 1988 when annual burning was initiated on N01B and the 4-year cycle was initiated on N04D.
Spring fires occurred after our trapping session in these and other treatment units. 
N20B was not burned in 1988, but it had been burned in 1980. 
Before the spring sampling period in 1989, the number of traplines sampled was reduced from 28 traplines to 14; the 14 traplines remaining included two traplines in each of seven experimental treatments (001D, 004B, 004F, 020B, N01B, N04D and N20B).
In addition, a second ungrazed 4-year fire treatment (004F) was continued to help monitor climatic effects on small mammals in prairie experiencing periodic fires
Data for small mammals captured on traplines that were discontinued (001A, 002C, 002D, 004D, 004G, 010A and 010D) can be found in CSM06
In May 1992, gates were opened between phase I and phase II of the bison area. N01B, N04D and N20B lie within the phase II area. 
All sampling periods from autumn 1981 through spring 1992 on these three treatment units occurred on traplines that had not been grazed by bison.


```{r}
#Seasonal summaries of small mammals on LTER traplines 
smallmammals_raw <- read_csv("Datasets_EAGER/Small_mammals/knb-lter-knz.88.9_FourteenTraplines/CSM011.csv")

#Data is for small mammal individual records
smallmammals_indiv <- read_csv("Datasets_EAGER/Small_mammals/knb-lter-knz.88.9_FourteenTraplines/CSM012.csv")


#Making some the comlumns "factors" for later manipulation and analysis
smallmammals <- smallmammals_raw %>% mutate(RECYEAR = factor(RECYEAR),
                                          SEASON = factor(SEASON),
                                          `WATERSHED/LINE` = factor(`WATERSHED/LINE`)) 
#Separating the watershed/line column into two separate columns
smallmammals <- separate(smallmammals, 
                                      `WATERSHED/LINE`, 
                                      into = c("WATERSHED", "LINE"), 
                                      sep = "-")

#Pivoting the table so that each species is a column and each species per year is a different row
smallmammals_1 <- pivot_longer(smallmammals, 
                                          cols = 8:21, names_to = "Species", 
                                          values_to = "Count")

#Renaming the species into their full names
smallmammals_1$Species <- recode(smallmammals_1$Species,
                                     "Pm" = "Peromyscus_maniculatus",
                                     "Rmeg" = "Reithrodontomys_megalotis",
                                     "Sh" = "Sigmodon_hispidus",
                                     "Bh" = "Blarina_hylophaga",
                                     "Rmon" = "Reithrodontomys_montanus",
                                     "St" = "Spermophilus_tridecemlineatus",
                                     "Mo" = "Microtus_ochrogaster",
                                     "Pl" = "Peromyscus_leucopus",
                                     "Ch" = "Chaetodipus_hispidus",
                                     "Mm" = "Mus_musculus",
                                     "Nf" = "Neotoma_floridana",
                                     "Sc" = "Synaptomys_cooperi",
                                     "Zh" = "Zapus_hudsonius",
                                     "Cp" = "Cryptotis_parva")

unique(smallmammals_1$WATERSHED)
smallmammals_1$WATERSHED <- recode(smallmammals_1$WATERSHED,
                                       "004F" = "4F",
                                       "004B" = "4B",
                                       "001D" = "1D",
                                       "020B" = "20B",
                                       "N20B" = "N20B",
                                       "N01B" = "N1B",
                                       "N04D" = "N4D")


watershed_info_smallmammals <- tibble(WATERSHED = c("4F", "4B", "1D",
                                                    "20B", "N20B", "N1B",
                                                    "N4D"),
                                      Fire_interval = c("4_years", "4_years", "1_years",
                                                        "20_years", "20_years", "1_years", 
                                                        "4_years"),
                                      Grazing = c("Ungrazed", "Ungrazed", "Ungrazed", 
                                                  "Ungrazed", "Grazed", "Grazed", 
                                                  "Grazed"))

smallmammals_2 <- left_join(smallmammals_1, watershed_info_smallmammals, by = "WATERSHED")
view(smallmammals_2)

```


```{r}

#Looking at the amount of time data was collected for each watershed, in each year, and in each month.
view(smallmammals)
unique(smallmammals$RECTYPE)

smallmammals_group_count1 <- smallmammals %>%  group_by(WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

smallmammals_group_count2 <- smallmammals %>%  group_by(RECYEAR, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

smallmammals_group_count2 %>% ggplot(aes(x = RECYEAR, y = total_count, color = WATERSHED)) + geom_jitter(width = 0.2, height = 0.2)


smallmammals_group_count3 <- smallmammals %>% 
  group_by(RECYEAR, SEASON, WATERSHED) %>% 
  summarise(total_count=n(),.groups = 'drop')

```


```{r}

#Looking at the total observations by year
smallmammals1_long %>% ggplot(aes(x = RECYEAR, y = Count)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))

smallmammals1_long %>% ggplot(aes(x = WATERSHED, y = Count, fill = Type)) + geom_col() + theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


#Grouping by species and then graphing by type
smallmammals1_grouping <- smallmammals1_long%>% 
  group_by(RECYEAR, Type, Species) %>% 
  dplyr::reframe(Count = sum(Count))


smallmammals1_grouping_1 <- smallmammals1_grouping %>%
  dplyr::reframe(Mean_count = mean(Count), .by = c(RECYEAR, Type))

smallmammals1_grouping_1 <- subset(smallmammals1_grouping_1, smallmammals1_grouping_1$Type %in% c("Control","Press","Pulse"))

smallmammals1_grouping_1 %>% ggplot(aes(x = RECYEAR, y = Mean_count, group = Type)) + 
  geom_line(aes(color = Type)) + geom_point(aes(color = Type)) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1))


```



##Plant Species Composition
This section is dedicated to cleaning the plant species composition/cover dataset. Data was collected from 1983 to 2022. Watersheds that are included: FA, SuB, N4A, R1A, 2D, WB, N20B, 1D, N1A, R1B, SpA, SpB, WA, 20B, 4A, 4F, SuA, N1B, N20A, N4D.

Author: David Hartnett, Scott Collins, Zak Ratajczak

ZIP file: knb.lter.knz.69.21

Data code: PVC02

File(s): PVC021.csv
 Pasture c01b was called Texas Hog (thp) in 2009. 
Vegetation species composition from 1983: The transects were permanently laid out in the current format of 4 transects (A-D), each with 5 plots. 
Transect E only occurred on watershed N20B florence in 1986 and 1987. This transect is the same as the current transect D for this watershed and soil type. 
The old transect D was abandoned in 1987 prior to bison reintroduction. 
Seven cover classes were used to estimate species canopy coverage. 1 - 0-1% cover; 2 - 2-5% cover; 3 - 5-25% cover; 4 - 25-50% cover; 5 - 50-75%; 6 - 75-95% cover; 7 - 95-100% cover. (Note: for the watershed r20b, no data were collected in the transect A and B in the fall of 2011 due to wildfire occurred.) 
Transect 'E' occurred only on the N20b florence site in 1986 and 1987 and was renamed the new 'transect' in 1988. 
The old 'D' transect was sampled from 1983 to 1987 and then abandoned when the bison fence was constructed. 
Thus, plots d1-d5 from 1983-1987 are NOT the same d1-d5 plots in subsequent years.


```{r}

plant_comp_raw <- read_csv("Datasets_EAGER/Plants/knb-lter-knz.69.21_SpeciesComposition/PVC021.csv")

view(plant_comp_raw)

plant_comp <- plant_comp_raw %>% mutate(RecType = factor(RecType),
                                        RecYear = factor(RecYear),
                                        RecMonth = factor(RecMonth),
                                        RecDay = factor(RecDay),
                                        WaterShed = factor(WaterShed),
                                        SoilType = factor(SoilType),
                                        Transect = factor(Transect),
                                        Plot = factor(Plot),
                                        SpeCode = factor(SpeCode))

#Combining the genus and species column and creating a new column. 
plant_comp1 <- add_column(unite(plant_comp, "Genus_spp", AB_genus:AB_species, sep = "_"))

unique(plant_comp1$WaterShed)


plant_comp1$WaterShed <- recode(plant_comp1$WaterShed,
                                "001c" = "1C",
                                "001d" = "1D",
                                "004b" = "4B",
                                "020b" = "20B",
                                "n01b" = "N1B",
                                "n04d" = "N4D",
                                "n20b" = "N20B",
                                "002c" = "2C",
                                "002d" = "2D",
                                "004a" = "4A",
                                "004f" = "4F",
                                "n04a" = "N4A",
                                "n20a" = "N20A",
                                "020d" = "20D",
                                "n01a" = "N1A",
                                "00wb" = "WB",
                                "0spa" = "SpA",
                                "0spb" = "SpB",
                                "00fa" = "FA",
                                "00fb" = "FB",
                                "00wa" = "WA",
                                "0sua" = "SuA",
                                "0sub" = "SuB",
                                "001a" = "1A",
                                "020a" = "20A",
                                "r01a" = "R1A",
                                "r01b" = "R1B",
                                "r20a" = "R20A",
                                "r20b" = "R20B",
                                "n02a" = "N2A",
                                "n02b" = "N2B")

```


```{r}
#Looking at the amount of time data was collected for each watershed, in each year, and in each month.


##grouping by the variables that we are interested in
plant_comp1_group <- plant_comp1 %>% group_by(RecYear, RecMonth, WaterShed, Genus_spp) %>% 
  dplyr::summarise(Cover = sum(Cover))




##pivoting the dataset wider so that each species is a column
plant_comp1_wide <- plant_comp1_group %>% 
  pivot_wider(names_from = Genus_spp, values_from = Cover)


plant_comp1_wide2 <- plant_comp1_wide %>% 
  subset(.$WaterShed =! c("FA", "FB", "WA", "WB", "SpA", "SpB", "SuA", "SuB", "R1A", "R1B", "R20A", "R20B", "4F", "20A", "20D", "N2A", "N2B"))

plant_comp_group_count1 <- plant_comp1_wide %>% group_by(WaterShed) %>% 
  summarise(total_count=n(),.groups = 'drop')


plant_comp_group_count2 <- plant_comp1_wide %>% group_by(RecYear, WaterShed) %>% 
  summarise(total_count=n(),.groups = 'drop')

plant_comp_group_count2 %>% ggplot(aes(x = RecYear, y = total_count, color = WaterShed)) + geom_jitter(width = 0.2, height = 0.2)


plant_comp_group_count3 <- plant_comp1_wide %>% group_by(RecYear, RecMonth, WaterShed) %>% 
  summarise(total_count=n(),.groups = 'drop')

```


##Plant Biomass
This section is dedicated to cleaning the plant bimoass dataset. Data was collected from 1984 to 2018. Watersheds that are included: 1D, R1A, R1B, 2C, 2D, 4A, 4B, 20B, R20A, R20B, SpA, SpB, SuA, SuB, FA, FB, WA, WB.

Author(s): John Blair and Jesse Nippert

ZIP code: knb-lter-knz.55.14

Data code: PAB01

File(s): PAB011.csv

End-of-season biomass samples from 1991 were sorted only into grass, forb, and previous dead (where applicable) categories. Due to an extensive wildfire in spring 1991, most watersheds had no litter accumulation, and 'grass' in 1991 included live grass + current year’s dead grass (no separation due to the large number of additional areas sampled that year and the need to process samples quickly). The April wildfire burned watersheds 001C, 001D, N01B, N01A, 004B, N04D, 020B, N20B, N20A, 002D.Prior to 1992: (for all PAB0_1) Woody plants were NOT separated from forbs. The most common woody species are: Amorpha canescens (lead plant), Ceanothus herbaceous (New Jersey tea), Cornus drummondii (dogwood), Rosa arkansana (wild rose), Rhus glabra (Smooth sumac), and Symphoricarpos orbiculatus (buckbrush).1994: seasonal burning began (PAB031). Prior to this all burning was done in the spring only.1994 and 1995: (PAB021) 020A was burned in a wildfire. 010B was clipped for two years to account for the p.dead component.1997: Watersheds 001A, 020A, 020D, 001C were set up to become part of the Fire Reversal. Experiment (PAB051). Fire Reversal did not take place until 2001.Watersheds 001A and 020A (Tully soils only) began being clipped at the end of the season as part of PAB051 and continued to be clipped biweekly throughout the summer as part of PAB021.Before 1999: All PAB0_1 samples were ground and saved for future chemical analysis. Currently, only 001D, 004B and 020B (PAB011) samples 1, 2 and 3 are ground and saved.1999: Watersheds 020D and 001C began being clipped biweekly as part of PAB021. They also continued to be clipped at the end of the season as part of PAB051.2000: PAB021 was terminated after the 2000 clip season.2001: Fire reversal experiment (PAB051) began. R01A was 020A, R20A was 001A, R01B was 020D, and R20B was 001C.2002: clipping on slope locations terminated for all PAB0_1.Prior to 2002: current year's dead (c.dead) was separated from live grass for all PAB0_1. After this time c.dead is included in live grass weights.

```{r}

plant_biomass_raw <- read_csv("~/Documents/GitHub/KNZ-incidence/Datasets_EAGER/Plants/knb-lter-knz.55.14_AbovegroundBiomass/PAB011.csv")
