---
title: "EAGER-Small_mammals"
author: "Maya Parker-Smith"
date: "2023-04-11"
output: 
  html_document:
    smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---
#  {.tabset .tabset-pills .tabset-fade}

```{r setup, include=FALSE}
## Clear environment 
rm(list=ls())

read.delim()
read_exce

""
##Libraries
library(tidyverse)
library(cowplot)
library(bbmle)
library(lmerTest)
library(car)
library(readxl)
library(multcomp)
library(readr)
library(incidence)
library(vegan)
library(goeveg)
library(randtests)
library(reshape2)
```

#Small-mammal trapping 

###List of changes
This section is dedicated to cleaning the small-mammal trapping dataset. Data was collected from 1981 to 2013. Watersheds that are supposedly included: 4B, 4F, N4D, N20B, 1D, 20B, N1B.

Author: Donald Kaufman

ZIP file: knb-lter-knz.88.9

Data code: CSM01

File(s): CSM011.csv, CSM012.csv

During winter 1981-1982, the Konza Prairie management committee shifted treatment boundaries, so our traplines in N00D were encompassed in the new boundaries of N01D. Therefore, we established two new traplines in another treatment unit (N00B) in spring 1982.
In autumn 1984, four new traplines were established with two traplines in 002C and two in 002D (24 total sampling lines). 
Four more traplines were added in autumn 1985 with two in 010D and two in 001A (28 total sampling lines). 
After 1987, summer sampling was discontinued because of the intensive labor required to close traps in the morning and open traps in late afternoon each day on each trapline in each trapping period.
In 1988, Konza Prairie management committee changed unburned research treatments to treatments with a 20-year frequency of occurrence of fire and, therefore, 000B and N00B became 020B and N20B, respectively. 
The treatment units (N01B and N04D) remained unburned from 1968 until spring 1988 when annual burning was initiated on N01B and the 4-year cycle was initiated on N04D.
Spring fires occurred after our trapping session in these and other treatment units. 
N20B was not burned in 1988, but it had been burned in 1980. 
Before the spring sampling period in 1989, the number of traplines sampled was reduced from 28 traplines to 14; the 14 traplines remaining included two traplines in each of seven experimental treatments (001D, 004B, 004F, 020B, N01B, N04D and N20B).
In addition, a second ungrazed 4-year fire treatment (004F) was continued to help monitor climatic effects on small mammals in prairie experiencing periodic fires
Data for small mammals captured on traplines that were discontinued (001A, 002C, 002D, 004D, 004G, 010A and 010D) can be found in CSM06
In May 1992, gates were opened between phase I and phase II of the bison area. N01B, N04D and N20B lie within the phase II area. 
All sampling periods from autumn 1981 through spring 1992 on these three treatment units occurred on traplines that had not been grazed by bison.



##Reading in the data and basic set-up
```{r}
#Seasonal summaries of small mammals on LTER traplines 
smammals_raw <- read_csv("Datasets_EAGER/Small_mammals/knb-lter-knz.88.9_FourteenTraplines/CSM011.csv")

smammals_raw
#Table including the watershed info for all watersheds
watershed_info <- read_excel("Datasets_EAGER/Other/Watershed Info.xlsx")


#Making some the columns "factors" for later manipulation and analysis
smammals <- smammals_raw %>% mutate(SEASON = factor(SEASON),
                                    `WATERSHED/LINE` = factor(`WATERSHED/LINE`)) 



#Separating the watershed/line column into two separate columns
smammals <- separate(smammals, `WATERSHED/LINE`, into = c("WATERSHED", "LINE"), sep = "-")


#Renaming columns just so that they're not all capitalized
smammals <- smammals %>% rename("Watershed_name" = "WATERSHED", 
                                "Datacode" = "DATACODE",
                                "Rectype" = "RECTYPE", 
                                "Recyear" = "RECYEAR", 
                                "Season" = "SEASON", 
                                "Line" = "LINE")

#Shortening the names of the watersheds
smammals$Watershed_name <- dplyr::recode(smammals$Watershed_name,
                                       "004F" = "4F",
                                       "004B" = "4B",
                                       "001D" = "1D",
                                       "020B" = "20B",
                                       "N20B" = "N20B",
                                       "N01B" = "N1B",
                                       "N04D" = "N4D")



#Pivoting the table so that each species is a column and each species per year is a different row
smammals_long <- pivot_longer(smammals, cols = 8:21, names_to = "Species", values_to = "Count")

#Renaming the species into their full names
smammals_long$Species <- dplyr::recode(smammals_long$Species,
                                     "Pm" = "Peromyscus_maniculatus",
                                     "Rmeg" = "Reithrodontomys_megalotis",
                                     "Sh" = "Sigmodon_hispidus",
                                     "Bh" = "Blarina_hylophaga",
                                     "Rmon" = "Reithrodontomys_montanus",
                                     "St" = "Spermophilus_tridecemlineatus",
                                     "Mo" = "Microtus_ochrogaster",
                                     "Pl" = "Peromyscus_leucopus",
                                     "Ch" = "Chaetodipus_hispidus",
                                     "Mm" = "Mus_musculus",
                                     "Nf" = "Neotoma_floridana",
                                     "Sc" = "Synaptomys_cooperi",
                                     "Zh" = "Zapus_hudsonius",
                                     "Cp" = "Cryptotis_parva")



#Joining the watershed_info dataset with the actual small mammal dataset
smammals_long <- left_join(smammals_long, watershed_info, by = "Watershed_name")


#Making variables factors
smammals_long <- smammals_long %>% mutate(Fire_interval = factor(Fire_interval),
                    Burn_season = factor(Burn_season),
                    Grazing = factor(Grazing),
                    Watershed_name = factor(Watershed_name),
                    Line = factor(Line))


#Separate the dataset from 1992 onwards. 
smammals_long <- smammals_long %>% subset(.$Recyear >= 1992)

smammals_long <- smammals_long %>% subset(Watershed_name != "4F")

#Since I likely won't differentiate the data based on the trap lines, I am summing the data based on just watershed
smammals_long <- smammals_long %>% group_by(Recyear, Watershed_name, 
                                            Species, Fire_interval, Grazing) %>% reframe(Count = sum(Count))

#Removing the 4F watershed to make sampling between all fire/grazing treatments equal (the 4 year fire interval with no grazing combination of disturbances has two watersheds represented, while all other combinations have just one).
####smallmammals_long1 <- smallmammals_long1 %>% subset(Watershed_name != "4F")



```


##Watershed Dissimilarity matrix
We want to check dissimilarity across the different treatments because...
```{r}
#Making dissimilarity matrices for the watersheds - abundance

#Selecting only necessary columns
smammals_group_diss <- smammals_long %>% 
  group_by(Watershed_name, Fire_interval, Grazing, Species) %>% 
  reframe(Total = sum(Count))

#Pivoting the the dataset wider 
smammals_diss <- smammals_group_diss %>% 
  pivot_wider(names_from = "Species", values_from = "Total", 
              values_fn = function(x) paste(sum(x)))


#Making an environment dataset with the classifying variables
smammals_diss_env <- smammals_diss[, c(1:3)]

#Removing the classifying variables for the dissimilarity analysis
smammals_diss <- smammals_diss[, -c(1:3)]



#Making the values numeric instead of characters
smammals_diss <- smammals_diss %>% mutate_if(is.character, as.numeric)

#Standardizing for the total abundance by watershed
decostand_tot_smammals <- decostand(smammals_diss, method = "total")

#Creating the dissimilarity matrix
diss_matrix <- vegdist(decostand_tot_smammals, method = "bray")

```

##Changing to presence/absence
```{r}

#Pivoting the the dataset wider, so that each species is its own column
smammals_wide <- smammals_long %>% 
  pivot_wider(names_from = "Species", values_from = "Count", 
              values_fn = function(x) paste(sum(x)))

#Making an environment table to add back onto the presence/absence data later
smammals_env <- smammals_wide[, 1:4]

#Standardizing by presence/absence
decostand_pa_smammals <- decostand(smammals_wide[, -c(1:4)], "pa", na.rm = TRUE )


#Checking the stress of this presence/absence data
dimcheckMDS(decostand_pa_smammals, distance = "jaccard", k=5)


#Adding the presence/absence dataset back with the enviornment dataset
smammals_wide <- decostand_pa_smammals %>% add_column(smammals_env, .before = "Blarina_hylophaga")

#Pivoting the presence/absence back to a longer format
smammals_long_pa <- smammals_wide %>% pivot_longer(cols = 5:18, 
                                                   names_to = "Species", 
                                                   values_to = "Presence")

#Making sure the species are coded as factors
smammals_long_pa <- smammals_long_pa %>% mutate(Species = factor(Species))

```


##Calculating persistence
```{r}

smammals_years_wide %>% ggplot(aes(x = Species, y = Persistence, fill = Species)) + 
  geom_col() +  
  facet_grid(Watershed_name~.) +
  theme(text = element_text(size=10))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 10)) 


# find number of years in the dataset
totalyears <- length(unique(smammals_long_pa$Recyear))

#Creating another dataset by species and years to calculate persistence
smammals_years_wide <- smammals_long_pa %>% 
  pivot_wider(names_from = "Recyear", values_from = "Presence")

#Making NA's zeroes... MIGHT HAVE TO UNDO THIS
smammals_years_wide <- smammals_years_wide %>% 
  mutate_all(~replace(., is.na(.), 0))

#Summing the amount of years each species was present at each watershed 
smammals_years_wide <- smammals_years_wide %>% 
  mutate(Total_Presence = rowSums(smammals_years_wide[, 5:26]))

#Dividing the number each species was present by the total number of years in the dataset
smammals_years_wide <- smammals_years_wide %>% 
  mutate(Persistence = (.$Total_Presence/totalyears))

#Summing the number of times each species occurs in the early half of the dataset and later half of the dataset for the incidence classification later on
smammals_years_wide <- smammals_years_wide %>% 
  mutate(Early_pres = rowSums(smammals_years_wide[,c(5:15)], na.rm = TRUE)) %>%
  mutate(Late_pres = rowSums(smammals_years_wide[,c(16:26)], na.rm = TRUE))


#Pivoting the dataset back to long format for further analysis; this dataset included the persistence calculations so that I can classify always_absent, always_present, increasing, and decreasing incidence later on
smammals_long_pers <- pivot_longer(smammals_years_wide, cols = 5:26,
                     names_to = "Recyear",
                     values_to = "Presence")



#Plotting persistence in a simple bar graph
smammals_years_wide %>% ggplot(aes(x = Species, y = Persistence, fill = Species)) + 
  geom_col() +  
  facet_grid(Watershed_name~.) +
  theme(text = element_text(size=10))+
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 10)) 

```


##Chi-square contingency tables
Move the notes that describes what this is and what it tells us about the ecology and the incidence of the species from the code chunk to here in the markdown part.
```{r}

#Alright this section is for trying the contingency table and chi-square and Wald-Wolfowitz runs test. But I dunno what I'm doing so I am just hoping for the best now. 

#Species incidence classification is determined by combined results from a contingency table with a chi-squared test and a runs test (Figs 3-4; ,8). Species present in all years are assigned “no change” and their sequential incidences are not tested. The following steps will be completed on each sample-based rarefaction run of each assemblage, and values will be compared to ensure that classification is robust to sampling error.
#Looking at the amount of time data was collected for each watershed, in each year, and in each month.



#Okay so I think I need to make a contingency table by splitting the time series into early/late and presence/absence 


#Making a new dataset just in case #FIXME: can probably get rid of this line
smammals_cont <- smammals_long_pers

#Adding empty columns onto new dataset
smammals_cont <- smammals_cont %>% mutate(Presence_cont = NA,
                                          Early_late = NA,
                                          Recyear = as.numeric(Recyear))

# finds the median (middle) year in the dataset
# if there are an even number of years, separate equally
# if there are an odd number of years, the late mid year is counted as late #FIXME: check
midpoint <- ceiling(median(unique(smammals_cont$Recyear)))

#Filling early/late contingency based on year
for(i in 1:nrow(smammals_cont)){
   if (smammals_cont[i,9] < midpoint){smammals_cont[i,12] <- "Early"}
   if (smammals_cont[i,9] >= midpoint){smammals_cont[i,12] <- "Late"}
}

#Converting the new presence/absence and early/late columns into factors
smammals_cont <- smammals_cont %>% mutate(Early_late = factor(Early_late))


#Creating a summary table to convert to matrices for chi.sq analysis
summary_tbl <- smammals_cont %>%
  group_by(Species, Watershed_name)%>%
    dplyr::count(Presence, Early_late)

#Uniting the species and watershed columns so there is a unique identifier column
summary_tbl <- summary_tbl %>% unite("Species_watershed", Species:Watershed_name, sep = "_") 
#Converting the new Species_watershed column into a factor
summary_tbl <- summary_tbl %>% mutate(Species_watershed = factor(Species_watershed))


#Creating the contingency matrices 
contabs <- xtabs(n ~ Presence + Early_late + Species_watershed, summary_tbl)
#Applying chi-sq tests to each individual matrix (MARGIN = 3 means it is looking at rows and columns)
chisqtests <- apply(contabs, 3, chisq.test)
#Extracting all the p.values from the chi.sq tests into a table
chisq_p.values <- tibble(names(chisqtests), do.call(rbind,lapply(chisqtests,function(v){v$p.value})))

#Renaming the variables from the chi-square p.value table
chisq_p.values <- chisq_p.values %>% rename("chisq_p.values" = "do.call(...)",
                                            "Species_watershed" = "names(chisqtests)")

#Just doing a test
test <- matrix(c(7,4,  9,2), nrow = 2, ncol = 2, byrow = TRUE,
       dimnames = list(c("absent", "present"),
                       c("early", "late")))

chisq.test(test)

```

##Runs-test 
```{r}

#Selecting the variables I need for the runs.test and again combining the species and watershed to create unique identifiers
smammals_long_runs <- smammals_long_pers %>% 
  group_by(Recyear, Species, Watershed_name) %>% reframe(Presence = Presence) %>%
  unite("Species_watershed", Species:Watershed_name, sep = "_")

#Creating the runs.test matrices 
presence.tabs <- xtabs(Presence ~ Recyear + Species_watershed, smammals_long_runs)
#Applying runs.tests to the matrices (MARGIN = 2 means that they are looking at the columns)
runstests <- apply(presence.tabs, 2, runs.test)
#Extracting the p.values from the runs.test into a table
runstest_p.values <- tibble(names(runstests), do.call(rbind,lapply(runstests,function(v){v$p.value})))
#Renaming the variables from the runs.test p.value table
runstest_p.values <- runstest_p.values %>% rename("runstest_p.values" = "do.call(...)",
                                                  "Species_watershed" = "names(runstests)")

```

##Incidence classification
```{r}

#Merging the p.value outputs from the chi.sq and runs.tests 
merged_p.values <- merge(chisq_p.values, runstest_p.values, by = "Species_watershed", all = TRUE)


#Separating the combined Species_watershed again so that I can join these p.values to previous datasets
merged_p.values1 <- merged_p.values %>% separate(Species_watershed, into = c("Species", "spp", "Watershed_name")) %>% 
  unite("Species", Species, spp)

#Joining the merged p.values with a previous dataset to get all the variables
merged_p.values1 <- left_join(merged_p.values1, (dplyr::select(smammals_years_wide, c(Watershed_name, Fire_interval, Grazing, Species, Persistence, Early_pres, Late_pres))), by = c("Watershed_name", "Species"))


#Copying the merged p.value dataset
merged_p.values2 <- merged_p.values1
#Adding new empty columns, one to indicate incidence change (chi.sq significance) and one to indicate sequential patterns(runs.test significance)
merged_p.values2 <- merged_p.values2 %>% mutate(Incidence_change = NA,
                                          Sequential_patterns = NA)

#Avoiding the NA/NaN error by changing all NA's/NaN to "2" (will change back later)
for (k in 1:nrow(merged_p.values2)){
  if (is.nan(merged_p.values2$chisq_p.values[k])) {merged_p.values2$chisq_p.values[k] <- 2}
  if (is.nan(merged_p.values2$runstest_p.values[k])) {merged_p.values2$runstest_p.values[k] <- 2}
}

#Copying the dataset again (this is all just for caution)
merged_p.values3 <- merged_p.values2

#Marking the incidence change and sequential patterns with either a "significant" (p.value < 0.05), a "-" (in the original output, p-value was NaN), or a "NS" (not significant, p.value > 0.05)
for (kk in 1:nrow(merged_p.values3)){
  if (merged_p.values3$chisq_p.values[kk] == 2){merged_p.values3$Incidence_change[kk] <- "-"}
  if (merged_p.values3$chisq_p.values[kk] <= 0.05){merged_p.values3$Incidence_change[kk] <- "Significant"}
  if (merged_p.values3$chisq_p.values[kk] > 0.05 & merged_p.values3$chisq_p.values[kk] < 2){merged_p.values3$Incidence_change[kk] <- "NS"}
  if (merged_p.values3$runstest_p.values[kk] == 2){merged_p.values3$Sequential_patterns[kk] <- "-"}
  if (merged_p.values3$runstest_p.values[kk] <= 0.05){merged_p.values3$Sequential_patterns[kk] <- "Significant"}
  if (merged_p.values3$runstest_p.values[kk] > 0.05 & merged_p.values3$runstest_p.values[kk] < 2){merged_p.values3$Sequential_patterns[kk] <- "NS"}
}

#Copying the dataset again and adding a classification column (this is all just for caution)
merged_p.values4 <- merged_p.values3 %>% mutate(Classification = NA)

#Classifying the incidence based on the "incidence change" and "sequential patterns" columns
for (jj in 1:nrow(merged_p.values4)){
  if (merged_p.values4$Incidence_change[jj] == "-" & merged_p.values4$Sequential_patterns[jj] == "-" & merged_p.values4$Persistence[jj] == 0){merged_p.values4$Classification[jj] <- "No_change.always_absent"} #Chi.sq and runs.tests p.value are NaN; persistence is "0"
  
  if (merged_p.values4$Incidence_change[jj] == "-" & merged_p.values4$Sequential_patterns[jj] == "-" & merged_p.values4$Persistence[jj] == 1){merged_p.values4$Classification[jj] <- "No_change.always_present"} #Chi.sq and runs.tests p.value are NaN; persistence is "1"
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "NS"){merged_p.values4$Classification[jj] <- "Random"} #Chi.sq and runs.test p.values are both > 0.05 
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "Significant"){merged_p.values4$Classification[jj] <- "Recurrent"} #Chi.sq p.value is > 0.05; runs.test p.value is < 0.05
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "-"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test p.value is NaN
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "-"){merged_p.values4$Classification[jj] <- "Random"} #Chi.sq p.value is > 0.05; runs.test is NaN
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "Significant"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test is < 0.05
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "NS"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test p.value > 0.05
}

#Copying the dataset again (this is all just for caution)
merged_p.values5 <- merged_p.values4  

#Classifying the Chi.sq p.values that are < 0.05 into 'decreasing' and 'increasing' based on its early presence vs late presence
for (xx in 1:nrow(merged_p.values5)){
  if (merged_p.values5$Classification[xx] == "Significant" & ((merged_p.values5$Early_pres[xx]) > (merged_p.values5$Late_pres[xx]))){merged_p.values5$Classification[xx] <- "Decreasing"} #Chi.sq p.value is < 0.05; early persistence is more than late presence
  if (merged_p.values5$Classification[xx] == "Significant" & ((merged_p.values5$Late_pres[xx]) > (merged_p.values5$Early_pres[xx]))){merged_p.values5$Classification[xx] <- "Increasing"} #Chi.sq p.value is < 0.05; early persistence is less than late presence
}

#Reverting the "2" for the p.values back into NAs
for (ii in 1:nrow(merged_p.values5)){
  if (merged_p.values5$chisq_p.values[ii] == 2) {merged_p.values5$chisq_p.values[ii] <- NA}
  if (merged_p.values5$runstest_p.values[ii] == 2) {merged_p.values5$runstest_p.values[ii] <- NA}
}

```


##Exploration Plots
```{r}

#Joining the merged p.value dataset that includes classification with a previous dataset to have all variables (year, fire-interval, grazing, etc.)
smammals_class <- left_join(smammals_long_pers, merged_p.values5)

#Reordering the classification variables for plotting
smammals_class$Classification <- factor(smammals_class$Classification, 
                               levels = c("No_change.always_absent", "No_change.always_present", 
                                          "Random", "Recurrent", "Increasing"))

#Making the zeroes in the presence column into NAs for plotting (that way they don't show up on the graph)
smammals_class$Presence[smammals_class$Presence == 0] <- NA
#Making sure the Recyear is coded as numeric 
smammals_class <- smammals_class %>% mutate(Recyear = as.numeric(Recyear))

#Reordering the fire intervals for plotting
smammals_class$Fire_interval <- factor(smammals_class$Fire_interval, 
                                           levels = c("1_year", "4_years", "20_years"))
#Reordering the grazing for plotting
smammals_class$Grazing <- factor(smammals_class$Grazing,
                                     levels = c("Ungrazed", "Bison"))
#Reordering the watersheds for plotting
smammals_class$Watershed_name <- factor(smammals_class$Watershed_name,
                                        levels = c("1D", "N1B", "4B", "4F", "N4D", "20B", "N20B"))

#Creating a graph to look at the classification by species and presence/absence by year
smammals_class %>% ggplot(aes(x = Recyear, y = Presence, fill = Classification)) + 
  geom_tile(width = .6, height = .6) + 
  facet_grid(Species~Watershed_name, switch = "y") +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =12),
        axis.text.y = element_text(size=12),
        strip.text.y.left = element_text(angle = 0, size = 10),
        strip.text.x.top = element_text(size = 10)) +
  geom_vline(xintercept = 2002.5, linetype= 2) +
  labs(x = "Year", y = "Presence") +
  scale_fill_manual(values = c("NA", "black", "dodgerblue2", "darkorchid", "darkolivegreen4", "firebrick3"))

  
smammals_class %>% ggplot(aes(x = Recyear, y = Presence, fill = Classification)) + 
  geom_tile(width = .9, height = .9) + 
  facet_grid(Species~Grazing~Fire_interval, switch = "y") +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =7),
        axis.text.y = element_text(size=7),
        strip.text.y.left = element_text(angle = 0)) +
  geom_vline(xintercept = 2002.5) +
  labs(x = "Year", y = "Presence", title = "Small mammals: incidence classification by species") 



```



