---
title: "EAGER-Birds"
author: "Maya Parker-Smith"
date: "2023-05-01"
output: html_document
 smart: no
    theme: flatly
    float: yes
    css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
---
#  {.tabset .tabset-pills .tabset-fade}

```{r setup, include=FALSE}
## Clear environment 
rm(list=ls())

##Libraries
library(tidyverse)
library(cowplot)
library(bbmle)
library(lmerTest)
library(car)
library(readxl)
library(multcomp)
library(readr)
library(incidence)
library(vegan)
library(goeveg)
library(randtests)
library(reshape2)
library(psych)

```



## Bird Sampling
This section is dedicated to cleaning the bird-sampling dataset. Data was collected from 1981 to 2009. Watersheds that are supposedly included: N4B, N4D, 4A, N1B, 1D, R20A, R1B, 20C, 20B, N20B. Censuses are conducted two times during the year: during the first two weeks of January as
a measure of wintering populations, and during the first two weeks of June as a measure of
breeding populations.

Author: Alice Boyle

ZIP file: knb-lter-knz.26.12

Data code: CBP01

File(s): CBP011.csv

Summary of All Changes Up to 1993: 
1981: Transects were as presently located except no transect in N20B and there were transects in  N20C (now N01A), N20D (now part of N01B), N01C (now N02B), N01D (now part of N01B), and N04C (now N04B). 
1982: Transects in N20C (N01A) and N01C (N02B) were dropped, N20D was maintained in what is now N01B. N01D became the second transect in what is now N01B, and N04C transect was maintained in what is now N04B. N20B was added. 
1990: 004d became SA and transect was dropped.

##Reading in the data and basic set-up
```{r}
#Read in and view bird data
birds_raw <- read_csv("Datasets_EAGER/Birds/knb-lter-knz.26.12/CBP011.csv")
watershed_info <- read_excel("Datasets_EAGER/Other/Watershed Info.xlsx")
View(birds_raw)


#Changing classifications and names of columns
birds_all <- birds_raw %>% mutate(RECMONTH = factor(RECMONTH),
                          RECDAY = factor(RECDAY),
                          SEASON = factor(SEASON),
                          TRANSNUM = factor(TRANSNUM),
                          SPECNAME = factor(SPECNAME),
                          AOUCODE = factor(AOUCODE),
                          COMMONNAME = factor(COMMONNAME),
                          DISTANCE = as.numeric(DISTANCE),
                          COUNT = as.numeric(COUNT),
                          SEX = factor(SEX),
                          STATUS = factor(STATUS),
                          COMMENTS = as.character(COMMENTS))

#Renaming the columns so they're not all capitalized
birds_all <- birds_all %>% rename("Watershed_name" = "WATERSHED",
                          "Datacode" = "DATACODE",
                          "Recyear" = "RECYEAR",
                          "Recmonth" = "RECMONTH",
                          "Recday" = "RECDAY",
                          "Season" = "SEASON",
                          "Transect" = "TRANSNUM",
                          "Observation_number" = "OBSNUM",
                          "Species" = "SPECNAME",
                          "AOU_code" = "AOUCODE",
                          "Common_name" = "COMMONNAME",
                          "Distance" = "DISTANCE",
                          "Count" = "COUNT",
                          "Sex" = "SEX",
                          "Status" = "STATUS",
                          "Comments" = "COMMENTS",
                          "Time" = "TIME",
                          "Duration" = "DURATION",
                          "Observer" = "OBSERVER")

#Renaming the watersheds to make the names shorter
birds_all$Watershed_name <- dplyr::recode(birds_all$Watershed_name,
                                       "N04D" = "N4D",
                                       "N04B" = "N4B",
                                       "004B" = "4B",
                                       "004A" = "4A",
                                       "004D" = "4D",
                                       "N01B" = "N1B",
                                       "001D" = "1D",
                                       "R20A" = "R20A",
                                       "R01B" = "R1B",
                                       "020C" = "20C",
                                       "020B" = "20B",
                                       "L00A" = "LA",
                                       "S00A" = "SA",
                                       "G00A" = "GA",
                                       "N20B" = "N20B")

#Joining the watershed_info dataset with the actual birds dataset
birds_all <- left_join(birds_all, watershed_info, by = "Watershed_name")

#Converting some of the variables into factors
birds_all <- birds_all %>% mutate(Fire_interval = factor(Fire_interval),
                                  Burn_season = factor(Burn_season),
                                  Grazing = factor(Grazing))


#I need to separate the dataset from 1992 onwards. 
birds <- birds_all %>% subset(.$Recyear >= 1992)


#Removing watersheds and transects I won't use for further analysis
birds <-  birds %>% subset(!(Watershed_name %in% c("4D", "R20A", "20C",
                                                     "R1B", "LA",
                                                     "GA", "SA")))
##This transect is being removed to not duplicate the observations at a single watershed; watershed N1B has two transects (6, 10) and all other watersheds just have one transect
###I did do a Bray-Curtis dissimilarity test on the two different transects to make sure they were similar enough, and it looks like they are about 23% dissimilar (~77% similar)
birds <-  birds %>% subset(Transect != "10")


#Also removing observations that are unspecified
birds <-  subset(birds, !(AOU_code %in% c("NONE", "VOID", "UNSP", "UNEM", "RSTO")))
birds$AOU_code <- dplyr::recode(birds$AOU_code,
                                "eawp" = "EAWP")

#Checking the lengths of the species (the 'species' variable has a lot more entries, but they are duplicates. So I'll use the AOU code instead)
length(unique(birds$Species))
length(unique(birds$AOU_code))


```

##Watershed Dissimilarity matrix
```{r}

##Creating a Bray-Curtis dissimilarity matrix for the watersheds
birds_diss <- birds %>% group_by(Watershed_name, Fire_interval, Grazing, AOU_code) %>% 
  reframe(Total = sum(Count))


birds_diss <- birds_diss %>% 
  pivot_wider(names_from = "AOU_code", values_from = "Total", 
              values_fn = function(x) paste(sum(x)))


birds_diss_env <- birds_diss[, c(1:3)]
birds_diss <- birds_diss[, -c(1:3)]
birds_diss <- birds_diss %>% mutate_if(is.character, as.numeric)
birds_diss[is.na(birds_diss)] <- 0

decostand_tot_birds <- decostand(birds_diss, method = "total")

diss_matrix <- vegdist(decostand_tot_birds, method = "bray")

```


##Changing to presence/absence
```{r}

#Selecting the columns I want for further analysis 
birds_group <- birds %>% group_by(Recyear, Watershed_name, Fire_interval, Grazing, AOU_code) %>% 
  reframe(Count = sum(Count))


#Pivoting the the dataset wider, so that each species is its own column
birds_wide <- birds_group %>% 
  pivot_wider(names_from = "AOU_code", values_from = "Count", 
              values_fn = function(x) paste(sum(x)))


#Making an environment table to add back onto the presence/absence data later
birds_env <- birds_wide[, 1:4]


#Standardizing by presence/absence
decostand_pa_birds <- decostand(birds_wide[, -c(1:4)], "pa", na.rm = FALSE)
decostand_pa_birds[is.na(decostand_pa_birds)] <- 0

#adding the presence/absence dataset back with the environment dataset
birds_wide <- cbind(birds_env, decostand_pa_birds)


#Pivoting the presence/absence back to a longer format
birds_pa <- birds_wide %>% pivot_longer(cols = -c(1:4),
                                           names_to = "AOU_code",
                                           values_to = "Presence")

#making the species names and watershed names into factors
birds_pa <- birds_pa %>% mutate(AOU_code = factor(AOU_code))
birds_pa <- birds_pa %>% mutate(Watershed_name = factor(Watershed_name))

```


##Calculating persistence
```{r}

#Using the wide dataset to calculate species incidence?
birds_years_wide <- birds_pa %>% pivot_wider(names_from = factor("Recyear"), values_from = "Presence")

#Summing the amount of years each species was present at each watershed
birds_years_wide <- birds_years_wide %>% mutate(Total_Presence = 
                                                  rowSums(birds_years_wide[, 5:22]))

#Dividing the number of years each species was present by the total number of years in the dataset
birds_years_wide <- birds_years_wide %>% mutate(Persistence = (Total_Presence/18))

#Summing the number of times each species occurs in the early half of the dataset and later half of the dataset for the incidence classification later on
birds_years_wide <- birds_years_wide %>% 
  mutate(Early_pres = rowSums(birds_years_wide[,c(5:13)], na.rm = TRUE)) %>%
  mutate(Late_pres = rowSums(birds_years_wide[,c(14:22)], na.rm = TRUE))

#Plotting persistence
##COMMENTING OUT FOR NOW UNTIL I FIGURE OUT A BETTER WAY TO GRAPH
# birds_years_wide %>% ggplot(aes(x = AOU_code, y = Persistence, color = Watershed_name)) + 
#   geom_point(alpha = 0.6) + 
#   facet_grid(Watershed_name~.) + 
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=45, hjust=1, size =7))
# 
# birds_years_wide$Watershed_name <- factor(birds_years_wide$Watershed_name,
#                                           levels = c("1D", "N1B", "4A", "4B", "N4B", "N4D", "20B", "N20B"))
# 
# birds_years_wide %>% ggplot(aes(x = Persistence, fill = Watershed_name)) + geom_histogram(stat = "count") + facet_grid(.~Watershed_name) +
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=90, hjust=1, size =10))
# 
# 
# 
# birds_years_wide %>% ggplot(aes(x = Watershed_name, y = Persistence, color = Watershed_name)) + 
#   geom_point(alpha = 0.6, position = position_jitter(width = 0.2)) + 
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=45, hjust=1, size =7))
# 
# birds_years_wide %>% ggplot(aes(x = Persistence)) + geom_histogram() +
#   facet_wrap(.~AOU_code) + 
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=45, hjust=1, size =7))

#Pivoting the dataset back to long format for further analysis; this dataset included the persistence calculations so that I can classify always_absent, always_present, increasing, and decreasing incidence later on
birds_long_pers <- pivot_longer(birds_years_wide, cols = 5:22,
                     names_to = "Recyear",
                     values_to = "Presence")




# birds_long4 %>% ggplot(aes(x = Recyear, y = factor(Presence), fill = factor(Presence))) + 
#   geom_tile() + 
#   facet_grid(Species~Grazing~Fire_interval) +
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=45, hjust=1, size =7),
#         axis.text.y = element_text(size=7),
#         strip.text.y.left = element_text(angle = 0)) +
#   scale_fill_manual(values = c("white", "black")) 
# 
#  birds_long4 %>%  ggplot(aes(Recyear, factor(Presence), fill = factor(Presence))) +
#   geom_tile(color = "gray30") +
#    facet_grid(Species~Grazing~Fire_interval, switch = "y") +
#    theme(text = element_text(size=7),
#         axis.text.x = element_text(angle=45, hjust=1, size =7),
#         axis.text.y = element_text(size=7),
#         strip.text.y.left = element_text(angle = 0)) +
#    coord_equal() +
#   scale_fill_manual(values = c("white", "black"), guide = "none") +
#   theme_minimal(base_size = 16)


```


##Chi-square contingency tables
```{r}

#Making a new dataset just in case
birds_cont <- birds_long_pers

#Adding empty columns onto new dataset
birds_cont <- birds_cont %>% mutate(Early_late = NA,
                                    Presence = factor(Presence))

#Filling presence contingency column based on presence/absence
 # for(j in 1:nrow(birds_cont)){
 #   if (birds_cont[j,10] == "1"){birds_cont[j,11] <- "Present"}
 #   if (birds_cont[j,10] == "0"){birds_cont[j,11] <- "Absent"}
 # } 

#Filling early/late contingency based on year
for(i in 1:nrow(birds_cont)){
   if (birds_cont[i,9] < 2001){birds_cont[i,11] <- "Early"}
   if (birds_cont[i,9] >= 2001){birds_cont[i,11] <- "Late"}
}


birds_cont <- birds_cont %>% mutate(Early_late = factor(Early_late))

summary_tbl <- birds_cont %>%
  group_by(AOU_code, Watershed_name, Fire_interval, Grazing, Persistence, Early_pres, Late_pres)%>%
    dplyr::count(Presence, Early_late)

summary_tbl <- summary_tbl %>% unite("Species_Watershed", AOU_code:Watershed_name, sep = "_") 
summary_tbl <- summary_tbl %>% mutate(Species_Watershed = factor(Species_Watershed))

contabs <- xtabs(n ~ Presence + Early_late + Species_Watershed, summary_tbl)
chisqtests <- apply(contabs, 3, chisq.test)

chisq_p.values <- tibble(names(chisqtests), do.call(rbind,lapply(chisqtests,function(v){v$p.value})))

chisq_p.values <- chisq_p.values %>% rename("chisq_p.values" = "do.call(...)",
                                            "Species_Watershed" = "names(chisqtests)")
chisq_p.values <- chisq_p.values %>% mutate(Species_Watershed = factor(Species_Watershed))



```

##Runs-test 
```{r}

#Selecting the variables I need for the runs.test and again combining the species and watershed to create unique identifiers
birds_long_runs <- birds_long_pers %>% 
  group_by(Recyear, AOU_code, Watershed_name, Persistence, Early_pres, Late_pres) %>% reframe(Presence = Presence) %>% unite("Species_Watershed", AOU_code:Watershed_name, sep = "_")

#Creating the runs.test matrices 
presence.tabs <- xtabs(Presence ~ Recyear + Species_Watershed, birds_long_runs)
#Applying runs.tests to the matrices (MARGIN = 2 means that they are looking at the columns)
runstests <- apply(presence.tabs, 2, runs.test)
#Extracting the p.values from the runs.test into a table
runstest_p.values <- tibble(names(runstests), do.call(rbind,lapply(runstests,function(v){v$p.value})))
#Renaming the variables from the runs.test p.value table
runstest_p.values <- runstest_p.values %>% rename("runstest_p.values" = "do.call(...)",
                                                  "Species_Watershed" = "names(runstests)")


```


##Incidence classification
```{r}

#Merging the p.value outputs from the chi.sq and runs.tests 
merged_p.values <- merge(chisq_p.values,runstest_p.values, by = "Species_Watershed", all = TRUE)


#Separating the combined Species_watershed again so that I can join these p.values to previous datasets
merged_p.values1 <- merged_p.values %>% separate(Species_Watershed, into = c("AOU_code", "Watershed_name"))

#Joining the merged p.values with a previous dataset to get all the variables
merged_p.values1 <- left_join(merged_p.values1, (dplyr::select(birds_years_wide, c(Watershed_name, Fire_interval, Grazing, AOU_code, Persistence, Early_pres, Late_pres))), by = c("Watershed_name", "AOU_code"))

#Adding new empty columns, one to indicate incidence change (chi.sq significance) and one to indicate sequential patterns(runs.test significance)
merged_p.values2 <- merged_p.values1 %>% mutate(Incidence_change = NA,
                                          Sequential_patterns = NA)

#Avoiding the NA/NaN error by changing all NA's/NaN to "2" (will change back later)
for (k in 1:nrow(merged_p.values2)){
  if (is.nan(merged_p.values2$chisq_p.values[k])) {merged_p.values2$chisq_p.values[k] <- 2}
  if (is.na(merged_p.values2$runstest_p.values[k])) {merged_p.values2$runstest_p.values[k] <- 2}
}

#Copying the dataset again (this is all just for caution)
merged_p.values3 <- merged_p.values2

#Marking the incidence change and sequential patterns with either a "significant" (p.value < 0.05), a "-" (in the original output, p-value was NaN), or a "NS" (not significant, p.value > 0.05)
for (kk in 1:nrow(merged_p.values3)){
  if (merged_p.values3$chisq_p.values[kk] == 2){merged_p.values3$Incidence_change[kk] <- "-"}
  if (merged_p.values3$chisq_p.values[kk] <= 0.05){merged_p.values3$Incidence_change[kk] <- "Significant"}
  if (merged_p.values3$chisq_p.values[kk] > 0.05 & merged_p.values3$chisq_p.values[kk] < 2){merged_p.values3$Incidence_change[kk] <- "NS"}
  if (merged_p.values3$runstest_p.values[kk] == 2){merged_p.values3$Sequential_patterns[kk] <- "-"}
  if (merged_p.values3$runstest_p.values[kk] <= 0.05){merged_p.values3$Sequential_patterns[kk] <- "Significant"}
  if (merged_p.values3$runstest_p.values[kk] > 0.05 & merged_p.values3$runstest_p.values[kk] < 2){merged_p.values3$Sequential_patterns[kk] <- "NS"}
}

#Copying the dataset again and adding a classification column (this is all just for caution)
merged_p.values4 <- merged_p.values3 %>% mutate(Classification = NA)

#Classifying the incidence based on the "incidence change" and "sequential patterns" columns
for (jj in 1:nrow(merged_p.values4)){
  if (merged_p.values4$Incidence_change[jj] == "-" & merged_p.values4$Sequential_patterns[jj] == "-" & merged_p.values4$Persistence[jj] == 0){merged_p.values4$Classification[jj] <- "No_change.always_absent"} #Chi.sq and runs.tests p.value are NaN; persistence is "0"
  
  if (merged_p.values4$Incidence_change[jj] == "-" & merged_p.values4$Sequential_patterns[jj] == "-" & merged_p.values4$Persistence[jj] == 1){merged_p.values4$Classification[jj] <- "No_change.always_present"} #Chi.sq and runs.tests p.value are NaN; persistence is "1"
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "NS"){merged_p.values4$Classification[jj] <- "Random"} #Chi.sq and runs.test p.values are both > 0.05 
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "Significant"){merged_p.values4$Classification[jj] <- "Recurrent"} #Chi.sq p.value is > 0.05; runs.test p.value is < 0.05
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "-"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test p.value is NaN
  
  if (merged_p.values4$Incidence_change[jj] == "NS" & merged_p.values4$Sequential_patterns[jj] == "-"){merged_p.values4$Classification[jj] <- "Random"} #Chi.sq p.value is > 0.05; runs.test is NaN
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "Significant"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test is < 0.05
  
  if (merged_p.values4$Incidence_change[jj] == "Significant" & merged_p.values4$Sequential_patterns[jj] == "NS"){merged_p.values4$Classification[jj] <- "Significant"} #Chi.sq p.value is < 0.05; runs.test p.value > 0.05
}

#Copying the dataset again (this is all just for caution)
merged_p.values5 <- merged_p.values4  

#Classifying the Chi.sq p.values that are < 0.05 into 'decreasing' and 'increasing' based on its early presence vs late presence
for (xx in 1:nrow(merged_p.values5)){
  if (merged_p.values5$Classification[xx] == "Significant" & ((merged_p.values5$Early_pres[xx]) > (merged_p.values5$Late_pres[xx]))){merged_p.values5$Classification[xx] <- "Decreasing"} #Chi.sq p.value is < 0.05; early persistence is more than late presence
  if (merged_p.values5$Classification[xx] == "Significant" & ((merged_p.values5$Late_pres[xx]) > (merged_p.values5$Early_pres[xx]))){merged_p.values5$Classification[xx] <- "Increasing"} #Chi.sq p.value is < 0.05; early persistence is less than late presence
}

#Reverting the "2" for the p.values back into NAs
for (ii in 1:nrow(merged_p.values5)){
  if (merged_p.values5$chisq_p.values[ii] == 2) {merged_p.values5$chisq_p.values[ii] <- NA}
  if (merged_p.values5$runstest_p.values[ii] == 2) {merged_p.values5$runstest_p.values[ii] <- NA}
}



# birds_check <- birds_long4 %>% subset(Species_Watershed %in% c("BARS_4A", "BGGN_N20B", "BGGN_N4D", "EAKI_N20B", "EAKI_N4D", "EATO_4A", "EATO_N4D", "INBU_20B", "INBU_N4D", "RNEP_1D", "RSTO_N20B", "RSTO_N4D", "RWBL_4A", "EATO_20B", "EATO_N20B", "FISP_4B", "INBU_N20B"))
# 
# birds_check %>% ggplot(aes(x = Recyear, y = factor(Presence), fill = factor(Presence))) + 
#   geom_tile(width = .9, height = .9) + 
#   facet_wrap(.~Species_Watershed, switch = "y") +
#   theme(text = element_text(size=10),
#         axis.text.x = element_text(angle=45, hjust=1, size =7),
#         axis.text.y = element_text(size=7),
#         strip.text.y.left = element_text(angle = 0)) +
#   labs(x = "Year", y = "Presence", fill = "Presence")
# 
# grouped_idk <- merged_p.values5 %>% group_by(Classification, Fire_interval, Grazing) %>% summarize(total_count=n(),.groups = 'drop')
# 
# grouped_idk %>% ggplot(aes(x = Classification, y = total_count, fill = Classification)) + geom_col() +
#   facet_wrap(Fire_interval~Grazing)

```



##Exploration Plots
```{r}

#Joining the merged p.value dataset that includes classification with a previous dataset to have all variables (year, fire-interval, grazing, etc.)
birds_class <- left_join(birds_long_pers, merged_p.values5)

#Reordering the classification variables for plotting
birds_class$Classification <- factor(birds_class$Classification, 
                               levels = c("No_change.always_absent", "No_change.always_present", 
                                          "Random", "Increasing", "Decreasing"))

birds_class %>% ggplot(aes(x = Recyear, y = factor(Presence), fill = factor(Presence))) + 
  geom_tile(width = .6, height = .6) + 
  facet_grid(AOU_code~Watershed_name, switch = "y") +
  theme(text = element_text(size=5),
        axis.text.x = element_text(angle=45, hjust=1, size =5),
        axis.text.y = element_text(size=5),
        strip.text.y.left = element_text(angle = 0)) +
  scale_fill_manual(values = c("NA", "black")) +
  labs(x = "Year", y = "Presence", fill = "Presence")

#Copying the dataset to manipulate it for graphing
birds_graphing <- birds_class

#Making all presence = "0" inti NA's so they don't show up in theb graph
birds_graphing$Presence[birds_graphing$Presence == 0] <- NA

#Making the years numeric instead of characters
birds_graphing <- birds_graphing %>% mutate(Recyear = as.character(Recyear))
birds_graphing <- birds_graphing %>% mutate(Recyear = as.numeric(Recyear))

#Changing the order of the watersheds so that the grazing and non-grazing watersheds of the same burn interval are next to each other
birds_graphing$Watershed_name <- factor(birds_graphing$Watershed_name,
                                levels = c("1D", "N1B", "4A", 
                                           "4B", "N4B", "N4D", 
                                           "20B", "N20B"))

birds_graphing$Classification <- factor(birds_graphing$Classification, 
                               levels = c("No_change.always_absent", "No_change.always_present", "Random", "Increasing", "Decreasing"))

#Graphing all species by classification and watershed
birds_graphing %>% ggplot(aes(x = Recyear, y = Presence, fill = Classification)) + 
  geom_tile(width = .6, height = .6) + 
  facet_grid(AOU_code~Watershed_name, switch = "y") +
  theme(text = element_text(size=7),
        axis.text.x = element_text(angle=45, hjust=1, size =7),
        axis.text.y = element_text(size=5),
        strip.text.y.left = element_text(angle = 0, size = 3),
        strip.text.x.top = element_text(size = 7)) +
  scale_fill_manual(values = c("NA", "black", "dodgerblue2", "darkolivegreen4", "firebrick3")) +
  labs(x = "Year", y = "Presence", fill = "Classification")


#Making a separate dataset with just the species that have a significant incidence classification (increasing or decreasing) in one of the watersheds
birds_graphing_sig <- birds_graphing %>% subset(AOU_code %in% c("BARS", "BGGN", "EAKI", "EATO", "INBU", "RNEP", "RWBL", "FISP"))

#Graphing the species that have a significant classifcation
birds_graphing_sig %>% ggplot(aes(x = Recyear, y = Presence, fill = Classification)) + 
  geom_tile(width = .6, height = .6) + 
  facet_grid(AOU_code~Watershed_name, switch = "y", scales = "free_y", space = "free_y") +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1, size =7),
        axis.text.y = element_text(size=7),
        strip.text.y.left = element_text(angle = 0)) +
  scale_fill_manual(values = c("NA", "black", "dodgerblue2", "darkolivegreen4", "firebrick3")) +
  geom_vline(xintercept = 2000.5, linetype= 3) +
  labs(x = "Year", y = "Presence") 


```



